---
title: "Session 3: EDA"
session_id: session-03
format:
  html: default
---

:::{.direction}

In this exercise, we cover:

| Exercise part | Time (min) |
|---|---:|
| [EDA: Clustering](#clustering) | 25 |
| [Wrap-up](#wrap-up) | 10 |
| **Overall** | **35** |

Work directly in this Quarto (`.qmd`) exercise using R in RStudio, Positron, or VS Code. Work through the tasks in R and compare your work with the solution.

### Package setup

Before starting this exercise, make sure the required R packages are installed.

Run this command once in the R console. You do not need to run it every time you open the exercise. You can skip this step if these packages are already installed.

```{r}
#| eval: false
install.packages("tidyverse")
```

:::

# EDA: Clustering {#clustering}

Now that the data is structured and cleaned, we explore whether companies group into similar patterns.
Clustering helps us move from preparation to insight generation: instead of looking at firms one by one, we ask whether there are recurring profiles in the data.

Dataset: `firm_data.csv`

```{r}
#| output: false

library(tidyverse)

firm_data_df <- read_csv(
  "https://raw.githubusercontent.com/fs-ise/machine-learning-for-big-data/main/exercises/data/firm_data.csv",
  show_col_types = FALSE
)

firm_data_df |>
  slice_head(n = 5)
```

:::{.direction}

## Tasks

### 1. Prepare the data for clustering

We decide to use `profit_margin` and `rd_intensity` to cluster firms.
In a first step, create a data frame `X` containing these two variables:

```{r}
X <- firm_data_df |>
  select(profit_margin, rd_intensity)

# Print X at the end.
X
```

Before clustering, do a quick exploratory check so you know what kind of patterns you are looking for.

```{r}
summary(X)
```

```{r}
X |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ variable, scales = "free") +
  labs(
    x = NULL,
    y = "Count"
  )
```

Considering that k-means is sensitive to different measurement scales, we need to standardize the values.
This ensures that they have approximately equal influence on the clusters. Base R's `scale()` centers each variable around zero and scales it to unit standard deviation.

Run the following code and observe how the scale of both variables changes:

```{r}
X_scaled <- scale(X)
head(X_scaled)
```

Before running a clustering algorithm, view the standardized data in a scatterplot:

```{r}
X_scaled_df <- as.data.frame(X_scaled)

ggplot(
  X_scaled_df,
  aes(x = profit_margin, y = rd_intensity)
) +
  geom_point() +
  labs(
    title = "Standardized data (no clustering)",
    x = "Profit margin (scaled)",
    y = "R&D intensity (scaled)"
  )
```

How many clusters would you expect? Briefly justify your answer based on the plot.

**Your answer:** *Write your answer here.*

### 2. Apply k-means clustering

* Next, run the k-means clustering algorithm for different values of `k`.

```{r}
# YOUR CODE HERE: try values from 2 to 10.
# Remember that kmeans() expects centers = k.
```

* Try different values of `k` (e.g., 2–10). Which number of clusters seems appropriate?

**Your answer:** *Write your answer here.*
### 3. Add cluster labels to the dataset

* Assign the cluster labels returned by `kmeans()` to the original data frame (`firm_data_df`).

```{r}
# YOUR CODE HERE
```


### 4. Create grouped boxplots

* Generate a boxplot, grouped according to the clusters, to compare differences in `revenue`.
* Try to complete the task using the [`ggplot2` boxplot documentation](https://ggplot2.tidyverse.org/reference/geom_boxplot.html).

```{r}
# YOUR CODE HERE
```

* What do the revenue distributions tell you about the firm profiles represented by the clusters?

**Your answer:** *Write your answer here.*

:::

:::{.sol}

## Solution

### 1. Prepare the data for clustering

We use only `profit_margin` and `rd_intensity` for clustering:

```{r}
X <- firm_data_df |>
  select(profit_margin, rd_intensity)

X |>
  slice_head(n = 5)
```

#### Exploratory check

```{r}
summary(X)
```

```{r}
X |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ variable, scales = "free") +
  labs(
    x = NULL,
    y = "Count"
  )
```

These summaries and histograms help us understand the distribution and scale of the variables.
#### Standardization

Since k-means is sensitive to scale differences, we standardize the variables. `scale()` centers each variable around zero and scales it to unit standard deviation.

```{r}
X_scaled <- scale(X)
head(X_scaled)
```

#### Scatterplot (before clustering)

```{r}
X_scaled_df <- as.data.frame(X_scaled)

ggplot(
  X_scaled_df,
  aes(x = profit_margin, y = rd_intensity)
) +
  geom_point() +
  labs(
    title = "Standardized data (no clustering)",
    x = "Profit margin (scaled)",
    y = "R&D intensity (scaled)"
  )
```

**Interpretation:**
From the scatterplot, we might expect a small number of clusters (e.g., 3–5), depending on visible groupings.
### 2. Apply k-means clustering

We try different values of `k` and visually inspect the cluster structure. Setting `nstart = 25` makes the result less sensitive to a single random initialization.

```{r}
for (k in 2:10) {
  set.seed(42)
  km <- kmeans(
    X_scaled,
    centers = k,
    nstart = 25
  )

  cluster_plot_data <- X_scaled_df |>
    mutate(cluster = factor(km$cluster))

  centers <- as.data.frame(km$centers)

  cluster_plot <- ggplot(
    cluster_plot_data,
    aes(
      x = profit_margin,
      y = rd_intensity,
      color = cluster
    )
  ) +
    geom_point() +
    geom_point(
      data = centers,
      aes(x = profit_margin, y = rd_intensity),
      inherit.aes = FALSE,
      shape = 4,
      size = 4,
      stroke = 1.5
    ) +
    labs(
      title = paste("k =", k),
      x = "Profit margin (scaled)",
      y = "R&D intensity (scaled)",
      color = "Cluster"
    )

  print(cluster_plot)
}
```

**Conclusion:**
A value around **k = 4** appears to provide a reasonable balance between simplicity and structure.
### 3. Add cluster labels to the dataset

We now fit the final model reproducibly and store the cluster labels as a factor because they represent categories rather than quantities:

```{r}
set.seed(42)

k <- 4
km <- kmeans(
  X_scaled,
  centers = k,
  nstart = 25
)

firm_data_df <- firm_data_df |>
  mutate(cluster = factor(km$cluster))

firm_data_df |>
  slice_head(n = 5)
```
### 4. Create grouped boxplots

We compare `revenue` across clusters:

```{r}
ggplot(
  firm_data_df,
  aes(x = cluster, y = revenue)
) +
  geom_boxplot() +
  labs(
    title = "Revenue by cluster",
    x = "Cluster",
    y = "Revenue"
  )
```

**Interpretation:**
The boxplots reveal how revenue differs across clusters, helping to characterize the identified firm profiles.

:::

:::{.direction}

## Wrap-up {#wrap-up}

🎉🎈 You have completed the exercise—good work! 🎈🎉

In this exercise, we have learned to

* Explore variable distributions and relationships
* Prepare selected variables for clustering
* Apply and interpret k-means clustering

:::
