---
session: "Session 7"
title: "Machine Learning Foundations: Algorithms"
session_id: session-07
format:
  html: default
---

::: {.direction}

In this exercise, we cover:

| Part | Topic | Time (min) |
|---|---|---:|
| 1 | Common evaluation setup | 10 |
| 2 | Regularized logistic regression | 15 |
| 3 | k-nearest neighbors | 11 |
| 4 | Decision tree | 12 |
| 5 | Random forest | 12 |
| 6 | RBF support vector machine | 12 |
| 7 | Validation comparison and recommendation | 8 |
| 8 | Final held-out evaluation | 7 |
| Wrap-up | Reflection | 5 |
| **Overall** | | **90** |

## Learning objectives

After this exercise, you should be able to:

- apply selected machine-learning algorithms in R;
- explain how their main tuning parameters control model flexibility;
- compare regularized logistic regression, k-NN, decision trees, random forests, and SVMs;
- compare algorithms using a common validation procedure;
- consider interpretability, computational cost, preprocessing requirements, and predictive performance;
- recommend an appropriate model for a business prediction task.

> In Session 6, we learned that models must be compared using the same training data and validation procedure. Here, we keep that workflow fixed and vary the algorithm.

### Package setup

Before starting this exercise, make sure the required R packages are installed.

Run this command once in the R console. You do not need to run it every time you open the exercise. You can skip this step if these packages are already installed.

```{r}
#| label: package-setup-install-packages
#| eval: false
install.packages(c(
  "tidyverse",
  "tidymodels",
  "glmnet",
  "kknn",
  "ranger",
  "kernlab",
  "rpart.plot"
))
```

:::

```{r}
#| label: package-setup-load-packages
#| solution-setup: true
library(tidyverse)
library(tidymodels)
```

# Part 1 — Establish the common evaluation setup {#setup}

We continue the LendWise default-classification task from Sessions 5 and 6. Every algorithm receives the same training observations and five cross-validation folds. The held-out test set remains unused until a recommendation has been made from validation results.

## Task 1.1 — Re-establish the Session 6 setup

::: {.direction}
Run the common setup and verify:

- `default` is the first factor level and therefore the event for sensitivity and ROC AUC;
- the split is stratified and reproducible;
- all five folds come only from `train_data`.
:::

```{r}
#| label: task-1-1-re-establish-the-session-6-setup-df
df <- read_csv(
  "https://raw.githubusercontent.com/fs-ise/machine-learning-for-big-data/main/exercises/data/credit_default.csv",
  show_col_types = FALSE
) |>
  mutate(
    default = factor(
      default,
      levels = c(1, 0),
      labels = c("default", "no_default")
    )
  )

set.seed(42)

data_split <- initial_split(
  df,
  prop = 0.8,
  strata = default
)

train_data <- training(data_split)
test_data <- testing(data_split)

set.seed(42)

folds <- vfold_cv(
  train_data,
  v = 5,
  strata = default
)

classification_metrics <- metric_set(
  roc_auc,
  accuracy,
  sens
)

levels(train_data$default)
nrow(train_data)
nrow(test_data)
folds
```

::: {.direction}
Why must every algorithm use these same folds? Why must `test_data` remain outside model comparison?

**Your answer:** *Write your answer here.*
:::

::: {.sol}
Using identical folds makes performance differences attributable to algorithms and tuning choices rather than different validation samples. The test set remains independent so that it can provide one final estimate of generalization after all model choices are fixed. Consulting it during comparison would introduce selection bias.
:::

## Task 1.2 — Define algorithm-specific preprocessing

Some algorithms depend directly on distances or coefficient penalties, while tree splits are largely insensitive to predictor scale.

::: {.direction}
Complete two recipes:

- `scaled_recipe` dummy-encodes nominal predictors and normalizes numeric predictors;
- `tree_recipe` keeps the original predictors without normalization.

Which model families should use each recipe?

**Your answer:** *Write your answer here.*

```{r}
#| label: task-1-2-define-algorithm-specific-preprocessing-student-response
# YOUR CODE HERE
```
:::

::: {.sol}

```{r}
#| label: task-1-2-define-algorithm-specific-preprocessing-scaled-recipe
scaled_recipe <- recipe(
  default ~ .,
  data = train_data
) |>
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors())

tree_recipe <- recipe(
  default ~ .,
  data = train_data
)

scaled_recipe
tree_recipe
```

Regularized regression, k-NN, and the SVM use `scaled_recipe`. The decision tree and random forest use `tree_recipe`. Scaling is essential for distance-based k-NN and helpful for penalty- and margin-based models. Tree-based predictions depend on ordered split points, so rescaling adds no useful information.
:::


# Part 2 — Regularized logistic regression {#regularized}

Regularization balances fit and simplicity. `penalty` controls its strength: larger values shrink coefficients more. `mixture = 0` gives ridge, `mixture = 1` gives lasso, and intermediate values give elastic net.

## Task 2.1 — Tune the model

::: {.direction}
Run the supplied specification and workflow to tune the 18 explicit combinations in `regularized_grid` on the shared folds. The wider penalty range makes both mild and substantial shrinkage available for comparison.

```{r}
#| label: task-2-1-specify-and-tune-the-model-student-response
regularized_grid <- expand_grid(
  penalty = c(0.0001, 0.001, 0.01, 0.1, 1, 10),
  mixture = c(0, 0.5, 1)
)

regularized_spec <- logistic_reg(
  penalty = tune(),
  mixture = tune()
) |>
  set_engine("glmnet") |>
  set_mode("classification")

regularized_workflow <- workflow() |>
  add_recipe(scaled_recipe) |>
  add_model(regularized_spec)

set.seed(42)

regularized_results <- tune_grid(
  regularized_workflow,
  resamples = folds,
  grid = regularized_grid,
  metrics = classification_metrics
)

best_regularized <- select_best(
  regularized_results,
  metric = "roc_auc"
)

best_regularized
show_best(regularized_results, metric = "roc_auc", n = 5)
```

How does increasing `penalty` change model flexibility? Why is normalization appropriate here?

**Your answer:** *Write your answer here.*
:::

::: {.sol}

```{r}
#| label: task-2-1-specify-and-tune-the-model-regularized-spec
regularized_spec <- logistic_reg(
  penalty = tune(),
  mixture = tune()
) |>
  set_engine("glmnet") |>
  set_mode("classification")

regularized_workflow <- workflow() |>
  add_recipe(scaled_recipe) |>
  add_model(regularized_spec)

regularized_grid <- expand_grid(
  penalty = c(0.0001, 0.001, 0.01, 0.1, 1, 10),
  mixture = c(0, 0.5, 1)
)

set.seed(42)

regularized_results <- tune_grid(
  regularized_workflow,
  resamples = folds,
  grid = regularized_grid,
  metrics = classification_metrics
)

best_regularized <- select_best(
  regularized_results,
  metric = "roc_auc"
)

best_regularized
show_best(regularized_results, metric = "roc_auc", n = 5)
```

A larger penalty constrains coefficients more strongly and reduces flexibility. Normalization makes the penalty comparable across predictors measured in euros, counts, months, and proportions.
:::

## Task 2.2 — Compare unregularized, ridge, and lasso coefficients

::: {.direction}
Task 2.1 used cross-validation to choose tuning parameters for predictive model selection. Here, instead, use one shared, moderate penalty chosen for a clear conceptual comparison. Keeping the penalty strength fixed lets you attribute differences between ridge and lasso coefficients to the type of penalty rather than to two different CV-selected penalty values.

The supplied code uses `comparison_penalty <- 0.05` for both ridge and lasso and fits all three models to `train_data` with `scaled_recipe`. This value gives more instructive contrast than the initial value of `0.01` without collapsing the ridge coefficients. Complete the plot by mapping coefficient `term` to the y-axis and `estimate` to the x-axis.

```{r}
#| label: task-2-2-compare-ridge-and-lasso-coefficients
comparison_penalty <- 0.05

penalized_coefficients <- tibble(
  model = c("Ridge", "Lasso"),
  mixture = c(0, 1)
) |>
  mutate(
    fitted_workflow = map(
      mixture,
      \(mixture_value) {
        workflow() |>
          add_recipe(scaled_recipe) |>
          add_model(
            logistic_reg(
              penalty = comparison_penalty,
              mixture = mixture_value
            ) |>
              set_engine("glmnet") |>
              set_mode("classification")
          ) |>
          fit(data = train_data)
      }
    ),
    coefficients = map(
      fitted_workflow,
      \(fitted) tidy(extract_fit_parsnip(fitted))
    )
  ) |>
  select(model, coefficients) |>
  unnest(coefficients) |>
  filter(term != "(Intercept)")

unregularized_workflow <- workflow() |>
  add_recipe(scaled_recipe) |>
  add_model(
    logistic_reg() |>
      set_engine("glm") |>
      set_mode("classification")
  )

unregularized_coefficients <- unregularized_workflow |>
  fit(data = train_data) |>
  extract_fit_parsnip() |>
  tidy() |>
  filter(term != "(Intercept)") |>
  transmute(
    model = "Unregularized",
    term,
    estimate
  )

coefficient_comparison <- bind_rows(
  unregularized_coefficients,
  penalized_coefficients |>
    select(model, term, estimate)
) |>
  mutate(
    model = factor(
      model,
      levels = c("Unregularized", "Ridge", "Lasso")
    )
  ) |>
  complete(
    model,
    term = unregularized_coefficients$term,
    fill = list(estimate = 0)
  )
```

```{r}
#| label: task-2-2-compare-ridge-and-lasso-coefficients-student-response
# YOUR CODE HERE: plot estimate by term, colored by model.
```


Compare ridge and lasso with the unregularized baseline. How does ridge shrink the coefficients while retaining meaningful contributions across predictors? Under the **same penalty strength**, how does lasso differ from ridge, and does it set any coefficients exactly to zero? Why does ridge usually keep all coefficients nonzero while lasso can perform variable selection? Finally, why is this teaching penalty appropriate for conceptual comparison but not a replacement for the cross-validation used for predictive model selection in Task 2.1?

**Your answer:** *Write your answer here.*
:::

::: {.sol}

```{r}
#| label: task-2-2-compare-ridge-and-lasso-coefficients-fit-and-plot
comparison_penalty <- 0.05

penalized_coefficients <- tibble(
  model = c("Ridge", "Lasso"),
  mixture = c(0, 1)
) |>
  mutate(
    fitted_workflow = map(
      mixture,
      \(mixture_value) {
        workflow() |>
          add_recipe(scaled_recipe) |>
          add_model(
            logistic_reg(
              penalty = comparison_penalty,
              mixture = mixture_value
            ) |>
              set_engine("glmnet") |>
              set_mode("classification")
          ) |>
          fit(data = train_data)
      }
    ),
    coefficients = map(
      fitted_workflow,
      \(fitted) tidy(extract_fit_parsnip(fitted))
    )
  ) |>
  select(model, coefficients) |>
  unnest(coefficients) |>
  filter(term != "(Intercept)")

unregularized_workflow <- workflow() |>
  add_recipe(scaled_recipe) |>
  add_model(
    logistic_reg() |>
      set_engine("glm") |>
      set_mode("classification")
  )

unregularized_coefficients <- unregularized_workflow |>
  fit(data = train_data) |>
  extract_fit_parsnip() |>
  tidy() |>
  filter(term != "(Intercept)") |>
  transmute(
    model = "Unregularized",
    term,
    estimate
  )

coefficient_comparison <- bind_rows(
  unregularized_coefficients,
  penalized_coefficients |>
    select(model, term, estimate)
) |>
  mutate(
    model = factor(
      model,
      levels = c("Unregularized", "Ridge", "Lasso")
    )
  ) |>
  complete(
    model,
    term = unregularized_coefficients$term,
    fill = list(estimate = 0)
  )

coefficient_comparison |>
  mutate(term = fct_reorder(term, abs(estimate), .fun = max)) |>
  ggplot(aes(x = estimate, y = term, color = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(position = position_dodge(width = 0.65), size = 2) +
  labs(
    title = "Coefficient comparison: unregularized, ridge, and lasso",
    x = "Coefficient estimate",
    y = NULL,
    color = NULL
  )

coefficient_comparison |>
  filter(model == "Lasso", estimate == 0) |>
  select(term, estimate)
```

Use the unregularized estimates as the baseline when identifying the largest changes in magnitude. At the common penalty of `r comparison_penalty`, ridge visibly shrinks the coefficients but retains meaningful nonzero contributions across predictors. Lasso applies stronger, selective shrinkage under that same penalty; use the printed output to report which coefficients, if any, are exactly zero.

Ridge's L2 penalty continuously reduces coefficient magnitudes and usually does not make them exactly zero. Lasso's L1 penalty can make coefficients exactly zero, so it can perform variable selection. The shared penalty here was deliberately chosen to make those different behaviors interpretable; it is not claimed to be the best predictive penalty. For predictive model selection, use the cross-validated ROC AUC procedure in Task 2.1.
:::

{{< needspace 8 >}}

# Part 3 — k-nearest neighbors {#knn}

k-NN predicts from nearby labeled observations. Small `k` produces flexible, high-variance boundaries; large `k` smooths over more observations and increases bias.

## Task 3.1 — Tune the number of neighbors

::: {.direction}
Use Euclidean distance, rectangular neighbor weights, and candidate values 3, 5, 9, 15, and 25.

::: {.callout-tip title="Useful documentation"}
See the [`nearest_neighbor()` documentation](https://parsnip.tidymodels.org/reference/nearest_neighbor.html) for its model-specific arguments and engine choices.
:::

```{r}
#| label: task-3-1-tune-the-number-of-neighbors-student-response
knn_grid <- tibble(
  neighbors = c(3L, 5L, 9L, 15L, 25L, 30L, 40L, 50L)
)

# YOUR CODE HERE: complete knn_spec, knn_workflow, and knn_results.
```

Why would unscaled `income` dominate predictors that range between 0 and 1?

**Your answer:** *Write your answer here.*
:::

::: {.sol}

```{r}
#| label: task-3-1-tune-the-number-of-neighbors-knn-spec
knn_spec <- nearest_neighbor(
  neighbors = tune(),
  weight_func = "rectangular",
  dist_power = 2
) |>
  set_engine("kknn") |>
  set_mode("classification")

knn_workflow <- workflow() |>
  add_recipe(scaled_recipe) |>
  add_model(knn_spec)

knn_grid <- tibble(
  neighbors = c(3L, 5L, 9L, 15L, 25L)
)

set.seed(42)

knn_results <- tune_grid(
  knn_workflow,
  resamples = folds,
  grid = knn_grid,
  metrics = classification_metrics
)

best_knn <- select_best(knn_results, metric = "roc_auc")

best_knn
show_best(knn_results, metric = "roc_auc", n = 5)

knn_results |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  ggplot(aes(x = neighbors, y = mean)) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_continuous(breaks = knn_grid$neighbors) +
  labs(
    title = "k-NN validation performance",
    x = "Number of neighbors (k)",
    y = "Mean cross-validation ROC AUC"
  )
```

Distance calculations use numerical differences. Without scaling, a difference of hundreds of euros can overwhelm meaningful differences in utilization or debt ratios merely because of units.
:::

## Task 3.2 — Interpret model flexibility

::: {.direction}
Inspect the ROC AUC plot. How do very small and larger values of `k` change the smoothness and flexibility of the decision boundary? Describe the observed validation pattern and explain it using bias, variance, and generalization. Does the most flexible candidate perform best?

**Your answer:** *Write your answer here.*
:::

::: {.sol}
Small `k` creates highly local boundaries with low bias but high variance: individual observations can substantially change predictions. Larger `k` averages over a wider neighborhood, producing a smoother, less variable boundary but potentially increasing bias. Use the plotted ROC AUC values to identify where that trade-off generalizes best; the answer should refer to the observed curve rather than assume that either extreme wins.
:::


# Part 4 — Decision tree {#tree}

A tree recursively partitions predictor space. Greater depth and smaller terminal-node requirements increase flexibility, while a larger `cost_complexity` encourages pruning.

## Task 4.1 — Tune tree complexity

::: {.direction}
Use eight combinations: cost complexity 0.001 or 0.01, depth 3 or 6, and minimum node size 10 or 25.

```{r}
#| label: task-4-1-tune-tree-complexity-student-response
# YOUR CODE HERE: complete the tree specification, workflow, grid, and tuning call.
```
:::

::: {.sol}

```{r}
#| label: task-4-1-tune-tree-complexity-tree-spec
tree_spec <- decision_tree(
  cost_complexity = tune(),
  tree_depth = tune(),
  min_n = tune()
) |>
  set_engine("rpart") |>
  set_mode("classification")

tree_workflow <- workflow() |>
  add_recipe(tree_recipe) |>
  add_model(tree_spec)

tree_grid <- expand_grid(
  cost_complexity = c(0.001, 0.01),
  tree_depth = c(3L, 6L),
  min_n = c(10L, 25L)
)

set.seed(42)

tree_results <- tune_grid(
  tree_workflow,
  resamples = folds,
  grid = tree_grid,
  metrics = classification_metrics
)

best_tree <- select_best(tree_results, metric = "roc_auc")

best_tree
show_best(tree_results, metric = "roc_auc", n = 5)
```

A single tree is often easy to explain as decision paths, but small changes in data can produce different splits. Complexity controls help limit overfitting.
:::

## Task 4.2 — Fit and interpret the selected tree

::: {.direction}
Run the supplied code to finalize the ROC-AUC-optimal tree, fit it to all training observations, and plot it. Interpret one or two important splits in substantive LendWise terms: which applicants are separated, and how do their predicted outcomes differ? Then explain how `tree_depth` and pruning through `cost_complexity` control flexibility.

```{r}
#| label: task-4-2-fit-and-interpret-the-selected-tree-student-response
#| fig-width: 14
#| fig-height: 9
#| out-width: 100%
final_tree_workflow <- finalize_workflow(
  tree_workflow,
  best_tree
)

fitted_tree_workflow <- fit(
  final_tree_workflow,
  data = train_data
)

fitted_tree <- extract_fit_engine(fitted_tree_workflow)

rpart.plot::rpart.plot(
  fitted_tree,
  type = 4,
  extra = 104,
  fallen.leaves = TRUE,
  tweak = 0.75,
  varlen = 0,
  faclen = 0,
  digits = 3,
  branch = 0.4,
  box.palette = "GnBu"
)
```

**Your answer:** *Write your answer here.*
:::

::: {.sol}

```{r}
#| label: task-4-2-fit-and-interpret-the-selected-tree-plot
#| fig-width: 14
#| fig-height: 9
#| out-width: 100%
final_tree_workflow <- finalize_workflow(
  tree_workflow,
  best_tree
)

fitted_tree_workflow <- fit(
  final_tree_workflow,
  data = train_data
)

fitted_tree <- extract_fit_engine(fitted_tree_workflow)

rpart.plot::rpart.plot(
  fitted_tree,
  type = 4,
  extra = 104,
  fallen.leaves = TRUE,
  tweak = 0.75,
  varlen = 0,
  faclen = 0,
  digits = 3,
  branch = 0.4,
  box.palette = "GnBu"
)
```

Begin at the root and state each split as an applicant rule—for example, whether values below or above a displayed threshold move an applicant toward a node with a different default composition. The exact variables and thresholds come from the fitted plot. Larger allowed depth permits more interactions and narrower subgroups; a larger cost-complexity value prunes branches unless they improve fit enough, yielding a less flexible tree that may generalize better.
:::

# Part 5 — Random forest {#forest}

A random forest aggregates trees trained on bootstrap samples while considering random predictor subsets. Aggregation generally reduces the instability of one tree, but the ensemble is harder to explain as a single decision path.

## Task 5.1 — Tune the forest

::: {.direction}
Keep `trees = 500` fixed. Tune `mtry` over 2, 4, and 6 predictors and `min_n` over 10 and 25.

::: {.callout-tip title="Useful documentation"}
See the [`rand_forest()` documentation](https://parsnip.tidymodels.org/reference/rand_forest.html) for its model-specific arguments, including `mtry`, `trees`, and `min_n`.
:::

```{r}
#| label: task-5-1-tune-the-forest-student-response
# YOUR CODE HERE: complete rf_spec, rf_workflow, rf_grid, and rf_results.
```

Why might a forest predict better than one tree? What interpretability is lost?

**Your answer:** *Write your answer here.*
:::

::: {.sol}

```{r}
#| label: task-5-1-tune-the-forest-rf-spec
rf_spec <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 500
) |>
  set_engine("ranger") |>
  set_mode("classification")

rf_workflow <- workflow() |>
  add_recipe(tree_recipe) |>
  add_model(rf_spec)

rf_grid <- expand_grid(
  mtry = c(2L, 4L, 6L),
  min_n = c(10L, 25L)
)

set.seed(42)

rf_results <- tune_grid(
  rf_workflow,
  resamples = folds,
  grid = rf_grid,
  metrics = classification_metrics
)

best_rf <- select_best(rf_results, metric = "roc_auc")

best_rf
show_best(rf_results, metric = "roc_auc", n = 5)
```

Averaging many decorrelated trees can reduce variance and improve generalization. Unlike one tree, however, a forest cannot be summarized by one transparent set of decision paths.
:::


{{< needspace 10 >}}

# Part 6 — RBF support vector machine {#svm}

An SVM seeks a decision boundary with a wide margin. `cost` controls how strongly classification errors are penalized. The RBF kernel permits nonlinear boundaries, while `rbf_sigma` controls how locally the kernel responds and therefore affects flexibility.

## Task 6.1 — Tune cost and kernel scale

::: {.direction}
Tune six combinations: cost 0.25, 1, or 4 and RBF sigma 0.01 or 0.1. Use normalized predictors.

::: {.callout-tip title="Useful documentation"}
See the [`svm_rbf()` documentation](https://parsnip.tidymodels.org/reference/svm_rbf.html) for model-specific guidance on `cost` and `rbf_sigma`.
:::

```{r}
#| label: task-6-1-tune-cost-and-kernel-scale-student-response
# YOUR CODE HERE: complete svm_spec, svm_workflow, svm_grid, and svm_results.
```
:::

::: {.sol}

```{r}
#| label: task-6-1-tune-cost-and-kernel-scale-svm-spec
svm_spec <- svm_rbf(
  cost = tune(),
  rbf_sigma = tune()
) |>
  set_engine("kernlab") |>
  set_mode("classification")

svm_workflow <- workflow() |>
  add_recipe(scaled_recipe) |>
  add_model(svm_spec)

svm_grid <- expand_grid(
  cost = c(0.25, 1, 4),
  rbf_sigma = c(0.01, 0.1)
)

set.seed(42)

svm_results <- tune_grid(
  svm_workflow,
  resamples = folds,
  grid = svm_grid,
  metrics = classification_metrics
)

best_svm <- select_best(svm_results, metric = "roc_auc")

best_svm
show_best(svm_results, metric = "roc_auc", n = 5)

svm_results |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  mutate(rbf_sigma = factor(rbf_sigma)) |>
  ggplot(
    aes(
      x = cost,
      y = mean,
      color = rbf_sigma,
      group = rbf_sigma
    )
  ) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_log10(breaks = c(0.25, 1, 4)) +
  labs(
    title = "RBF SVM validation performance",
    x = "Cost (log scale)",
    y = "Mean cross-validation ROC AUC",
    color = "RBF sigma"
  )
```

Scaling prevents the margin and RBF distances from being dominated by predictors with numerically large units.
:::

## Task 6.2 — Interpret the SVM tuning surface

::: {.direction}
Use the validation plot to answer: which `cost` and `rbf_sigma` combinations allow more local, flexible decision boundaries, and which encourage smoother boundaries? In this grid, does greater flexibility consistently improve validation ROC AUC? Support your answer with the plotted pattern.

**Your answer:** *Write your answer here.*
:::

::: {.sol}
A larger `rbf_sigma` makes each training observation's influence more local, permitting a more rapidly changing boundary. A larger `cost` penalizes errors more strongly and usually allows the fitted boundary to follow the training observations more closely. Thus high cost with high sigma is the most flexible corner of this grid, while low cost with low sigma is smoother. The validation plot—not training fit—shows whether that flexibility generalizes; compare all six means and do not assume that the most flexible combination must win.
:::

# Part 7 — Compare validation performance and recommend a model {#comparison}

All five algorithms have now been evaluated using the same folds. ROC AUC is the primary selection metric, while accuracy, sensitivity, uncertainty, and organizational requirements provide additional context.

## Task 7.1 — Collect each model's best metrics

::: {.direction}
Run the supplied helper and comparison code. It retrieves ROC AUC, accuracy, and sensitivity at each model's ROC-AUC-optimal configuration.

```{r}
#| label: task-7-1-collect-each-model-s-best-metrics-student-response
collect_best_metrics <- function(results, best_parameters, model_name) {
  all_metrics <- collect_metrics(results)
  parameter_names <- intersect(
    names(best_parameters),
    names(all_metrics)
  ) |>
    setdiff(".config")

  all_metrics |>
    inner_join(
      best_parameters,
      by = parameter_names
    ) |>
    mutate(model = model_name) |>
    select(
      model,
      .metric,
      mean,
      std_err,
      any_of(parameter_names)
    )
}

validation_comparison <- bind_rows(
  collect_best_metrics(
    regularized_results,
    best_regularized,
    "Regularized logistic"
  ),
  collect_best_metrics(
    knn_results,
    best_knn,
    "k-NN"
  ),
  collect_best_metrics(
    tree_results,
    best_tree,
    "Decision tree"
  ),
  collect_best_metrics(
    rf_results,
    best_rf,
    "Random forest"
  ),
  collect_best_metrics(
    svm_results,
    best_svm,
    "RBF SVM"
  )
)

validation_comparison |>
  arrange(.metric, desc(mean))
```
:::

::: {.sol}

```{r}
#| label: task-7-1-collect-each-model-s-best-metrics-collect-best-metrics
collect_best_metrics <- function(results, best_parameters, model_name) {
  all_metrics <- collect_metrics(results)
  parameter_names <- intersect(
    names(best_parameters),
    names(all_metrics)
  ) |>
    setdiff(".config")

  all_metrics |>
    inner_join(
      best_parameters,
      by = parameter_names
    ) |>
    mutate(model = model_name) |>
    select(
      model,
      .metric,
      mean,
      std_err,
      any_of(parameter_names)
    )
}

validation_comparison <- bind_rows(
  collect_best_metrics(
    regularized_results,
    best_regularized,
    "Regularized logistic"
  ),
  collect_best_metrics(
    knn_results,
    best_knn,
    "k-NN"
  ),
  collect_best_metrics(
    tree_results,
    best_tree,
    "Decision tree"
  ),
  collect_best_metrics(
    rf_results,
    best_rf,
    "Random forest"
  ),
  collect_best_metrics(
    svm_results,
    best_svm,
    "RBF SVM"
  )
)

validation_comparison |>
  arrange(.metric, desc(mean))
```
:::

## Task 7.2 — Visualize the comparison

::: {.direction}
Create a plot that compares the mean validation metrics across algorithms and includes uncertainty based on the cross-validation standard errors.

::: {.callout-tip title="Plotting hint"}
Use `geom_errorbar()` for intervals such as `mean ± 1.96 * std_err`. `position_dodge()` can keep metrics from overlapping, and `coord_flip()` can make model names easier to read.
:::

:::

::: {.sol}

```{r}
#| label: task-7-2-visualize-the-comparison-validation-comparison-pipeline
validation_comparison |>
  ggplot(
    aes(
      x = reorder(model, mean),
      y = mean,
      color = .metric
    )
  ) +
  geom_point(
    position = position_dodge(width = 0.5),
    size = 2
  ) +
  geom_errorbar(
    aes(
      ymin = mean - 1.96 * std_err,
      ymax = mean + 1.96 * std_err
    ),
    position = position_dodge(width = 0.5),
    width = 0.2
  ) +
  coord_flip() +
  labs(
    title = "Cross-validation performance by algorithm",
    x = NULL,
    y = "Mean validation metric",
    color = "Metric"
  )
```
:::

## Task 7.3 — Interpretability and predictive power

Typical properties—not absolute rules—include:

| Model | Typical strength | Typical limitation | Scaling needed here? |
|---|---|---|---|
| Regularized logistic | interpretable and stable | mostly linear decision boundary | yes |
| k-NN | simple and locally flexible | scaling and dimensionality sensitive | yes |
| Decision tree | transparent decision paths | unstable and can overfit | no |
| Random forest | strong flexible predictions | less directly interpretable | no |
| RBF SVM | powerful nonlinear boundaries | harder to interpret and tune | yes |

::: {.direction}
Answer:

1. Inspect the full comparison table—not only the ROC AUC ranking. Which model has the strongest mean validation ROC AUC? Is there a model with a competitive ROC AUC but clearly problematic accuracy or sensitivity?

**Your answer:** *Write your answer here.*

2. What does `sens = 0` mean substantively for LendWise?

**Your answer:** *Write your answer here.*

3. How can a model obtain a reasonably strong ROC AUC while predicting no defaults at the standard classification threshold?

**Your answer:** *Write your answer here.*

4. Are ROC AUC differences large relative to their standard errors? Which models naturally capture nonlinear relationships, and is the most flexible model necessarily best here?

**Your answer:** *Write your answer here.*

5. What matters for LendWise besides the highest score when choosing a model and its operating threshold?

**Your answer:** *Write your answer here.*
:::

::: {.sol}
The RBF SVM has the highest mean validation ROC AUC, about 0.730, but the full table reveals an important anomaly. The ROC-AUC-optimal regularized logistic regression (`penalty = 10`, `mixture = 0`, or ridge) is almost tied at about 0.729 ROC AUC—a difference of only about 0.001—yet its sensitivity is 0 and its accuracy is about 0.538. That threshold-based behavior differs much more substantially than the two models' ROC AUC values do. A sensitivity of 0 for the event class `default` means that, at the operating threshold used to create `.pred_class`, the model identifies none of the applicants who actually default as defaults. It predicts no observations as `default`, which would make it operationally useless for detecting risky applicants at that threshold.

This is not a defect in ROC AUC. ROC AUC evaluates how well predicted scores rank defaults above non-defaults over all possible thresholds; it does not evaluate classifications at one particular threshold. Accuracy and sensitivity computed from `.pred_class`, in contrast, describe the classifications at the software's standard threshold. Heavy ridge shrinkage can pull predicted probabilities toward the center—and, here, keep them on the `no_default` side of that threshold—while preserving enough ordering among applicants to achieve a reasonably strong ROC AUC. A model can therefore rank cases reasonably well while making unusable classifications at an inappropriate operating threshold.

The roughly 0.001 ROC-AUC gap should also be treated cautiously relative to fold-to-fold uncertainty. Trees, forests, k-NN, and the RBF SVM can capture nonlinear structure in different ways, but greater flexibility does not guarantee better generalization. LendWise should consider stability, computation, preprocessing, calibration, transparent explanations and governance for credit decisions, as well as the intended operating point and the costs of missed defaults and false alarms. Sensitivity, specificity, and other threshold-specific measures should be assessed at that operating point, and the threshold may need tuning on validation data rather than assuming 0.5 is appropriate. Ranking performance, model-family selection, and operating-threshold selection are related but distinct decisions; the held-out test set must remain untouched until all three have been fixed.
:::

{{< needspace 8 >}}

## Task 7.4 — Lock the recommendation using validation only

::: {.direction}
The code below identifies the highest mean validation ROC AUC. Before running the final evaluation, write a short recommendation that also considers stability and interpretability.
:::

::: {.sol}

```{r}
#| label: task-7-4-lock-the-recommendation-using-validation-only-roc-auc-compari
roc_auc_comparison <- validation_comparison |>
  filter(.metric == "roc_auc") |>
  arrange(desc(mean))

selected_model <- roc_auc_comparison |>
  slice_head(n = 1) |>
  pull(model)

selected_model
```

For this exercise, the quantitative rule selects the strongest mean validation ROC AUC. A business recommendation may reasonably prefer a nearly equivalent, more interpretable model when score differences are small and regulated credit decisions require clear explanations. The choice is locked before any test outcome is inspected.
:::


# Part 8 — Final held-out test evaluation {#final-test}

We now use the test set once. This is not another opportunity to tune the winner.

## Task 8.1 — Finalize the selected workflow

::: {.direction}
Run the selection map, finalize the chosen workflow with its best parameters, and evaluate it with `last_fit()`.
:::

::: {.sol}

```{r}
#| label: task-8-1-finalize-the-selected-workflow-workflow-lookup
workflow_lookup <- list(
  "Regularized logistic" = regularized_workflow,
  "k-NN" = knn_workflow,
  "Decision tree" = tree_workflow,
  "Random forest" = rf_workflow,
  "RBF SVM" = svm_workflow
)

parameter_lookup <- list(
  "Regularized logistic" = best_regularized,
  "k-NN" = best_knn,
  "Decision tree" = best_tree,
  "Random forest" = best_rf,
  "RBF SVM" = best_svm
)

selected_workflow <- workflow_lookup[[selected_model]]
best_parameters <- parameter_lookup[[selected_model]]

final_workflow <- finalize_workflow(
  selected_workflow,
  best_parameters
)

set.seed(42)

final_fit <- last_fit(
  final_workflow,
  split = data_split,
  metrics = classification_metrics
)

final_test_metrics <- collect_metrics(final_fit)
final_test_metrics
```
:::

## Task 8.2 — Compare validation and test performance

::: {.direction}
Compare the selected model's cross-validation metrics with its held-out test metrics. Then inspect the confusion matrix and ROC curve. Does the held-out result broadly support the validation-based recommendation?

**Your answer:** *Write your answer here.*
:::

::: {.sol}

```{r}
#| label: task-8-2-compare-validation-and-test-performance-selected-cv-metrics
#| out-width: 63%
selected_cv_metrics <- validation_comparison |>
  filter(model == selected_model) |>
  select(.metric, cv_mean = mean, cv_std_err = std_err)

performance_comparison <- selected_cv_metrics |>
  left_join(
    final_test_metrics |>
      select(.metric, test_estimate = .estimate),
    by = ".metric"
  )

performance_comparison

final_predictions <- collect_predictions(final_fit)

final_predictions |>
  conf_mat(
    truth = default,
    estimate = .pred_class
  )

final_predictions |>
  roc_curve(
    truth = default,
    .pred_default,
    event_level = "first"
  ) |>
  autoplot() +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed"
  ) +
  coord_equal() +
  labs(title = paste("Held-out ROC curve:", selected_model))
```

Does the test result broadly support the validation-based recommendation? A difference is expected because both estimates vary by sample. The held-out result should be reported honestly, not used to reopen the tuning grids.
:::

::: {.direction}


## Wrap-up

🎉🎈 You have completed the exercise—good work! 🎈🎉

You have learned to:

- apply regularized logistic regression, k-NN, decision trees, random forests, and RBF SVMs;
- connect tuning parameters to model flexibility;
- recognize algorithm-specific preprocessing requirements;
- compare algorithms under one cross-validation framework;
- balance predictive performance, stability, computation, and interpretability;
- select and evaluate a final model on unseen data.

<!--
::: {.callout-tip title="Session 7 survey"}
{{< survey session_07 exercise >}}
:::
-->

:::
