---
session: "Session 9"
title: "Deployment in Organizations: Explainable AI"
session_id: session-09
format: html
---

::: {.direction}

In this exercise, we return to the **LendWise loan-default case** from previous sessions. The bank has developed several predictive models. Before deploying them, we need to understand how their predictions are produced and how they can be explained to different stakeholders.

We first examine models that are **inherently interpretable**, where the model structure itself provides an explanation. We then treat a random forest as a black box and apply **post-hoc explainability methods**.

| Part | Topic | Time (min) |
|---|---|---:|
| 1 | Three models, one applicant | 15 |
| 2 | Logistic regression: coefficients as explanations | 15 |
| 3 | Decision tree: rules as explanations | 10 |
| 4 | Global post-hoc explanation: permutation importance | 15 |
| 5 | From average effects to individual what-if profiles | 15 |
| 6 | Local post-hoc explanation with SHAP | 10 |
| Wrap-up | Match the explanation to the question | 10 |
| **Overall** | | **90** |

### Learning objectives

After this exercise, you should be able to:

- distinguish **inherently interpretable models** from **post-hoc explanations**;
- distinguish **global** from **local** explanations;
- interpret logistic-regression coefficients and shallow decision trees;
- use permutation importance to investigate which variables a black-box model relies on;
- use partial-dependence and individual profiles to investigate how predictions change;
- use SHAP values to explain an individual prediction;
- identify important limitations of model explanations;
- select explanation methods for different organizational stakeholders.

### Package setup

Run this command once in the R console. You can skip this step if the packages are already installed.

```{r}
#| label: package-setup-install
#| eval: false
install.packages(c(
  "tidyverse",
  "tidymodels",
  "DALEX",
  "DALEXtra",
  "ranger",
  "rpart",
  "rpart.plot"
))
```

:::

```{r}
#| label: setup
#| solution-setup: true
library(tidyverse)
library(tidymodels)

tidymodels_prefer()

df <- read_csv(
  "https://raw.githubusercontent.com/fs-ise/machine-learning-for-big-data/main/exercises/data/credit_default.csv",
  show_col_types = FALSE
) |>
  mutate(
    customer_category = factor(customer_category),
    # The event of interest ("default") is the SECOND level.
    # R's glm() models the probability of the second level,
    # so raw glm coefficients then refer directly to default.
    # tidymodels and yardstick treat the FIRST level as the event
    # by default, so metric calls need event_level = "second".
    default = factor(
      default,
      levels = c(0, 1),
      labels = c("no_default", "default")
    )
  )

set.seed(42)

data_split <- initial_split(
  df,
  prop = 0.8,
  strata = default
)

train_data <- training(data_split)
test_data <- testing(data_split)

logistic_recipe <- recipe(default ~ ., data = train_data) |>
  step_dummy(all_nominal_predictors())

tree_recipe <- recipe(default ~ ., data = train_data)

logistic_spec <- logistic_reg() |>
  set_engine("glm")

tree_spec <- decision_tree(
  tree_depth = 3,
  min_n = 20,
  cost_complexity = 0
) |>
  set_engine("rpart") |>
  set_mode("classification")

rf_spec <- rand_forest(
  trees = 500,
  mtry = 3,
  min_n = 10
) |>
  set_engine("ranger") |>
  set_mode("classification")

logistic_fit <- workflow() |>
  add_recipe(logistic_recipe) |>
  add_model(logistic_spec) |>
  fit(data = train_data)

tree_fit <- workflow() |>
  add_recipe(tree_recipe) |>
  add_model(tree_spec) |>
  fit(data = train_data)

set.seed(42)

rf_fit <- workflow() |>
  add_recipe(tree_recipe) |>
  add_model(rf_spec) |>
  fit(data = train_data)

predict_default_probability <- function(model, new_data) {
  predict(
    model,
    new_data = new_data,
    type = "prob"
  ) |>
    pull(.pred_default)
}

rf_test_probability <- predict_default_probability(
  rf_fit,
  test_data
)

applicant_index <- which.min(
  abs(rf_test_probability - 0.5)
)

applicant <- test_data |>
  slice(applicant_index) |>
  select(-default)

applicant_outcome <- test_data$default[[applicant_index]]

set.seed(42)

explanation_data <- test_data |>
  slice(-applicant_index) |>
  slice_sample(
    n = min(500, nrow(test_data) - 1)
  )

explainer_rf <- DALEXtra::explain_tidymodels(
  rf_fit,
  data = explanation_data |> select(-default),
  y = as.numeric(explanation_data$default == "default"),
  predict_function = predict_default_probability,
  label = "Random forest",
  type = "classification",
  verbose = FALSE
)
```

::: {.direction}

Inspect the variables to familiarize yourself with the predictors and outcome.

```{r}
#| label: inspect-data-student
glimpse(df)
```

:::

# Part 1 — Three models, one applicant {#models}

The three models solve exactly the same prediction problem:

- **logistic regression** uses one additive equation;
- the **decision tree** uses explicit decision rules;
- the **random forest** combines predictions from 500 trees.

Throughout the exercise, we explain the same applicant. It was selected because the random-forest probability is closest to 0.5, making it a borderline case rather than an obviously easy prediction. The shallow tree uses depth 3 and `min_n = 20`.

An **inherently interpretable** model exposes understandable prediction logic; a **post-hoc** method explains a fitted model after training. **Global** explanations describe behavior across cases, while **local** explanations describe one case. **Model-specific** methods use a particular model structure; **model-agnostic** methods use inputs and predictions and can be applied across model types.

## Task 1.1 — Compare the predictions

::: {.direction}

Inspect `applicant`.

Then calculate the predicted probability of default for the applicant using:

- `logistic_fit`;
- `tree_fit`;
- `rf_fit`.

The helper function

```r
predict_default_probability(model, applicant)
```

returns the probability of default.

```{r}
#| label: task-1-1-student
applicant

# YOUR CODE HERE
```

At a 0.5 threshold, which class does each model predict? Even when predictions agree, could the reasons behind them differ?

**Your answer:** *Write your answer here.*

Can the predicted probability by itself explain **why** the model produced the prediction?

**Your answer:** *Write your answer here.*

:::

::: {.sol}

```{r}
#| label: task-1-1-solution
applicant

tibble(
  model = c(
    "Logistic regression",
    "Decision tree",
    "Random forest"
  ),
  default_probability = c(
    predict_default_probability(logistic_fit, applicant),
    predict_default_probability(tree_fit, applicant),
    predict_default_probability(rf_fit, applicant)
  )
) |>
  mutate(
    predicted_class = if_else(
      default_probability >= 0.5,
      "default",
      "no_default"
    )
  )
```

The probabilities need not be identical because the models represent relationships differently. At a 0.5 threshold, probabilities at least 0.5 predict `default`; models can agree on that class while relying on different relationships.

A probability alone does not explain the reasoning behind the prediction. To understand the prediction, we need information about the model structure or an additional explanation method.

**Answer.** In the reference run, the logistic regression (about 0.54), the decision tree (about 0.61) and the random forest (about 0.50, only just above the threshold) all predict `default`. Agreeing on the class does not mean agreeing on the reasons: the logistic regression adds weighted contributions of all predictors, the tree applies a few thresholds on a handful of variables, and the forest averages hundreds of trees. The probability only says *what* the model predicts, not *why*.

:::

{{< needspace 20 >}}

## Task 1.2 — Classify the models

::: {.direction}

Classify each model before continuing.

| Model | Inherently interpretable or black box? | Why? |
|--|----|---|
| Logistic regression | | |
| Shallow decision tree | | |
| Random forest | | |

:::

::: {.sol}

```{=latex}
\begingroup
\footnotesize
```

| Model | Inherently interpretable or black box? | Why? |
|--|----|----|
| Logistic regression | Inherently interpretable | A small set of coefficients describes how predictors enter the model. |
| Shallow decision tree | Inherently interpretable | The complete prediction logic can be represented as a small set of decision rules. |
| Random forest | Black box for practical interpretation | Hundreds of trees contribute to each prediction, so directly tracing the complete model is impractical. |

```{=latex}
\endgroup
```

Interpretability is not an absolute property. A very large regression specification or deep decision tree can also become difficult to understand.

:::

## Task 1.3 — Check predictive quality

::: {.direction}

Compute test-set ROC AUC for all three models with yardstick. Why does model quality matter before explaining a model, and what would a much weaker tree imply about the fidelity of its explanation to the random forest?

```{r}
#| label: task-1-3-student
model_fits <- list(logistic = logistic_fit, tree = tree_fit, random_forest = rf_fit)

# Hint: `default` is the second factor level, so pass `event_level = "second"`
# to yardstick metrics such as roc_auc().
# YOUR CODE HERE
```

**Your answer:** *Write your answer here.*

:::

::: {.sol}

```{r}
#| label: task-1-3-solution
logistic_auc <- predict(logistic_fit, test_data, type = "prob") |>
  bind_cols(test_data |> select(default)) |>
  roc_auc(truth = default, .pred_default, event_level = "second")

tree_auc <- predict(tree_fit, test_data, type = "prob") |>
  bind_cols(test_data |> select(default)) |>
  roc_auc(truth = default, .pred_default, event_level = "second")

rf_auc <- predict(rf_fit, test_data, type = "prob") |>
  bind_cols(test_data |> select(default)) |>
  roc_auc(truth = default, .pred_default, event_level = "second")

logistic_auc
tree_auc
rf_auc
```

Explanations are useful only if the model is sufficiently valid for its intended use. A much weaker tree can explain its own predictions faithfully, but it is a poor surrogate for the random forest and should not be presented as explaining that forest.

**Answer.** In the reference run the logistic regression has the highest test AUC (about 0.78), followed by the random forest (about 0.71) and the tree (about 0.66). All three are only moderately accurate, so any explanation describes a model of modest quality. Note also that the least transparent model does not outperform the transparent logistic regression here, which raises the question of whether the black box is needed at all.

:::

# Part 2 — Logistic regression: coefficients as explanations {#logistic}

A logistic regression exposes its internal structure directly. Each predictor has a coefficient determining how it changes the **log-odds** of default, holding other predictors constant.

## Task 2.1 — Inspect the coefficients

:::: {.direction}

Extract the fitted `glm` model with `extract_fit_engine()` and inspect its coefficients using `broom::tidy()`.

Because `default` is the second factor level, raw `glm` models the log-odds of **default** directly, so the coefficients need no sign adjustment. Add `odds_ratio = exp(estimate)` to express each effect as a multiplicative change in the odds of default.

Ignore `(Intercept)`, inspect `p.value`, and remember that coefficient magnitudes are not comparable across predictors on different scales unless predictors are standardized. Check the reference level with `levels(train_data$customer_category)`.

::: {.callout-tip title="Useful documentation"}
See the [`logistic_reg()` documentation](https://parsnip.tidymodels.org/reference/logistic_reg.html).
:::

```{r}
#| label: task-2-1-student
coef_tbl <- logistic_fit |>
  extract_fit_engine() |>
  broom::tidy() |>
  mutate(odds_ratio = exp(estimate))

# YOUR CODE HERE
```

Identify one predictor associated with **higher** predicted default risk and one associated with **lower** predicted default risk.

**Your answer:** *Write your answer here.*

As a sanity check, does each direction agree with the tree or partial-dependence profile?

**Your answer:** *Write your answer here.*

What does an odds ratio greater than `1` mean?

**Your answer:** *Write your answer here.*

::::

::: {.sol}

```{r}
#| label: task-2-1-solution
coef_tbl <- logistic_fit |>
  extract_fit_engine() |>
  broom::tidy() |>
  mutate(
    odds_ratio = exp(estimate)
  )

coef_tbl
```

Positive coefficients increase the predicted log-odds of default; negative coefficients decrease them, holding the other predictors constant.

An odds ratio above `1` means that a one-unit increase in the predictor multiplies the odds of default by a factor greater than one. An odds ratio below `1` reduces the odds.

Use `p.value` as evidence about sampling uncertainty, not as a measure of practical importance. Ignore the intercept when identifying characteristics, and sanity-check directions against the tree or a PDP. The interpretation depends on the predictor's unit; magnitudes across different scales are not directly comparable. Dummy variables are interpreted relative to the omitted reference category, which `levels(train_data$customer_category)` helps identify.

**Answers.**

- **Higher risk:** `missed_payments_12m` (odds ratio about 2.06 per additional missed payment, p < 0.001) and `credit_utilization` (odds ratio about 6.3 for a one-unit increase, i.e. the whole 0–1 range, p = 0.002). **Lower risk:** `months_with_company` (odds ratio about 0.98 per additional month, borderline with p ≈ 0.05) and `income` (the odds ratio of about 1.000 per unit looks like "no effect" only because one unit is tiny; per 1,000 units it is exp(−0.323) ≈ 0.72, p = 0.017).
- **Sanity check:** the tree splits first on `missed_payments_12m` (more missed payments, higher default share), sends high `credit_utilization` (at least 0.75) to a riskier branch, and has lower-income leaves with higher default shares. The PDPs for `credit_utilization` and `debt_to_income` also rise on average. All directions agree.
- **Odds ratio above 1:** a one-unit increase in the predictor multiplies the odds of default by that factor, holding the other predictors constant (2.06 is roughly a doubling of the odds per additional missed payment).
- **Reference category:** the omitted first level is `early_career_urban`. The `customer_category_platform_worker` coefficient (odds ratio about 1.5, p ≈ 0.16) compares platform workers with that reference group and is not statistically distinguishable from zero.
- Since `default` is the modeled level, `estimate` and `statistic` now have the same sign.

:::

## Task 2.2 — Explain the applicant

::: {.direction}

Use the applicant's encoded values (and reference categories) together with `coef_tbl`. Here, the **contribution** of a predictor is its coefficient times the difference between the applicant's value and the training average of that predictor, $\beta_j \times (x_j^{\text{applicant}} - \bar{x}_j^{\text{train}})$. Coefficient × raw value is usually uninformative because it depends on where the predictor's zero lies.

```{r}
#| label: task-2-2-student
# Bake the applicant and the training data with the fitted recipe.
# For each predictor, compute: coefficient * (applicant value - training mean).
# Sort by absolute contribution.
# YOUR CODE HERE
```

Which characteristics of this applicant appear to push the logistic-regression prediction toward **higher** default risk, and which push it toward **lower** risk?

**Your answer:** *Write your answer here.*

Why is this explanation **not** evidence that changing one of these variables would causally change the applicant's default risk?

**Your answer:** *Write your answer here.*

:::

::: {.sol}

```{r}
#| label: task-2-2-solution
baked_applicant <- extract_recipe(logistic_fit) |>
  bake(applicant)
baked_train <- extract_recipe(logistic_fit) |>
  bake(train_data)

term_contributions <- coef_tbl |>
  filter(term != "(Intercept)") |>
  transmute(
    term,
    coefficient = estimate,
    applicant_value = unlist(baked_applicant[1, term], use.names = FALSE),
    training_mean = unname(colMeans(baked_train[term])),
    contribution_default_log_odds =
      coefficient * (applicant_value - training_mean)
  ) |>
  arrange(desc(abs(contribution_default_log_odds)))

term_contributions
```

```{r}
#| label: task-2-2-check-solution
# Check: the contributions add up to the applicant's log-odds minus the
# log-odds of a hypothetical applicant at the training mean of every predictor.
glm_engine <- extract_fit_engine(logistic_fit)

average_profile_log_odds <- coef(glm_engine)[["(Intercept)"]] +
  sum(term_contributions$coefficient * term_contributions$training_mean)

applicant_log_odds <- as.numeric(
  predict(glm_engine, newdata = baked_applicant, type = "link")
)

tibble(
  sum_of_contributions = sum(term_contributions$contribution_default_log_odds),
  applicant_minus_average_profile = applicant_log_odds - average_profile_log_odds
)
```

Positive values push the applicant's log-odds of default **above**, and negative values **below**, those of a hypothetical applicant who sits at the training average on every predictor. The contributions add up exactly to the difference between the two log-odds (see the check). A coefficient alone is not an applicant-specific contribution: it also matters how far the applicant's value lies from the average (for a dummy variable, from the share of training cases in that category).

**Answer.** In the reference run, the largest push toward **higher** risk comes from `missed_payments_12m` (about +0.53; the applicant has 2, above the training average), followed by `income` (about +0.20; below-average income); `debt_to_income` is negligible (about +0.01). Pushing toward **lower** risk are `customer_category_platform_worker` (about −0.22; the applicant is not a platform worker, and platform workers carry higher modeled risk than the reference group `early_career_urban`), `credit_utilization` (about −0.10; below average) and `months_with_company` (about −0.08; above-average tenure).

However, the model estimates predictive associations. The coefficients do not establish causal effects. A predictor can be associated with default because of confounding, selection effects, correlated variables, or other characteristics represented in the training data.

:::

# Part 3 — Decision tree: rules as explanations {#tree}

A shallow decision tree provides another form of inherent interpretability. Instead of coefficients, it represents the prediction as a sequence of **if/then rules**.

## Task 3.1 — Visualize the tree

:::: {.direction}

Extract the fitted `rpart` model and visualize it. With `extra = 104`, each node shows the predicted class, class probabilities, and percentage of observations; probabilities are listed in factor-level order, so `no_default` comes first and `default` second. Use `nn = TRUE` to number nodes, and `rpart.plot::rpart.predict(..., nn = TRUE)` to check the applicant's leaf.

::: {.callout-tip title="Useful documentation"}
See the [`decision_tree()` documentation](https://parsnip.tidymodels.org/reference/decision_tree.html) and [`rpart.plot`](https://cran.r-project.org/package=rpart.plot).
:::

```{r}
#| label: task-3-1-student
tree_engine <- extract_fit_engine(tree_fit)
applicant

# YOUR CODE HERE
```

Trace the applicant manually through the tree.

Which variables determine the path?

**Your answer:** *Write your answer here.*

A shallow tree may predict `no_default` in most leaves when defaults are rare.

Which terminal node does the applicant reach, and what does that node tell you?

**Your answer:** *Write your answer here.*

::::

::: {.sol}

```{r}
#| label: task-3-1-solution

tree_engine <- extract_fit_engine(tree_fit)

rpart.plot::rpart.plot(
  tree_engine,
  type = 2,
  extra = 104,
  nn = TRUE,
  roundint = FALSE
)

applicant_leaf <- rpart.plot::rpart.predict(
  tree_engine,
  newdata = applicant,
  nn = TRUE
)
applicant_leaf
```

```{r}
#| label: task-3-1-path-solution
leaf_node <- as.numeric(applicant_leaf[1, "nn"])

tree_path <- rpart::path.rpart(
  tree_engine,
  nodes = leaf_node,
  print.it = FALSE
)[[1]]

tree_path
```

The applicant can be explained by following the applicable split at each node until reaching a leaf. In the reference run the path is:

1. `missed_payments_12m >= 1.5`: yes (the applicant has 2), go to node 2.
2. `missed_payments_12m >= 2.5`: no (2 < 2.5), go to node 5.
3. `income < 3542`: yes (income is about 2,211), reach **leaf 10**.

Only two variables determine the path: `missed_payments_12m` (twice) and `income`. `credit_utilization` and `debt_to_income` are not used for this applicant. Leaf 10 holds about 22% of the training cases, of which about 61% defaulted, which is where the tree's predicted probability of 0.61 comes from. Defaults are not rare in this data (about 46% of training cases), so several leaves predict `default`. The leaf is far from pure: roughly 4 in 10 similar applicants did not default.

The variables encountered on that path constitute a compact local explanation of this prediction. The leaf summarizes the class distribution and prediction for training observations satisfying the same rules.

:::

## Task 3.2 — Compare two forms of inherent interpretability

::: {.direction}

Compare the logistic regression and decision tree.

| Question | Logistic regression | Decision tree |
|---|---|---|
| How is the model represented? | | |
| How can we obtain a global understanding? | | |
| How can we explain one prediction? | | |
| What could make the model difficult to interpret? | | |

:::

::: {.sol}

```{=latex}
\begingroup
\footnotesize
```

| Question | Logistic regression | Decision tree |
|---|---|---|
| How is the model represented? | Additive equation on the log-odds scale | Hierarchy of if/then rules |
| How can we obtain a global understanding? | Inspect all coefficients | Inspect the complete tree |
| How can we explain one prediction? | Combine the applicant's values with coefficient directions and magnitudes | Follow the applicant's decision path |
| What could make the model difficult to interpret? | Many predictors, transformations, interactions, or nonlinear terms | A deep or very large tree |

```{=latex}
\endgroup
```

Both models expose their internal logic. They differ in the form in which that logic is represented.

:::

# Part 4 — Global post-hoc explanation: permutation importance {#importance}

The random forest contains hundreds of decision trees. Instead of inspecting the complete model, we now use **model-agnostic post-hoc explanations**.

DALEX provides a common interface for explaining different predictive models. An `explainer` connects a fitted model, data, observed outcomes, and its prediction function; `DALEXtra` integrates this interface with tidymodels. The random-forest explainer `explainer_rf` has already been prepared for you. Explanations use a held-out test sample so their summaries are not based on the same cases used to fit the model.

| Method | Scope | Main question | DALEX function |
|----|--|-----|---|
| Permutation importance | Global | Which variables does the model rely on? | `model_parts()` |
| Partial dependence | Global | How does the average prediction change with X? | `model_profile()` |
| Individual profile | Local | How would this prediction change if X changed? | `predict_profile()` |
| SHAP | Local | Which variables contributed to this prediction? | `predict_parts()` |

Permutation feature importance asks:

> How much worse does the model become when the information in one predictor is destroyed?

The predictor is randomly shuffled while the fitted model remains unchanged.
For classification, DALEX uses `1 - AUC` as its default loss; with `type = "difference"`, importance is the AUC lost relative to the full model. `_full_model_` is the unpermuted loss, `_baseline_` is the loss after all predictors are removed, and the boxplots show variation across repeated permutations.

## Task 4.1 — Calculate permutation importance

:::: {.direction}

Use `DALEX::model_parts()` with the prepared `explainer_rf`.

Use `B = 10` permutations. Then plot the result.

::: {.callout-tip title="Useful documentation"}
See the [DALEX model-parts documentation](https://modeloriented.github.io/DALEX/reference/model_parts.html).
:::

```{r}
#| label: task-4-1-student
set.seed(42)

# YOUR CODE HERE
```

Which predictors does the random forest rely on most strongly?

**Your answer:** *Write your answer here.*

::::

::: {.sol}

```{r}
#| label: task-4-1-solution
#| warning: false

set.seed(42)

importance <- DALEX::model_parts(
  explainer_rf,
  type = "difference",
  B = 10
)

plot(importance)
importance
```

Variables with a larger deterioration in model performance after permutation are more important to the random forest's predictions.

**Answer.** In the reference run the forest relies most on `missed_payments_12m` (AUC loss of about 0.11), then `credit_utilization` (about 0.06) and `debt_to_income` (about 0.03). The importances of `customer_category`, `income` and `months_with_company` are around zero or slightly negative, with whiskers that span zero. A negative value means that shuffling the variable happened to improve AUC marginally, which is noise: the forest does not detectably rely on these variables. This contrasts with the logistic regression, where `income` has a clear coefficient. Only the top one or two variables are ranked reliably given `B = 10` permutations on 500 rows.

:::

## Task 4.2 — Interpret importance carefully

::: {.direction}

Does permutation importance tell you whether increasing a variable increases or decreases default risk?

**Your answer:** *Write your answer here.*

Does it show that the variable causes default?

**Your answer:** *Write your answer here.*

Does it show that the variable is important for this particular applicant?

**Your answer:** *Write your answer here.*

What could happen to importance estimates when two predictors contain very similar information?

**Your answer:** *Write your answer here.*

:::

::: {.sol}

Permutation importance measures **global predictive reliance**. It does not provide the direction of an effect, establish causality, or explain one particular applicant.

When predictors are strongly correlated or redundant, one variable may replace information removed from another. Importance can therefore be distributed across correlated predictors or appear smaller than expected. Moreover, independently permuting one variable can create unrealistic predictor combinations.

:::

{{< needspace 20 >}}

# Part 5 — From average effects to individual what-if profiles {#profiles}

Feature importance tells us **whether** a variable matters, but not **how predictions change** as the variable changes.

We investigate two complementary views:

- a **partial-dependence profile (PDP)** summarizes average model predictions;
- an **individual profile** changes one predictor for one applicant while holding the applicant's other characteristics constant.

A PDP averages many individual what-if profiles and is therefore global. An individual profile instead keeps one observation fixed and varies one predictor at a time.

## Task 5.1 — Partial-dependence profiles

:::: {.direction}

Create partial-dependence profiles for `debt_to_income` and `credit_utilization`.

Use `N = 200` observations to keep computation fast.

::: {.callout-tip title="Useful documentation"}
See [Partial-dependence profiles](https://ema.drwhy.ai/partialDependenceProfiles.html). A minimal call is `DALEX::model_profile(explainer_rf, variables = "debt_to_income", type = "partial")`; use `plot(..., geom = "profiles")` to reveal individual lines under the average.
:::

```{r}
#| label: task-5-1-student
set.seed(42)

# YOUR CODE HERE
```

For each predictor, describe the overall direction of the relationship, whether it appears approximately linear, and any regions where predictions change particularly strongly.

**Your answer:** *Write your answer here.*

::::

::: {.sol}

```{r}
#| label: task-5-1-solution
#| out-width: 80%

set.seed(42)

partial_profiles <- DALEX::model_profile(
  explainer_rf,
  variables = c(
    "debt_to_income",
    "credit_utilization"
  ),
  type = "partial",
  N = 200
)

plot(partial_profiles)
```

The profiles show how the random forest's prediction changes **on average** when the selected predictor is varied while averaging over the remaining observations.

Unlike logistic-regression coefficients, the resulting relationship does not have to be linear or monotonic. PDPs can mislead when predictors are correlated because they may average predictions for unrealistic combinations.

**Answer.** Both profiles trend upward: higher `credit_utilization` and higher `debt_to_income` are associated with a higher average predicted default probability. Neither is linear. `credit_utilization` rises in steps, with a pronounced jump around 0.75 (matching the tree's split at 0.75). `debt_to_income` is roughly flat until about 0.6 and then climbs, levelling off at the highest values. Small wiggles are probably sampling noise (`N = 200`) and should not be interpreted.

:::

## Task 5.2 — Explain the same applicant with an individual profile

:::: {.direction}

Now calculate individual profiles for the same two variables and the selected `applicant`.

A **counterfactual explanation** seeks the smallest feasible change that flips a decision; a **ceteris-paribus profile** varies one variable while holding the others fixed. Here we create the latter.

::: {.callout-tip title="Useful documentation"}
See [Ceteris-paribus / individual profiles](https://ema.drwhy.ai/ceterisParibus.html). Use `predict_profile()` with arguments `new_observation` and `variables`.
:::

```{r}
#| label: task-5-2-student
# YOUR CODE HERE
```

Overlay the applicant profile and the PDP when comparing them. Where do they answer different questions?

**Your answer:** *Write your answer here.*

Suppose the individual profile shows that lower credit utilization would reduce the model's predicted default probability. Why should this **not automatically be presented as advice** to the applicant?

**Your answer:** *Write your answer here.*

::::

::: {.sol}

```{r}
#| label: task-5-2-solution
#| out-width: 80%
individual_profile <- DALEX::predict_profile(
  explainer_rf,
  new_observation = applicant,
  variables = c(
    "debt_to_income",
    "credit_utilization"
  )
)

plot(
  individual_profile,
  variables = c(
    "debt_to_income",
    "credit_utilization"
  )
) +
  ingredients::show_aggregated_profiles(
    partial_profiles$agr_profiles,
    variables = c(
      "debt_to_income",
      "credit_utilization"
    ),
    size = 1
  )
```

The PDP asks how the model's **average prediction** changes as a predictor changes. The individual profile asks how **this applicant's prediction** changes when one predictor changes while the other characteristics remain fixed.

Neither establishes causality. One-at-a-time changes can also create unrealistic combinations of characteristics, so a model-based what-if result is not automatically valid real-world advice or actionable recourse.

**Answers.**

- **Different questions:** the PDP asks, across applicants, how the average prediction changes with a variable. The individual profile asks how *this* applicant's prediction would change with everything else fixed. For this applicant the individual curve lies mostly above the PDP, because other characteristics (two missed payments) put them above the average applicant, and it is more irregular because it is not smoothed by averaging. Where the two curves differ, the average effect is a poor description of this person.
- **Not automatically advice:** the profile describes the model's behavior, not a causal effect, so reducing utilization may not lower the applicant's actual risk. Changing one variable alone may be unrealistic or infeasible, since utilization, debt-to-income and missed payments are linked. The curve is also an estimate with noise, and presenting it as advice invites gaming the score and creates expectations the bank may not honor.

:::

{{< needspace 20 >}}

# Part 6 — Local post-hoc explanation with SHAP {#shap}

We now ask:

> Why did the random forest assign this particular probability to this applicant?

SHAP explains one prediction relative to a **baseline prediction**. Each feature receives a contribution showing how much it pushes the prediction above or below that baseline. Together, the contributions explain the difference between the baseline and the prediction.

:::: {.direction}

Suppose the model's **baseline probability of default is 20%**, while it predicts **55%** for one applicant.

SHAP distributes this difference across the applicant's feature values:

| Component | Contribution |
|---|---:|
| **Baseline prediction** | **20%** |
| High debt-to-income ratio | +18 pp |
| High credit utilization | +12 pp |
| Long customer relationship | −4 pp |
| Other features | +9 pp |
| **Prediction** | **55%** |

$$\text{Prediction} = \text{Baseline} + \sum_{j=1}^{p} \text{SHAP}_j$$

In this example:

$$55\% = 20\% + 18\% + 12\% - 4\% + 9\%$$

**Interpretation:** Positive SHAP values push this prediction **above** the baseline; negative values push it **below** the baseline.

SHAP explains the **model's prediction**, not the causal effect of changing a feature.

::::


## Task 6.1 — Calculate a SHAP explanation

:::: {.direction}

DALEX approximates Shapley values by sampling feature orderings.

Use `DALEX::predict_parts()` with:

- `new_observation = applicant`;
- `type = "shap"`;
- `B = 25`.

Here, `B = 25` controls the number of sampled orderings used in the approximation. Plot the result and print its baseline. The baseline is the mean prediction over `explanation_data`; check that baseline plus the summed contributions is approximately the applicant probability. Boxplots show variation across the `B` sampled orderings, and results can change with the seed.

::: {.callout-tip title="Useful documentation"}
See the [conceptual Shapley-value documentation](https://ema.drwhy.ai/shapley.html) and the DALEX [`predict_parts()` reference](https://modeloriented.github.io/DALEX/reference/predict_parts.html).
:::

```{r}
#| label: task-6-1-student
set.seed(42)

# YOUR CODE HERE
```

Which applicant characteristics push the prediction most strongly toward default and away from default?

**Your answer:** *Write your answer here.*

::::

::: {.sol}

```{r}
#| label: task-6-1-solution
set.seed(42)

shap_explanation <- DALEX::predict_parts(
  explainer_rf,
  new_observation = applicant,
  type = "shap",
  B = 25
)

plot(shap_explanation)
shap_explanation

shap_baseline <- mean(explainer_rf$y_hat)
shap_check <- shap_explanation |>
  filter(!variable %in% c("_baseline_", "_full_model_")) |>
  group_by(variable) |>
  summarise(contribution = mean(contribution), .groups = "drop") |>
  summarise(reconstructed_probability = shap_baseline + sum(contribution))

shap_baseline
shap_check
predict_default_probability(rf_fit, applicant)
```

Positive contributions move the prediction above the baseline prediction; negative contributions move it below the baseline. The boxplots summarize variation across sampled orderings, so a different seed can change the approximation.

The SHAP explanation is **local**: it explains this prediction rather than the behavior of the model as a whole.

**Answer.** In the reference run, `missed_payments_12m = 2` pushes the prediction most strongly toward default (about +0.15 on the probability scale). `credit_utilization = 0.545` (about −0.10) and `debt_to_income = 0.536` (about −0.03) push it away from default; `months_with_company` and `customer_category` have small negative contributions (about −0.01), and `income` is close to zero. The baseline (the mean predicted probability, about 0.49) plus the contributions reproduces the applicant's 0.50. The upward push of the missed payments is largely offset by the other variables, which is why this is a borderline case.

:::

## Task 6.2 — Compare the local explanations

::: {.direction}

Compare the explanations produced for this applicant by logistic regression, the decision tree, and SHAP applied to the random forest. Logistic-regression contributions use the log-odds scale while SHAP contributions use the probability scale, so their numerical values are not directly comparable.

| Method | Top three drivers for this applicant |
|---|---|
| Logistic regression | |
| Decision tree | |
| SHAP | |

Do the three approaches emphasize broadly similar characteristics?

**Your answer:** *Write your answer here.*

Which explanation would be easiest to communicate to a nontechnical applicant?

**Your answer:** *Write your answer here.*

Does the SHAP explanation tell us what **caused** the applicant's predicted default risk?

**Your answer:** *Write your answer here.*

:::

::: {.sol}

```{r}
#| label: task-6-2-solution
shap_top <- shap_explanation |>
  filter(!variable %in% c("_baseline_", "_full_model_")) |>
  group_by(variable) |>
  summarise(contribution = mean(contribution), .groups = "drop") |>
  arrange(desc(abs(contribution))) |>
  slice_head(n = 3)

tree_path_variables <- tree_path |>
  setdiff("root") |>
  str_extract("^[A-Za-z0-9_]+") |>
  unique()

tibble(
  method = c("Logistic regression", "Decision tree", "SHAP"),
  top_drivers = c(
    term_contributions |>
      slice_head(n = 3) |>
      mutate(
        label = sprintf("%s (%+.2f log-odds)", term, contribution_default_log_odds)
      ) |>
      pull(label) |>
      paste(collapse = "; "),
    paste(tree_path_variables, collapse = ", "),
    shap_top |>
      mutate(label = sprintf("%s (%+.2f)", variable, contribution)) |>
      pull(label) |>
      paste(collapse = "; ")
  )
) |>
  knitr::kable()
```

1. **Agreement:** All three methods put `missed_payments_12m` first, pushing toward default, so the dominant driver is robust across models. Beyond that they diverge. The logistic regression adds `customer_category` and `income` (below-average income raises risk). SHAP gives `income` almost no weight, consistent with its near-zero permutation importance, while the tree uses `income` at its last split. `credit_utilization` pushes toward lower risk in both the logistic regression (about −0.10) and SHAP (about −0.10). `debt_to_income` is negligible in the logistic regression, slightly negative in SHAP and absent from the tree path. These differences are expected: the models represent relationships differently, predictors are correlated, and a tree path only shows the variables needed to reach one leaf. Compare ranks and directions, not magnitudes across the log-odds and probability scales.

2. **Communication:** The tree path is easiest to communicate to a nontechnical applicant because it can be stated as a short sequence of plain-language rules, for example: "Applicants with two missed payments in the last year and an income below 3,542 defaulted in about six out of ten historical cases." One caveat is that thresholds simplify relationships and a shallow tree may be less accurate than the random forest.

3. **Causality:** SHAP attributes a **model prediction** relative to its baseline. It does not establish that the attributed characteristics caused default risk.

:::

# Wrap-up — Match the explanation to the question {#wrap-up}

::: {.direction}

**Model-specific** methods use a model's internal structure; **model-agnostic** methods explain it through inputs and predictions.

Complete the comparison.

| Method | Global or local? | Model-specific? | Main question answered | Important limitation |
|---|---|---|---|---|
| Logistic coefficients | | | | |
| Decision tree | | | | |
| Permutation importance | | | | |
| Partial dependence | | | | |
| Individual profile | | | | |
| SHAP | | | | |

Finally, select the explanation you would prioritize for each stakeholder.

| Stakeholder | Most useful explanation(s) | Why? |
|---|---|---|
| Credit analyst reviewing one case | | |
| Applicant affected by a decision | | |
| Model-governance team | | |

Compare `applicant_outcome` with the three predictions from Task 1.1. What does this case suggest, and why can one case prove little about model quality?

**Your answer:** *Write your answer here.*

Optional discussion: How should regulation, contestability, and governance shape which explanations the organization records and communicates?

**Your answer:** *Write your answer here.*

:::

::: {.sol}

```{=latex}
\begingroup
\footnotesize
```

| Method | Global or local? | Model-specific? | Main question answered | Important limitation |
|--|--|---|----|----|
| Logistic coefficients | Primarily global | Yes | How are predictors associated with the model's log-odds? | Assumes the specified functional form; coefficient interpretation depends on units and reference categories |
| Decision tree | Global and local | Yes | Which explicit rules produce predictions? | Large trees quickly become difficult to understand and can be unstable |
| Permutation importance | Global | No | Which predictors does the model rely on? | Gives no direction or causality; correlated predictors complicate interpretation |
| Partial dependence | Global | No | How does the average prediction vary with a predictor? | Can hide heterogeneous effects and create unrealistic combinations |
| Individual profile | Local | No | How would one prediction vary when one predictor changes? | One-at-a-time changes may be unrealistic and are not causal |
| SHAP | Local | DALEX sampling: no; TreeSHAP: yes | Which variables contribute to this prediction relative to a baseline? | Attribution depends on the model and background distribution and is not causal |

```{=latex}
\endgroup
```

{{< pagebreak >}}

Possible stakeholder choices:

```{=latex}
\begingroup
\footnotesize
```

| Stakeholder | Most useful explanation(s) | Why? |
|---|---|---|
| Credit analyst reviewing one case | SHAP, individual profile, or decision path | Provides case-specific evidence that can support review |
| Applicant affected by a decision | Simplified local explanation derived from SHAP/tree plus carefully validated what-if information | Focuses on factors relevant to the individual and can support understanding or contestability |
| Model-governance team | Permutation importance, PDP, model structure, plus performance and monitoring evidence | Requires a global view of model behavior rather than only individual predictions |

```{=latex}
\endgroup
```

```{r}
#| label: wrap-up-solution
tibble(
  model = c("Logistic regression", "Decision tree", "Random forest"),
  default_probability = c(
    predict_default_probability(logistic_fit, applicant),
    predict_default_probability(tree_fit, applicant),
    predict_default_probability(rf_fit, applicant)
  ),
  predicted_class = if_else(default_probability >= 0.5, "default", "no_default"),
  actual_outcome = as.character(applicant_outcome)
) |>
  mutate(correct = predicted_class == actual_outcome)
```

All three models predict `default` in the reference run (the random forest only just, at about 0.50), so they are either all right or all wrong, as shown in the `correct` column. Either way, one case proves little. This applicant was chosen because the forest's probability is near 0.5, so the outcome is close to a coin flip by construction: a miss would not show the model is bad, and a hit would not show it is good. Model quality has to be judged with test-set metrics (such as the AUC from Task 1.3), calibration, the costs of the different error types and, ideally, cross-validation.

Regulation, contestability, and governance may require auditable explanations, review and appeal routes, documented limitations, and monitoring.

The central lesson is that **explainability is question-dependent**. There is no single explanation that is appropriate for every model, stakeholder, or deployment decision.

:::
