---
title: "Session 5: ML Baselines (Classification)"
session_id: session-05
format:
  html: default
---

::: {.direction}

In this exercise, we cover:

| Part | Topic | Time |
|---|---|---:|
| 1 | Case understanding and first model | 15 min |
| 2 | Probabilities and thresholds | 10 min |
| 3 | Confusion matrix and metrics | 10 min |
| 4 | ROC and AUC | 10 min |
| 5 | Comparing customer categories | 10 min |
| Wrap-up | Reflection | 5 min |
| **Overall** | | **60 min** |

Work directly in this Quarto (`.qmd`) exercise using R in RStudio, Positron, or VS Code.

# Case: Rising defaults at LendWise

You are a business analyst working for LendWise, a fintech company that offers fast digital consumer loans.
Over the last six months, LendWise has seen a noticeable spike in credit defaults in two customer categories:

- **Early-career urban borrowers:** typically younger customers, often with lower savings, moderate income, and irregular repayment histories.
- **Self-employed platform workers:** customers with fluctuating income, often higher short-term debt exposure, and less predictable monthly cash flows.

The risk team wants to understand whether default risk follows the same logic in both categories, or whether the predictors and model quality differ substantially.

You first focus on early-career urban borrowers and build a logistic regression model to predict default. Afterward, you compare the results with platform workers.

LendWise is interested in questions such as:

- Which factors are associated with a higher probability of default in early-career urban borrowers?
- How can predicted probabilities be translated into default/non-default classifications?
- How well does the model perform, and how informative are the ROC curve and AUC?
- Does a model work equally well across customer categories?

## Data description

- `customer_category`: customer segment
- `income`: monthly income in EUR
- `debt_to_income`: debt-to-income ratio
- `missed_payments_12m`: number of missed payments in the last 12 months
- `credit_utilization`: share of available credit currently used (between 0 and 1)
- `months_with_company`: months since first activity with LendWise
- `default`: whether the customer defaulted

Throughout the exercise, **`default = 1` means default** and **`default = 0` means no default**. This convention determines how predicted classes, confusion matrices, and business decisions are interpreted.

The dataset was curated by the company's Risk Analytics Department and has already been cleaned, validated, and prepared for modeling.

:::

```{r}
library(tidyverse)
library(pROC)

df <- read_csv(
  "https://raw.githubusercontent.com/fs-ise/machine-learning-for-big-data/main/exercises/data/credit_default.csv",
  show_col_types = FALSE
)

glimpse(df)
```

# Part 1 — Analyze early-career urban borrowers with logistic regression

::: {.direction}
In this first part, focus only on early-career urban borrowers.
:::

### Task 1.1 — Prepare the subset

Create a filtered data frame containing only early-career urban borrowers.

::: {.direction}

```{r}
# Add your solution below.
```
:::

::: {.sol}

## Solution

```{r}
df_ec_urban <- df |>
  filter(customer_category == "early_career_urban")

glimpse(df_ec_urban)
```
:::

### Task 1.2 — Estimate the model

Estimate a logistic regression model for credit default.

::: {.direction}

- Use `default` as the outcome.
- Use `income`, `debt_to_income`, `missed_payments_12m`, `credit_utilization`, and `months_with_company` as predictors.
- Fit the model with `glm(..., family = binomial)`.
- Inspect `coef(model)` and `summary(model)`.

```{r}
# Add your solution below.
```
:::

::: {.sol}

## Solution

```{r}
model <- glm(
  default ~
    income +
    debt_to_income +
    missed_payments_12m +
    credit_utilization +
    months_with_company,
  data = df_ec_urban,
  family = binomial
)

coef(model)
summary(model)
```

`glm()` estimates the unregularized maximum-likelihood logistic model. Its summary includes coefficient estimates, standard errors, z-statistics, and p-values.
:::

### Task 1.3 — Interpret the model

Logistic-regression coefficients describe changes in **log-odds**, holding the other predictors constant. Exponentiating them produces **odds ratios**: the factor by which a one-unit increase multiplies the odds of default.

::: {.sol}

```{r}
coefficient_table <- tibble(
  predictor = names(coef(model)),
  estimate = coef(model),
  odds_ratio = exp(coef(model))
)

coefficient_table
```
:::

- Which variables have positive associations with default?

::: {.sol}
::: {.callout-tip title="Solution"}
`missed_payments_12m` and `credit_utilization` have positive coefficients. Holding the other predictors constant:

- one additional missed payment multiplies the estimated odds of default by about **3.06** (coefficient 1.119, p < 0.001);
- a one-unit increase in credit utilization—from 0 to 1—multiplies the estimated odds by about **6.47** (coefficient 1.867, p ≈ 0.019). A smaller change, such as 0.10, has a more realistic interpretation.
:::
:::

- Which variables have negative associations with default?

::: {.sol}
::: {.callout-tip title="Solution"}
`income`, `debt_to_income`, and `months_with_company` have negative estimates. The evidence is weak at the 5% level for these three predictors: income is close (p ≈ 0.067), while debt-to-income and tenure have larger p-values. The debt-to-income estimate is very close to zero relative to its uncertainty.
:::
:::

- Can the largest raw coefficient automatically be called the most important predictor?

::: {.sol}
::: {.callout-tip title="Solution"}
No. Predictors use different units: income is measured in euros, missed payments as a count, and utilization as a proportion. Raw coefficient magnitudes are therefore not directly comparable. Interpret directions, uncertainty, odds ratios, and changes in substantively meaningful units instead.
:::
:::

### Task 1.4 — From coefficients to classification

Explain how the model moves from a linear combination of predictors to a predicted probability and then a default/non-default class.

::: {.sol}

## Solution

The sequence is:

1. predictor values are combined into a linear predictor, or log-odds;
2. `plogis()` transforms log-odds into a probability;
3. a chosen threshold converts the probability into a class.

```{r}
z <- predict(
  model,
  newdata = df_ec_urban[1, ],
  type = "link"
)

p <- plogis(z)
y_class <- if_else(p >= 0.5, 1L, 0L)

z
p
y_class
```

The first observation's predicted default probability is approximately **48.1%**, so it is classified as no default at a 50% threshold.
:::
# Part 2 — Probabilities and thresholds

::: {.direction}
Logistic regression produces probabilities. A classification is a separate decision created by comparing those probabilities with a threshold.

Because the modeled event is `default = 1`, the returned values estimate **P(default = 1)**.
:::

### Task 2.1 — Understanding outputs

Run the code and compare the probabilities with the classes.

```{r}
y_prob <- predict(
  model,
  type = "response"
)

threshold <- 0.5

y_pred <- if_else(
  y_prob >= threshold,
  1L,
  0L
)

head(y_prob)
head(y_pred)
```

- What is the difference between `y_prob` and `y_pred`?
- Why is it useful to define the threshold explicitly?

::: {.sol}

- `y_prob` contains estimated probabilities of default.
- `y_pred` contains decisions based on the chosen 0.5 threshold.
- Writing the threshold explicitly makes clear that the fitted model estimates risk while the organization chooses how that risk becomes a decision.
:::

### Task 2.2 — Manual thresholding

::: {.direction}
Try thresholds of 0.2, 0.4, 0.6, and 0.7:

```{r}
t <- 0.4

y_pred_t <- if_else(
  y_prob >= t,
  1L,
  0L
)

sum(y_pred_t)
```

- How does the number classified as high default risk change?
- What pattern appears as the threshold increases?

Then identify the 20% riskiest and 20% safest customers using `quantile()`.

```{r}
# Add your solution below.
```
:::

::: {.sol}

## Solution

At thresholds 0.2, 0.4, 0.6, and 0.7, respectively, the model flags **138, 86, 52, and 31** customers as risky.

- A **lower threshold** flags more customers as risky.
- A **higher threshold** flags fewer customers as risky.

```{r}
threshold_risky <- quantile(
  y_prob,
  probs = 0.80
)

risky_20 <- y_prob >= threshold_risky

threshold_safe <- quantile(
  y_prob,
  probs = 0.20
)

safe_20 <- y_prob <= threshold_safe

risky_customers <- df_ec_urban |>
  mutate(
    predicted_default_probability = y_prob,
    risky_20 = risky_20
  ) |>
  filter(risky_20)

safe_customers <- df_ec_urban |>
  mutate(
    predicted_default_probability = y_prob,
    safe_20 = safe_20
  ) |>
  filter(safe_20)

threshold_risky
threshold_safe
risky_customers |> slice_head(n = 5)
safe_customers |> slice_head(n = 5)
```

The 80th-percentile risk threshold is approximately **0.652**, and the 20th-percentile threshold is approximately **0.151**.
:::

::: {.callout-important title="Threshold and approval logic"}
Assume customers classified as likely defaults are not automatically approved:

- **Lower risk threshold** → more customers flagged as risky → fewer approvals and lower default exposure.
- **Higher risk threshold** → fewer customers flagged as risky → more approvals and potentially more defaults.

Changing the threshold does not change estimated probabilities; it changes how the organization acts on them.
:::

# Part 3 — Confusion matrix and metrics

::: {.callout-note}
For simplicity, predictions are evaluated on the same observations used for estimation. This deliberately optimistic in-sample evaluation is sufficient for introducing metrics. Train/test splits, cross-validation, and out-of-sample evaluation follow in the machine-learning workflow session.
:::

### Task 3.1 — Create and interpret the matrix

::: {.direction}
Create a labeled 2 × 2 table. Rows should be actual outcomes and columns predicted outcomes.

```{r}
# Add your solution below.
```

Identify true negatives, false positives, false negatives, and true positives.
:::

::: {.sol}

```{r}
cm <- table(
  actual = df_ec_urban$default,
  predicted = y_pred
)

cm

tn <- cm["0", "0"]
fp <- cm["0", "1"]
fn <- cm["1", "0"]
tp <- cm["1", "1"]

c(tn = tn, fp = fp, fn = fn, tp = tp)
```

At the 0.5 threshold:

- TN: **91**
- FP: **22**
- FN: **32**
- TP: **48**

A false positive is a customer predicted to default who does not; this could mean rejecting or intervening on a good customer. A false negative is predicted not to default but actually defaults; this could mean approving a risky loan.
:::

### Task 3.2 — Compute key metrics

::: {.direction}
Compute and interpret accuracy, precision, and recall. Recall is also called sensitivity or the true-positive rate.
:::

::: {.sol}

```{r}
accuracy <- (tp + tn) / sum(cm)
precision <- tp / (tp + fp)
recall <- tp / (tp + fn)

c(
  accuracy = accuracy,
  precision = precision,
  recall = recall
)
```

- Accuracy: approximately **0.720**—72.0% of classifications are correct.
- Precision: approximately **0.686**—68.6% of customers flagged as risky actually default.
- Recall/sensitivity/TPR: **0.600**—60.0% of actual defaults are flagged.
:::

### Task 3.3 — Business reflection

LendWise is trying to grow its customer base aggressively.

- Why might the company accept more false negatives temporarily?

::: {.sol}
::: {.callout-tip title="Solution"}
An aggressive growth strategy may use a **higher default-risk threshold**. Fewer customers are flagged, so more may be approved, but more actual defaults can be missed. The resulting growth comes with greater credit risk.
:::
:::

- Why might LendWise lower the threshold during an economic downturn?

::: {.sol}
::: {.callout-tip title="Solution"}
A risk-control strategy may use a **lower threshold**. More customers are flagged, reducing false negatives and default exposure, at the cost of more false positives and potentially fewer approvals.
:::
:::

# Part 4 — ROC curve and AUC

::: {.direction}
Use `pROC` to evaluate discrimination across all possible thresholds. Here, higher predicted probabilities should correspond to `default = 1`, so `direction = "<"` states that controls tend to have lower scores than cases.

```{r}
# Add your solution below.
```
:::

::: {.sol}

```{r}
roc_ec <- roc(
  response = df_ec_urban$default,
  predictor = y_prob,
  levels = c(0, 1),
  direction = "<",
  quiet = TRUE
)

auc_ec <- auc(roc_ec)

roc_ec_df <- tibble(
  false_positive_rate = 1 - roc_ec$specificities,
  true_positive_rate = roc_ec$sensitivities
)

ggplot(
  roc_ec_df,
  aes(
    x = false_positive_rate,
    y = true_positive_rate
  )
) +
  geom_path() +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed"
  ) +
  coord_equal() +
  labs(
    title = "ROC curve: early-career urban borrowers",
    subtitle = paste("AUC =", round(as.numeric(auc_ec), 3)),
    x = "False-positive rate",
    y = "True-positive rate"
  )

auc_ec
```

The AUC is approximately **0.801**.
:::

### Task 4.1 — Understanding ROC

- What does the ROC curve represent?
- What are TPR and FPR?

::: {.sol}

- The ROC curve shows the trade-off between the true-positive rate and false-positive rate across thresholds.
- TPR, or recall/sensitivity, is `TP / (TP + FN)`.
- FPR is `FP / (FP + TN)`.
:::

### Task 4.2 — Interpreting AUC

- What does AUC represent?
- What would values near 0.5, 0.7, and 1.0 suggest?

::: {.sol}

AUC summarizes how well predicted scores rank defaults above non-defaults across thresholds.

- 0.5 indicates no discrimination beyond random ranking.
- Around 0.7 indicates moderate discrimination.
- 1.0 indicates perfect discrimination on the evaluated data.

AUC is not a complete measure of business utility and does not determine the operating threshold.
:::

### Task 4.3 — Threshold perspective

- Why is ROC/AUC useful before the company fixes one decision threshold?

::: {.sol}
ROC/AUC evaluates ranking performance across thresholds and supports model comparison before selecting a cutoff. The eventual threshold must still reflect the unequal business costs of false positives and false negatives.
:::
# Part 5 — Compare early-career urban borrowers and platform workers

Repeat the same model specification for platform workers, then compare coefficients and discrimination.

### Task 5.1 — Fit the second model

::: {.direction}

```{r}
# Add your solution below.
```
:::

::: {.sol}

## Solution

```{r}
df_platform <- df |>
  filter(customer_category == "platform_worker")

model_platform <- glm(
  default ~
    income +
    debt_to_income +
    missed_payments_12m +
    credit_utilization +
    months_with_company,
  data = df_platform,
  family = binomial
)

coef(model_platform)
summary(model_platform)
```
:::

### Task 5.2 — Compare coefficients

- Are directions and statistical evidence the same in both groups?
- How do meaningful unit changes or odds ratios differ?

::: {.direction}

```{r}
# Add your solution below.
```
:::

::: {.sol}

```{r}
comparison <- tibble(
  predictor = names(coef(model)),
  early_career_urban_estimate = coef(model),
  platform_worker_estimate = coef(model_platform),
  early_career_urban_odds_ratio = exp(coef(model)),
  platform_worker_odds_ratio = exp(coef(model_platform))
)

comparison
```

For both segments, `credit_utilization` and `missed_payments_12m` have positive, statistically significant associations with default. Their exact estimates differ:

- credit utilization: **1.87** for early-career urban borrowers and **2.10** for platform workers;
- missed payments: **1.12** for early-career urban borrowers and **0.40** for platform workers.

Debt-to-income differs particularly strongly: it is near zero and not significant for early-career urban borrowers (**-0.05**, p ≈ 0.968), but positive and significant for platform workers (**2.18**, p ≈ 0.004).

Income and months with company are negative in both models but do not reach the 5% significance level. Because predictors use different units, raw coefficient size alone is not a valid measure of importance.
:::

### Task 5.3 — Compare ROC and AUC

- Which category has the higher in-sample AUC?
- What might explain the difference?

::: {.direction}

```{r}
# Add your solution below.
```
:::

::: {.sol}

```{r}
y_prob_platform <- predict(
  model_platform,
  type = "response"
)

roc_platform <- roc(
  response = df_platform$default,
  predictor = y_prob_platform,
  levels = c(0, 1),
  direction = "<",
  quiet = TRUE
)

auc_platform <- auc(roc_platform)

roc_comparison_df <- bind_rows(
  tibble(
    false_positive_rate = 1 - roc_ec$specificities,
    true_positive_rate = roc_ec$sensitivities,
    segment = "Early-career urban"
  ),
  tibble(
    false_positive_rate = 1 - roc_platform$specificities,
    true_positive_rate = roc_platform$sensitivities,
    segment = "Platform worker"
  )
)

ggplot(
  roc_comparison_df,
  aes(
    x = false_positive_rate,
    y = true_positive_rate,
    color = segment
  )
) +
  geom_path() +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed"
  ) +
  coord_equal() +
  labs(
    title = "ROC curves by customer segment",
    subtitle = paste(
      "AUC: early-career urban =",
      round(as.numeric(auc_ec), 3),
      "; platform worker =",
      round(as.numeric(auc_platform), 3)
    ),
    x = "False-positive rate",
    y = "True-positive rate",
    color = "Segment"
  )

c(
  early_career_urban = as.numeric(auc_ec),
  platform_worker = as.numeric(auc_platform)
)
```

The early-career urban model has the higher in-sample AUC: approximately **0.801**, compared with **0.718** for platform workers. One possible explanation is that the available predictors separate defaults more consistently in the early-career segment. AUC alone does not establish why the difference exists or whether it will persist out of sample.
:::

::: {.direction}
# 🧩 Wrap-up <a id="wrap-up"></a>

🎉🎈 You have completed the exercise—good work! 🎈🎉

In this exercise, we have learned to:

- Fit and interpret logistic regression models in R with `glm()`
- Interpret log-odds coefficients and odds ratios
- Generate predicted probabilities with `predict()`
- Convert probabilities into decisions using explicit thresholds
- Understand threshold trade-offs and their business costs
- Interpret confusion matrices and compute accuracy, precision, and recall
- Understand ROC curves and AUC
- Compare model behavior across customer segments

<!--
::: {.callout-tip title="Session 5 survey"}
{{< survey session_05 exercise >}}
:::
-->

:::
