---
title: "Session 2: Data Preparation"
session_id: session-02
format:
  html: default
---

:::{.direction}

In this exercise, we cover:

| Exercise part                          | Time (min) |
|----------------------------------------|------------|
| [Data structuring 1](#structuring_1)   |     15     |
| [Data structuring 2](#structuring_2)  |     20     |
| [Data cleansing](#cleansing)           |     20     |
| [Wrap-up](#wrap-up)                    |     10     |
| Overall                                |     65     |

<br>

> **Recommended workflow:**
>
> 1. ✏️ **Think & plan first:** Draft your answers and outline the transformation steps before coding.
> 2. 💻 **Then code:** Work directly in this Quarto (`.qmd`) exercise using R in RStudio, Positron, or VS Code.
>
> For data transformation guidance, refer to the [tidyr documentation](https://tidyr.tidyverse.org/).

{{< pagebreak >}}

:::

# Part 1: Data structuring 1 <a id="structuring_1"></a>

We start with the `market_data_log.csv` dataset.

```{r}
#| output: false

library(tidyverse)

market_data_log_df <- read_csv("data/market_data_log.csv", show_col_types = FALSE)
market_data_log_df |> head()
```

<div class="cell-output cell-output-display">
<div>
<table class="dataframe caption-top table table-sm table-striped small" data-quarto-postprocess="true" data-border="1">
<thead>
<tr class="header">
<th data-quarto-table-cell-role="th"></th>
<th data-quarto-table-cell-role="th">timestamp</th>
<th data-quarto-table-cell-role="th">logger_sys</th>
<th data-quarto-table-cell-role="th">Record</th>
</tr>
</thead>
<tbody>
<tr class="odd">
<td data-quarto-table-cell-role="th">0</td>
<td>2024-10-16 00:00:00</td>
<td>FIN-PIPE-V1</td>
<td>SAP-DE-Tech-2023Q1-MarketCap-1154.9B</td>
</tr>
<tr class="even">
<td data-quarto-table-cell-role="th">1</td>
<td>2023-04-25 00:00:00</td>
<td>FIN-PIPE-V1</td>
<td>SAP-DE-Tech-2023Q1-Revenue-475.6B</td>
</tr>
<tr class="odd">
<td data-quarto-table-cell-role="th">2</td>
<td>2023-01-26 00:00:00</td>
<td>FIN-PIPE-V1</td>
<td>SAP-DE-Tech-2023Q1-EBITDA-146.7B</td>
</tr>
<tr class="even">
<td data-quarto-table-cell-role="th">3</td>
<td>2023-10-09 00:00:00</td>
<td>FIN-PIPE-V1</td>
<td>SAP-DE-Tech-2023Q2-MarketCap-1816.0B</td>
</tr>
<tr class="odd">
<td data-quarto-table-cell-role="th">4</td>
<td>2023-09-08 00:00:00</td>
<td>FIN-PIPE-V1</td>
<td>SAP-DE-Tech-2023Q2-Revenue-82.2B</td>
</tr>
</tbody>
</table>

</div>
</div>

:::{.direction}

### Tasks

1. Identify:

   * Values: ________________
   * Observations: ________________
   * Variables: ________________
   * Units of observation: ________________

2. Draft the **tidy structure** of the dataset:

   * Fill in the column names and at least 3 example rows.

<div class="cell-output cell-output-display">
<div>
<table style="width:100%; border-collapse:collapse; font-family:Arial, sans-serif; color:#222;">
  <thead>
    <tr>
      <th style="background:#f2f2f2; border:1px solid #ccc; padding:10px;">&nbsp;</th>
      <th style="background:#f2f2f2; border:1px solid #ccc; padding:10px;">&nbsp;</th>
      <th style="background:#f2f2f2; border:1px solid #ccc; padding:10px;">&nbsp;</th>
      <th style="background:#f2f2f2; border:1px solid #ccc; padding:10px;">&nbsp;</th>
      <th style="background:#f2f2f2; border:1px solid #ccc; padding:10px;">&nbsp;</th>
      <th style="background:#f2f2f2; border:1px solid #ccc; padding:10px;">&nbsp;</th>
      <th style="background:#f2f2f2; border:1px solid #ccc; padding:10px;">&nbsp;</th>
      <th style="background:#f2f2f2; border:1px solid #ccc; padding:10px;">&nbsp;</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
    </tr>

    <tr>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
    </tr>
    <tr>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
    </tr>
    <tr>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
    </tr>
  </tbody>
</table>
</div>
</div>


3. Describe how the dataset should be transformed.

<br><br><br><br>

4. **(Optional challenge)**: Implement the transformation in R (`tidyverse`)

```{r}
market_data_df <- read_csv("data/market_data_log.csv", show_col_types = FALSE)

# Add your solution below.
# Note: you may need string operations.
```

{{< iconify fluent-mdl2:insights >}} If you didn’t get to this part in class, try working through it at home.


::: {.callout-note title="Expectations in the exam"}
<br>

<table style="border-collapse: collapse; width: 100%; font-family: Arial, sans-serif;">
  <thead>
    <tr style="background-color: #f2f2f2;">
      <th style="border: 1px solid #ccc; padding: 8px;">Grade</th>
      <th style="border: 1px solid #ccc; padding: 8px;">Expectations</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td style="border: 1px solid #ccc; padding: 8px;"><strong>1 (Best)</strong></td>
      <td style="border: 1px solid #ccc; padding: 8px;">
        <strong>Recommend</strong> how to fix the dataset using appropriate transformations, e.g., describe restructuring steps, select the correct method from examples, or provide pseudocode.
      </td>
    </tr>
    <tr>
      <td style="border: 1px solid #ccc; padding: 8px;"><strong>2</strong></td>
      <td style="border: 1px solid #ccc; padding: 8px;">
        <strong>Evaluate</strong> and justify the decision using tidy data principles, e.g., "This is not tidy because column 2 contains two variables instead of one."
      </td>
    </tr>
    <tr>
      <td style="border: 1px solid #ccc; padding: 8px;"><strong>3</strong></td>
      <td style="border: 1px solid #ccc; padding: 8px;">
        <strong>Apply</strong> tidy data principles to evaluate a dataset (correctly identify whether it is tidy or not).
      </td>
    </tr>
    <tr>
      <td style="border: 1px solid #ccc; padding: 8px;"><strong>4 (Pass)</strong></td>
      <td style="border: 1px solid #ccc; padding: 8px;">
        <strong>Understand</strong> and reproduce the principles of tidy data (e.g., each variable in a column, each observation in a row).
      </td>
    </tr>
  </tbody>
</table>

<p style="font-family: Arial, sans-serif; margin-top: 10px;">
  <strong>Note:</strong> Even for the best grade, you are not expected to write fully functional code from memory. Still, practicing with code will help you build a deeper conceptual understanding.
</p>

:::

:::


::: {.sol}

## Solution

### 1. Identify:

* **Values:** Examples: 2024-10-16 00:00:00, FIN-PIPE-V1, SAP, DE, 2023Q1, 1154.9
* **Observations:** Individual logged measurements for a specific firm and period
* **Variables:** Timestamp, Logger system, Firm, Country, Sector, Period, MarketCap, Revenue, EBITDA
* **Units of observation:** One logged **firm–period** record (aggregating multiple metrics)

---

### 2. Draft the **tidy structure** of the dataset (wide format):

| timestamp           | logger_sys  | Firm | Country | Sector | Period | MarketCap | Revenue | EBITDA |
| ------------------- | ----------- | ---- | ------- | ------ | ------ | --------- | ------- | ------ |
| 2024-10-16 00:00:00 | FIN-PIPE-V1 | SAP  | DE      | Tech   | 2023Q1 | 1154.9B   | 475.6B  | 146.7B |
| 2023-10-09 00:00:00 | FIN-PIPE-V1 | SAP  | DE      | Tech   | 2023Q2 | 1816.0B   | 82.2B   |        |
| ...                 | ...         | ...  | ...     | ...    | ...    | ...       | ...     | ...    |

---

### 3. Describe how the dataset should be transformed.

* **Parse encoded column:**

  * The `Record` column contains multiple variables concatenated with `-`
  * Split into: `Firm`, `Country`, `Sector`, `Period`, `Metric`, `Value`

* **Reshape (long → wide):**

  * Pivot the `Metric` column into separate columns (`MarketCap`, `Revenue`, `EBITDA`, …)
  * Use `Value` as cell entries

* **Optional refinement:**

  * Convert `Value` into numeric format (e.g., remove `B` and cast to float)
  * Keep or drop `timestamp` depending on whether it represents data collection or observation time

---

### 4. (Optional challenge): Implement the transformation in R (`tidyverse`)

```{r}
market_data_df <- read_csv("data/market_data_log.csv", show_col_types = FALSE)

# Split the values encoded in Record into separate variables.
market_data_long <- market_data_df |>
  separate_wider_delim(
    Record,
    delim = "-",
    names = c("Firm", "Country", "Sector", "Period", "Metric", "Value")
  )

# Give each metric its own column.
df_wide <- market_data_long |>
  pivot_wider(names_from = Metric, values_from = Value) |>
  mutate(across(c(MarketCap, Revenue, EBITDA), parse_number))

df_wide |> head()
```

:::


---

{{< pagebreak >}}

# Part 2: Data structuring 2 <a id="structuring_2"></a>

Dataset `sales_data_channels_full_year.csv`

```{r}
#| output: false

sales_data_df <- read_csv("data/sales_data_channels_full_year.csv", show_col_types = FALSE)
sales_data_df |> head()
```

<div class="cell-output cell-output-display">
<div>
<table class="dataframe caption-top table table-sm table-striped small" data-quarto-postprocess="true" data-border="1">
<thead>
<tr class="header">
<th data-quarto-table-cell-role="th"></th>
<th data-quarto-table-cell-role="th">Customer_ID</th>
<th data-quarto-table-cell-role="th">Region</th>
<th data-quarto-table-cell-role="th">Jan_Online</th>
<th data-quarto-table-cell-role="th">Jan_Store</th>
<th data-quarto-table-cell-role="th">Feb_Online</th>
<th data-quarto-table-cell-role="th">Feb_Store</th>
<th data-quarto-table-cell-role="th">Mar_Online</th>
<th data-quarto-table-cell-role="th">Mar_Store</th>
<th data-quarto-table-cell-role="th">Apr_Online</th>
<th data-quarto-table-cell-role="th">Apr_Store</th>
<th data-quarto-table-cell-role="th">...</th>
<th data-quarto-table-cell-role="th">Aug_Online</th>
<th data-quarto-table-cell-role="th">Aug_Store</th>
<th data-quarto-table-cell-role="th">Sep_Online</th>
<th data-quarto-table-cell-role="th">Sep_Store</th>
<th data-quarto-table-cell-role="th">Oct_Online</th>
<th data-quarto-table-cell-role="th">Oct_Store</th>
<th data-quarto-table-cell-role="th">Nov_Online</th>
<th data-quarto-table-cell-role="th">Nov_Store</th>
<th data-quarto-table-cell-role="th">Dec_Online</th>
<th data-quarto-table-cell-role="th">Dec_Store</th>
</tr>
</thead>
<tbody>
<tr class="odd">
<td data-quarto-table-cell-role="th">0</td>
<td>ACC-E0IFD0</td>
<td>APAC</td>
<td>215</td>
<td>201</td>
<td>240</td>
<td>73</td>
<td>80</td>
<td>184</td>
<td>224</td>
<td>240</td>
<td>...</td>
<td>182</td>
<td>180</td>
<td>108</td>
<td>154</td>
<td>278</td>
<td>76</td>
<td>85</td>
<td>211</td>
<td>148</td>
<td>132</td>
</tr>
<tr class="even">
<td data-quarto-table-cell-role="th">1</td>
<td>ACC-DPBHSA</td>
<td>EU</td>
<td>315</td>
<td>108</td>
<td>251</td>
<td>163</td>
<td>133</td>
<td>245</td>
<td>170</td>
<td>194</td>
<td>...</td>
<td>300</td>
<td>157</td>
<td>289</td>
<td>233</td>
<td>173</td>
<td>121</td>
<td>172</td>
<td>213</td>
<td>260</td>
<td>158</td>
</tr>
<tr class="odd">
<td data-quarto-table-cell-role="th">2</td>
<td>CUST-ZZPQK5</td>
<td>APAC</td>
<td>142</td>
<td>167</td>
<td>195</td>
<td>81</td>
<td>175</td>
<td>107</td>
<td>296</td>
<td>210</td>
<td>...</td>
<td>101</td>
<td>221</td>
<td>141</td>
<td>116</td>
<td>142</td>
<td>109</td>
<td>182</td>
<td>55</td>
<td>258</td>
<td>124</td>
</tr>
<tr class="even">
<td data-quarto-table-cell-role="th">3</td>
<td>CUST-ZMF8MD</td>
<td>APAC</td>
<td>218</td>
<td>209</td>
<td>154</td>
<td>224</td>
<td>205</td>
<td>163</td>
<td>124</td>
<td>206</td>
<td>...</td>
<td>83</td>
<td>228</td>
<td>264</td>
<td>172</td>
<td>235</td>
<td>223</td>
<td>239</td>
<td>88</td>
<td>285</td>
<td>178</td>
</tr>
<tr class="odd">
<td data-quarto-table-cell-role="th">4</td>
<td>CUST-MMJBQE</td>
<td>EU</td>
<td>180</td>
<td>145</td>
<td>212</td>
<td>135</td>
<td>217</td>
<td>124</td>
<td>231</td>
<td>190</td>
<td>...</td>
<td>253</td>
<td>151</td>
<td>115</td>
<td>78</td>
<td>159</td>
<td>194</td>
<td>140</td>
<td>219</td>
<td>111</td>
<td>137</td>
</tr>
</tbody>
</table>

<p>5 rows × 26 columns</p>
</div>
</div>

:::{.direction}

### Tasks

1. Identify:

   * Values: ________________
   * Observations: ________________
   * Variables: ________________
   * Units of observation: ________________

2. Draft the **tidy structure** of the dataset:


<div class="cell-output cell-output-display">
<div>
<table style="width:100%; border-collapse:collapse; font-family:Arial, sans-serif; color:#222;">
  <thead>
    <tr>
      <th style="background:#f2f2f2; border:1px solid #ccc; padding:10px;">&nbsp;</th>
      <th style="background:#f2f2f2; border:1px solid #ccc; padding:10px;">&nbsp;</th>
      <th style="background:#f2f2f2; border:1px solid #ccc; padding:10px;">&nbsp;</th>
      <th style="background:#f2f2f2; border:1px solid #ccc; padding:10px;">&nbsp;</th>
      <th style="background:#f2f2f2; border:1px solid #ccc; padding:10px;">&nbsp;</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
    </tr>

    <tr>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
    </tr>

    <tr>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
    </tr>

    <tr>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
      <td contenteditable="true" style="border:1px solid #ccc; padding:10px;">&nbsp;</td>
    </tr>
  </tbody>
</table>
</div>
</div>


3. Describe how the dataset should be transformed.

<br><br><br>

4. **(Optional challenge)**: Implement the transformation in R (`tidyverse`)

```{r}
sales_data_df <- read_csv("data/sales_data_channels_full_year.csv", show_col_types = FALSE)

# Add your solution below.
```

{{< iconify fluent-mdl2:insights >}} In addition to the transformation, you may need string operations

:::

:::{.sol}

## Solution

### 1. Identify:

   * Values: Numerical sales values (215, 201, 240, ...)
   * Observations: Individual sales measurements for a specific customer, month, and channel
   * Variables: CustomerID, Region, Month, Channel, Sales
   * Units of observation: A customer–month–channel combination

### 2. Draft the **tidy structure** of the dataset:

| Customer_ID | Region | Month | Channel | Sales |
| ----------- | ------ | ----- | ------- | ----- |
| CUST-AB12CD | EU     | Jan   | Online  | 120   |
| CUST-AB12CD | EU     | Jan   | Store   | 95    |
| CUST-AB12CD | EU     | Feb   | Online  | 135   |


{{< pagebreak >}}

### 3. Describe how the dataset should be transformed.

- Reshape (wide → long): Convert all monthly/channel columns into two columns
- Split the combined column (Month, Channel)

### 4. (Optional challenge): Implement the transformation in R (`tidyverse`)

```{r}
sales_data_df <- read_csv("data/sales_data_channels_full_year.csv", show_col_types = FALSE)

# Step 1: reshape the month/channel columns to long format.
tidy_df <- sales_data_df |>
  pivot_longer(
    cols = -c(Customer_ID, Region),
    names_to = "Month_Channel",
    values_to = "Sales"
  ) |>
  # Step 2: split the combined column into Month and Channel.
  separate_wider_delim(
    Month_Channel,
    delim = "_",
    names = c("Month", "Channel")
  )

tidy_df |> head()
```

:::

---

{{< pagebreak >}}

# Part 3: Data cleansing <a id="cleansing"></a>

Dataset `messy_customer_data.csv`

```{r}
#| output: false

customer_df <- read_csv(
  "data/messy_customer_data.csv",
  na = c("", "NA", "NaN"),
  show_col_types = FALSE
)
customer_df |> head()
```

<div class="cell-output cell-output-display">
<div>
<table class="dataframe caption-top table table-sm table-striped small" data-quarto-postprocess="true" data-border="1">
<thead>
<tr class="header">
<th data-quarto-table-cell-role="th"></th>
<th data-quarto-table-cell-role="th">customer_id</th>
<th data-quarto-table-cell-role="th">date_of_birth</th>
<th data-quarto-table-cell-role="th">income</th>
<th data-quarto-table-cell-role="th">gender</th>
<th data-quarto-table-cell-role="th">signup_date</th>
<th data-quarto-table-cell-role="th">country</th>
</tr>
</thead>
<tbody>
<tr class="odd">
<td data-quarto-table-cell-role="th">0</td>
<td>CUST_00000001809</td>
<td>1980-04-09</td>
<td>4896 USD</td>
<td>M</td>
<td>2026-12-14</td>
<td>USA</td>
</tr>
<tr class="even">
<td data-quarto-table-cell-role="th">1</td>
<td>CUST_00000000695</td>
<td>1945-08-02</td>
<td>7029</td>
<td>F</td>
<td>2023-11-26</td>
<td>BR</td>
</tr>
<tr class="odd">
<td data-quarto-table-cell-role="th">2</td>
<td>CUST_00000000907</td>
<td>1971-07-25</td>
<td>6432</td>
<td>F</td>
<td>2024-06-25</td>
<td>Austria</td>
</tr>
<tr class="even">
<td data-quarto-table-cell-role="th">3</td>
<td>CUST_00000000545</td>
<td>2000-03-01</td>
<td>6107</td>
<td>M</td>
<td>2023-06-29</td>
<td>JP</td>
</tr>
<tr class="odd">
<td data-quarto-table-cell-role="th">4</td>
<td>CUST_00000001848</td>
<td>1943-09-06</td>
<td>7655</td>
<td>F</td>
<td>2027-01-22</td>
<td>Italy</td>
</tr>
<tr class="even">
<td data-quarto-table-cell-role="th">5</td>
<td>CUST_00000001567</td>
<td>1980-09-14</td>
<td>2244</td>
<td>F</td>
<td>2026-04-16</td>
<td>CH</td>
</tr>
<tr class="odd">
<td data-quarto-table-cell-role="th">6</td>
<td>CUST_00000001573</td>
<td>1950-04-07</td>
<td>3262</td>
<td>F</td>
<td>2026-04-22</td>
<td>India</td>
</tr>
<tr class="even">
<td data-quarto-table-cell-role="th">7</td>
<td>CUST_00000000945</td>
<td>1989-08-23</td>
<td>3976</td>
<td>M</td>
<td>2024-08-02</td>
<td>BR</td>
</tr>
<tr class="odd">
<td data-quarto-table-cell-role="th">8</td>
<td>CUST_00000000819</td>
<td>1946-03-09</td>
<td>3077 USD</td>
<td>F</td>
<td>2024-03-29</td>
<td>NL</td>
</tr>
<tr class="even">
<td data-quarto-table-cell-role="th">9</td>
<td>CUST_00000000830</td>
<td>1981-03-30</td>
<td>NaN</td>
<td>M</td>
<td>2024-04-09</td>
<td>BR</td>
</tr>
<tr class="odd">
<td data-quarto-table-cell-role="th">10</td>
<td>CUST_00000001531</td>
<td>1995-03-08</td>
<td>4667</td>
<td>F</td>
<td>2026-03-11</td>
<td>DE</td>
</tr>
<tr class="even">
<td data-quarto-table-cell-role="th">11</td>
<td>CUST_00000000979</td>
<td>1990-01-24</td>
<td>NaN</td>
<td>F</td>
<td>2024-09-05</td>
<td>Switzerland</td>
</tr>
<tr class="odd">
<td data-quarto-table-cell-role="th">12</td>
<td>CUST_00000001576</td>
<td>1989-06-18</td>
<td>2920</td>
<td>F</td>
<td>2026-04-25</td>
<td>Brazil</td>
</tr>
<tr class="even">
<td data-quarto-table-cell-role="th">13</td>
<td>CUST_00000000786</td>
<td>1994-06-24</td>
<td>6613</td>
<td>F</td>
<td>2024-02-25</td>
<td>IT</td>
</tr>
<tr class="odd">
<td data-quarto-table-cell-role="th">14</td>
<td>CUST_00000000071</td>
<td>1960-11-19</td>
<td>4613</td>
<td>F</td>
<td>2022-03-12</td>
<td>FR</td>
</tr>
<tr class="even">
<td data-quarto-table-cell-role="th">15</td>
<td>CUST_00000001007</td>
<td>1989-07-09</td>
<td>4551</td>
<td>F</td>
<td>2024-10-03</td>
<td>BR</td>
</tr>
<tr class="odd">
<td data-quarto-table-cell-role="th">16</td>
<td>CUST_00000000189</td>
<td>1969-01-29</td>
<td>6779</td>
<td>M</td>
<td>2022-07-08</td>
<td>China</td>
</tr>
<tr class="even">
<td data-quarto-table-cell-role="th">17</td>
<td>CUST_00000001670</td>
<td>1987-08-04</td>
<td>7295</td>
<td>M</td>
<td>2026-07-28</td>
<td>Munch</td>
</tr>
<tr class="odd">
<td data-quarto-table-cell-role="th">18</td>
<td>CUST_00000001270</td>
<td>1969-02-06</td>
<td>4717</td>
<td>F</td>
<td>2025-06-23</td>
<td>USA</td>
</tr>
<tr class="even">
<td data-quarto-table-cell-role="th">19</td>
<td>CUST_00000001305</td>
<td>1989-01-13</td>
<td>7954</td>
<td>F</td>
<td>2025-07-28</td>
<td>ES</td>
</tr>
</tbody>
</table>

</div>
</div>

{{< pagebreak >}}

:::{.direction}


## Tasks

In this part, move from **detecting problems** to **making the dataset analysis-ready**. The goal is not only to spot errors, but also to document a reproducible cleaning strategy.

### 1. Identify scales of variables

* What type of scale is each variable?
  Run the following and assign `nominal`, `ordinal`, `interval`, or `ratio`.

```{r}
#| output: false

customer_df |> glimpse()
```


* What type of scale should each variable be?
  Assign `nominal`, `ordinal`, `interval`, or `ratio`.

  - customer_id: ________________
  - date_of_birth: ________________
  - income: ________________
  - gender: ________________
  - signup_date: ________________
  - country: ________________

{{< pagebreak >}}

### 2. Analyze data quality

Examine the dataset `data/messy_customer_data.csv` and run the following:

```{r}
summary(customer_df)
```

* Summarize all data quality issues you observe (open the [data/messy_customer_data.csv](data/messy_customer_data.csv)).

<br><br><br><br><br><br><br>

{{< pagebreak >}}

### 3. Propose data preparation steps

* For each data quality issue, suggest how you would address it.

<br><br><br><br><br><br><br>

---

### 4. *(Optional challenge)* Implement in R (`tidyverse`)

* Implement the data preparation steps in R.
* Run the code after each step and inspect the results.

```{r}
# Add your solution below.
```

Note: Check how the `customer_df` changes after each preparation step!

:::


:::{.sol}

## Solution

### 1. Identify scales of variables

* What type of scale is each variable?

  All are `nominal` (strings).


* What type of scale should each variable be?
  Assign `nominal`, `ordinal`, `interval`, or `ratio`.

  - customer_id: **nominal**
  - date_of_birth: **interval**
  - income: **ratio**
  - gender: **nominal**
  - signup_date: **interval**
  - country: **nominal**

### 2. Analyze data quality

Examine the dataset `data/messy_customer_data.csv` and run the following:

```{r}
summary(customer_df)
```

* Summarize all data quality issues you observe.

- **Missing values**
  - `date_of_birth`
  - `income`
  - `signup_date`

- **Implausible / invalid values**
  - `date_of_birth` (e.g., 1890, future dates like 2030)
  - `signup_date` (e.g., invalid dates like 2023-13-01)
  - `income` (e.g., negative values)

{{< pagebreak >}}

- **Inconsistent formats**
  - `income` (e.g., "5000", "5000 EUR", "5000 USD")
  - `signup_date` (multiple date formats)

- **Outliers**
  - `income` (e.g., very high values like 200000)

- **Inconsistent categorical values**
  - `gender` (M, F, Male, Female, m, f)
  - `country` (codes, names, typos: DE, Germany, Ger, U.S., Brasil, etc.)

- **Duplicates**
  - Duplicate records identifiable via repeated `customer_id` (high frequency)

---

### 3. Propose data preparation steps

* For each data quality issue, suggest how you would address it.

* **Missing values:** impute (e.g., median for income) or remove rows
* **Implausible / invalid values:** filter out or correct (e.g., negative income, invalid dates)
* **Inconsistent formats:** standardize (convert income to numeric, parse dates to datetime)
* **Outliers:** inspect and cap or remove extreme values
* **Inconsistent categorical values:** unify categories (e.g., gender labels, country formats)
* **Duplicates:** remove duplicates based on `customer_id`

---

### 4. *(Optional challenge)* Implement in R (`tidyverse`)

```{r}
customer_df <- read_csv(
  "data/messy_customer_data.csv",
  na = c("", "NA", "NaN"),
  show_col_types = FALSE
)

# --- Income: remove currency labels and parse a numeric value ---
customer_df <- customer_df |>
  mutate(income = parse_number(income))

# --- Missing income: replace with the median ---
income_median <- median(customer_df$income, na.rm = TRUE)
customer_df <- customer_df |>
  mutate(income = replace_na(income, income_median))

# --- Dates: parse valid ISO and day/month/year values ---
parse_customer_date <- function(x) {
  parsed <- as.Date(rep(NA_character_, length(x)))
  accepted_formats <- c("%Y-%m-%d", "%Y/%m/%d", "%d/%m/%Y", "%d/%m/%y")

  for (date_format in accepted_formats) {
    still_missing <- is.na(parsed)
    parsed[still_missing] <- as.Date(x[still_missing], format = date_format)
  }

  parsed
}

customer_df <- customer_df |>
  mutate(
    date_of_birth = parse_customer_date(date_of_birth),
    signup_date = parse_customer_date(signup_date)
  )

# --- Invalid or implausible values ---
today <- Sys.Date()
customer_df <- customer_df |>
  filter(
    !is.na(date_of_birth),
    date_of_birth > as.Date("1900-01-01"),
    date_of_birth < today,
    !is.na(signup_date),
    signup_date <= today,
    income >= 0
  )

# --- Income outliers: cap values at the 99th percentile ---
upper_cap <- quantile(customer_df$income, 0.99, na.rm = TRUE)
customer_df <- customer_df |>
  mutate(income = pmin(income, upper_cap))

# --- Standardize categorical values ---
customer_df <- customer_df |>
  mutate(
    gender = case_when(
      str_to_lower(gender) %in% c("m", "male") ~ "Male",
      str_to_lower(gender) %in% c("f", "female") ~ "Female",
      TRUE ~ gender
    ),
    country = recode(
      str_trim(country),
      "Germany" = "DE", "Ger" = "DE", "DEU" = "DE", "Munch" = "DE",
      "USA" = "US", "U.S." = "US",
      "France" = "FR", "Spain" = "ES", "Italy" = "IT",
      "Netherlands" = "NL", "Switzerland" = "CH", "Austria" = "AT",
      "China" = "CN", "Japan" = "JP", "India" = "IN",
      "Brazil" = "BR", "Brasil" = "BR", "Canada" = "CA"
    )
  )

# --- Keep the first record for each customer ID ---
customer_df <- customer_df |>
  distinct(customer_id, .keep_all = TRUE)

customer_df |> head()
```

:::


:::{.direction}

---

{{< pagebreak >}}

## Wrap-up <a id="wrap-up"></a>

🎉🎈 You have completed the exercise - good work! 🎈🎉

In this exercise, we have learned to

* Structure data in a tidy format
* Prepare and explore datasets


::: {.callout-note title="Stop the Codespace"}
Remember to <b>stop</b> your Codespace [here](https://github.com/codespaces).
:::

::: {.callout-tip title="Session 2 survey"}
{{< survey session_02 exercise >}}
:::

:::
