---
session: "Session 8"
title: "Machine Learning for Non-Tabular Data"
session_id: session-08
format:
  html: default
---

::: {.direction}

In this exercise, we move from a general-purpose multilayer perceptron (MLP) to architectures that preserve the structure of images and sequences.

| Part | Topic | Time (min) |
|---|---|---:|
| 1 | Train a small MLP | 20 |
| 2 | Architecture and training choices | 20 |
| 3 | Images: MLP versus CNN | 20 |
| 4 | Sequences: MLP versus RNN | 25 |
| Wrap-up | Compare representations and models | 5 |
| **Overall** | | **90** |

## Learning objectives

You are **not expected to memorize the torch API**. Use the linked official documentation to check constructors, arguments, tensor shapes, and return values; the assessed challenge is choosing and reasoning about the architecture, not recalling function signatures.

After this exercise, you should be able to:

- connect the forward pass, loss, backpropagation, and optimizer update;
- investigate how depth, width, and learning rate affect training;
- explain why CNNs preserve spatial structure and RNNs preserve sequential structure;
- select an output layer and loss for classification or regression;
- compare specialized architectures with an MLP baseline using held-out data.

:::

:::: {.direction}

::: {.callout-note title="Data and package setup"}

This exercise uses two datasets with different underlying structures: images and temporal sequences. They let us compare a generic MLP with architectures that preserve those structures, because observations are not always naturally represented as unordered feature vectors.

| Dataset | Underlying structure | Prediction task | Comparison |
|---|---|---|---|
| Fashion-MNIST | $28 \times 28$ image | Product category | Flattened MLP versus CNN |
| Store Item Demand | Ordered daily sales | Next-day sales | Lag-vector MLP versus RNN |

**Fashion-MNIST** contains 60,000 training and 10,000 test images. Each is a $28 \times 28$ grayscale image with one of 10 product-category labels. An MLP can flatten the pixels into 784 features; the values remain, but their two-dimensional neighborhoods are no longer explicit. A CNN instead uses local filters and shared weights on the image grid, providing a clean MLP-versus-CNN comparison.

**Store Item Demand Forecasting** contains five years of daily sales for 10 stores and 50 items, with columns for date, store, item, and sales. Although stored as a CSV table, it becomes a sequential task here: 28 consecutive days from one store-item series predict the next day. An MLP treats the lags as separate features; an RNN processes them in order while maintaining a recurrent hidden state.

The same underlying observations can often be represented as a flat vector, but specialized neural-network architectures encode useful assumptions about spatial or sequential structure.

### Download the data

The Fashion-MNIST dataset comes from the official [Zalando Research repository](https://github.com/zalandoresearch/fashion-mnist), and Store Item Demand [`train.csv`](https://raw.githubusercontent.com/DharitShah13/Kaggle-Store-Item-Demand-Forecasting-Challenge/master/train.csv) from [GitHub](https://github.com/DharitShah13/Kaggle-Store-Item-Demand-Forecasting-Challenge).

From the folder containing this notebook, create exactly this structure:

```text
data/
├── fashion-mnist/
│   ├── train-images-idx3-ubyte.gz
│   ├── train-labels-idx1-ubyte.gz
│   ├── t10k-images-idx3-ubyte.gz
│   └── t10k-labels-idx1-ubyte.gz
└── store-item-demand/
    └── train.csv
```

Copy and paste these shell commands:

```bash
mkdir -p data/fashion-mnist data/store-item-demand

curl -L https://raw.githubusercontent.com/zalandoresearch/fashion-mnist/master/data/fashion/train-images-idx3-ubyte.gz \
  -o data/fashion-mnist/train-images-idx3-ubyte.gz

curl -L https://raw.githubusercontent.com/zalandoresearch/fashion-mnist/master/data/fashion/train-labels-idx1-ubyte.gz \
  -o data/fashion-mnist/train-labels-idx1-ubyte.gz

curl -L https://raw.githubusercontent.com/zalandoresearch/fashion-mnist/master/data/fashion/t10k-images-idx3-ubyte.gz \
  -o data/fashion-mnist/t10k-images-idx3-ubyte.gz

curl -L https://raw.githubusercontent.com/zalandoresearch/fashion-mnist/master/data/fashion/t10k-labels-idx1-ubyte.gz \
  -o data/fashion-mnist/t10k-labels-idx1-ubyte.gz

curl -L https://raw.githubusercontent.com/DharitShah13/Kaggle-Store-Item-Demand-Forecasting-Challenge/master/train.csv \
  -o data/store-item-demand/train.csv
```

### Packages

Install the required R packages and the torch runtime once:

```r
install.packages(c(
  "torch",
  "readr",
  "dplyr",
  "tidyr",
  "ggplot2",
  "purrr",
  "tibble"
))

torch::install_torch()
```

:::

::::

::: {.sol}
The data should have been downloaded from GitHub as described in the exercise instructions.
:::

```{r}
#| label: setup
#| solution-setup: true
library(torch)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(tibble)

set.seed(42)
torch_manual_seed(42)
device <- torch_device(if (cuda_is_available()) "cuda" else "cpu")

fashion_train_images_path <- "data/fashion-mnist/train-images-idx3-ubyte.gz"
fashion_train_labels_path <- "data/fashion-mnist/train-labels-idx1-ubyte.gz"
fashion_test_images_path <- "data/fashion-mnist/t10k-images-idx3-ubyte.gz"
fashion_test_labels_path <- "data/fashion-mnist/t10k-labels-idx1-ubyte.gz"
demand_path <- "data/store-item-demand/train.csv"

required_files <- c(
  fashion_train_images_path,
  fashion_train_labels_path,
  fashion_test_images_path,
  fashion_test_labels_path,
  demand_path
)
if (!all(file.exists(required_files))) {
  missing_files <- required_files[!file.exists(required_files)]
  stop(
    "Missing required data file(s):\n- ",
    paste(missing_files, collapse = "\n- "),
    "\nReturn to 'Data and package setup' and run the provided shell commands."
  )
}
```

# Part 1 — Train a small MLP {#part-1}

Fashion-MNIST contains grayscale images of Zalando products. A classifier could support catalogue tagging, but here it first provides a controlled setting in which to observe neural-network training.

## Task 1.1 — Prepare one reproducible prediction task

The boilerplate below uses modest, fixed subsets for laptop-friendly training. Pixel intensities are scaled to $[0,1]$. Both the MLP and CNN later receive **exactly these images and labels**.

```{r}
#| label: fashion-data
read_idx_integer <- function(connection) {
  readBin(connection, what = integer(), n = 1, size = 4, endian = "big")
}

read_idx_images <- function(path) {
  connection <- gzfile(path, open = "rb")
  on.exit(close(connection))

  magic_number <- read_idx_integer(connection)
  n_images <- read_idx_integer(connection)
  n_rows <- read_idx_integer(connection)
  n_columns <- read_idx_integer(connection)
  if (magic_number != 2051L || n_rows != 28L || n_columns != 28L) {
    stop("Unexpected Fashion-MNIST image-file header: ", path)
  }

  pixels <- readBin(
    connection,
    what = integer(),
    n = n_images * n_rows * n_columns,
    size = 1,
    signed = FALSE
  )
  if (length(pixels) != n_images * n_rows * n_columns) {
    stop("Incomplete Fashion-MNIST image file: ", path)
  }
  matrix(pixels, nrow = n_images, byrow = TRUE) / 255
}

read_idx_labels <- function(path) {
  connection <- gzfile(path, open = "rb")
  on.exit(close(connection))

  magic_number <- read_idx_integer(connection)
  n_labels <- read_idx_integer(connection)
  if (magic_number != 2049L) {
    stop("Unexpected Fashion-MNIST label-file header: ", path)
  }

  labels <- readBin(connection, what = integer(), n = n_labels, size = 1, signed = FALSE)
  if (length(labels) != n_labels) {
    stop("Incomplete Fashion-MNIST label file: ", path)
  }
  labels + 1L # torch's classification targets are indexed from 1 in R
}

fashion_all <- list(
  x = read_idx_images(fashion_train_images_path),
  y = read_idx_labels(fashion_train_labels_path)
)
fashion_test_all <- list(
  x = read_idx_images(fashion_test_images_path),
  y = read_idx_labels(fashion_test_labels_path)
)

stopifnot(
  identical(dim(fashion_all$x), c(60000L, 784L)),
  length(fashion_all$y) == 60000L,
  identical(dim(fashion_test_all$x), c(10000L, 784L)),
  length(fashion_test_all$y) == 10000L
)

set.seed(42)
development_ids <- sample(seq_len(nrow(fashion_all$x)), 9000)
train_ids <- development_ids[1:8000]
validation_ids <- development_ids[8001:9000]
test_ids <- sample(seq_len(nrow(fashion_test_all$x)), 1000)

make_classification_loader <- function(x, y, batch_size = 128, shuffle = FALSE) {
  dataloader(
    tensor_dataset(
      torch_tensor(x, dtype = torch_float()),
      torch_tensor(y, dtype = torch_long())
    ),
    batch_size = batch_size,
    shuffle = shuffle
  )
}

fashion <- list(
  train = list(x = fashion_all$x[train_ids, ], y = fashion_all$y[train_ids]),
  validation = list(x = fashion_all$x[validation_ids, ], y = fashion_all$y[validation_ids]),
  test = list(x = fashion_test_all$x[test_ids, ], y = fashion_test_all$y[test_ids])
)

fashion_loaders <- list(
  train = make_classification_loader(fashion$train$x, fashion$train$y, shuffle = TRUE),
  validation = make_classification_loader(fashion$validation$x, fashion$validation$y),
  test = make_classification_loader(fashion$test$x, fashion$test$y)
)

dim(fashion$train$x)
table(fashion$train$y)
```

## Task 1.2 — Complete and train the MLP

::: {.callout-tip title="Useful torch documentation and syntax"}
Read the official references for [`nn_module()`](https://torch.mlverse.org/docs/reference/nn_module.html), [`nn_linear()`](https://torch.mlverse.org/docs/reference/nn_linear.html), and [`nnf_relu()`](https://torch.mlverse.org/docs/reference/nnf_relu.html). This unrelated miniature module shows where layers are created and how data flows through them:

```r
measurement_net <- nn_module(
  initialize = function() {
    self$projection <- nn_linear(4, 2)
  },
  forward = function(x) {
    nnf_relu(self$projection(x))
  }
)
```

`initialize` registers trainable layers; `forward` describes the computation. Adapt the pattern to the dimensions requested below rather than copying those example dimensions.
:::

::: {.direction}
Complete a `784 -> 64 -> 10` MLP. The hidden layer needs ReLU. The final layer should return ten **raw logits**: `nnf_cross_entropy()` combines the appropriate multiclass transformation and loss, so do not append softmax.

```{r}
#| label: task-1-2-mlp-student-response
# YOUR CODE HERE: define fashion_mlp with two linear layers and ReLU.
```
:::

::: {.sol}
```{r}
#| label: task-1-2-mlp-solution
fashion_mlp <- nn_module(
  initialize = function() {
    self$hidden <- nn_linear(28 * 28, 64)
    self$output <- nn_linear(64, 10)
  },
  forward = function(x) {
    x |>
      self$hidden() |>
      nnf_relu() |>
      self$output()
  }
)
```
:::

::: {.direction}
Create a small optimizer factory that returns Adam for a model and learning rate. Adam is the gradient-based update rule used throughout the image experiments.

::: {.callout-tip title="Adam syntax"}
The [`optim_adam()` documentation](https://torch.mlverse.org/docs/reference/optim_adam.html) lists its options. The minimal generic form is `optim_adam(model$parameters, lr = learning_rate)`; the factory should return that optimizer for the model it receives.
:::

```{r}
#| label: task-1-2-optimizer-student-response
# YOUR CODE HERE: make_adam <- function(model, lr) ...
```
:::

::: {.sol}
```{r}
#| label: task-1-2-optimizer-solution
make_adam <- function(model, lr) optim_adam(model$parameters, lr = lr)
```
:::

The reusable helpers below handle batching, backpropagation, metric collection, and plotting. They work for any image classifier that returns ten logits; `transform` determines whether a batch is flattened or kept as an image.

```{r}
#| label: classification-helpers
evaluate_classifier <- function(model, loader, transform = identity) {
  model$eval()
  total_loss <- 0
  total_correct <- 0
  total_n <- 0
  with_no_grad({
    coro::loop(for (batch in loader) {
      x <- transform(batch[[1]]$to(device = device))
      y <- batch[[2]]$to(device = device)
      logits <- model(x)
      n <- y$size(1)
      total_loss <- total_loss + nnf_cross_entropy(logits, y)$item() * n
      total_correct <- total_correct + (logits$argmax(dim = 2) == y)$sum()$item()
      total_n <- total_n + n
    })
  })
  c(loss = total_loss / total_n, accuracy = total_correct / total_n)
}

train_classifier <- function(model, loaders, epochs = 8, lr = 1e-3,
                             transform = identity, verbose = TRUE) {
  model <- model$to(device = device)
  optimizer <- make_adam(model, lr)
  history <- vector("list", epochs)

  for (epoch in seq_len(epochs)) {
    model$train()
    coro::loop(for (batch in loaders$train) {
      x <- transform(batch[[1]]$to(device = device))
      y <- batch[[2]]$to(device = device)
      optimizer$zero_grad()
      loss <- nnf_cross_entropy(model(x), y)
      loss$backward()
      optimizer$step()
    })
    train_metrics <- evaluate_classifier(model, loaders$train, transform)
    validation_metrics <- evaluate_classifier(model, loaders$validation, transform)
    history[[epoch]] <- tibble(
      epoch = epoch,
      split = c("training", "validation"),
      loss = c(train_metrics[["loss"]], validation_metrics[["loss"]]),
      accuracy = c(train_metrics[["accuracy"]], validation_metrics[["accuracy"]])
    )
    if (verbose) message("epoch ", epoch, ": validation accuracy = ",
                         round(validation_metrics[["accuracy"]], 3))
  }
  list(model = model, history = bind_rows(history))
}

plot_learning_curves <- function(history) {
  history |>
    pivot_longer(c(loss, accuracy), names_to = "metric", values_to = "value") |>
    ggplot(aes(epoch, value, colour = split)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.2) +
    facet_wrap(~metric, scales = "free_y") +
    labs(colour = NULL) +
    theme_minimal()
}
```

::: {.direction}
Instantiate the model and use your Adam factory by calling `train_classifier()` with `lr = 1e-3`. Train for eight epochs and inspect both curves.

```{r}
#| label: task-1-2-train-student-response
# YOUR CODE HERE
```
:::

::: {.sol}
```{r}
#| label: task-1-2-train-solution
torch_manual_seed(42)
mlp_fit <- train_classifier(fashion_mlp(), fashion_loaders, epochs = 8, lr = 1e-3)
plot_learning_curves(mlp_fit$history)
```
:::

::: {.direction}
In this run, identify the code line corresponding to each stage:

1. **forward pass**;
2. computation of the **loss**;
3. **backpropagation**;
4. the Adam **parameter update**.

Why are ten output logits appropriate, and why is cross-entropy preferable to a regression loss for the ten product classes? What do the gap and direction of the training/validation curves suggest?

**Your answer:** *Write your answer here.*
:::

::: {.sol}
The forward pass is `model(x)`, the loss is `nnf_cross_entropy(...)`, `loss$backward()` computes gradients by backpropagation, and `optimizer$step()` uses those gradients to update weights. Ten logits represent scores for ten mutually exclusive classes. Cross-entropy evaluates the probability assigned to the correct class, unlike MSE, which would incorrectly treat class numbers as continuous quantities. Falling losses indicate learning; a widening train-validation gap suggests overfitting, while two high, flat losses suggest insufficient learning or capacity.
:::

# Part 2 — Experiment with architecture and training {#part-2}

## Task 2.1 — Change depth and width

The lecture distinguishes **width** (units per layer) from **depth** (number of layers). This factory removes repeated model code.

```{r}
#| label: flexible-mlp
make_mlp <- function(hidden_sizes) {
  nn_module(
    initialize = function() {
      sizes <- c(784, hidden_sizes, 10)
      layers <- list()
      for (i in seq_len(length(sizes) - 1)) {
        layers <- append(layers, list(nn_linear(sizes[i], sizes[i + 1])))
        if (i < length(sizes) - 1) layers <- append(layers, list(nn_relu()))
      }
      self$network <- do.call(nn_sequential, layers)
    },
    forward = function(x) self$network(x)
  )
}
```

::: {.direction}
Specify three configurations: shallow/narrow `784 -> 32 -> 10`, wider `784 -> 128 -> 10`, and deeper `784 -> 128 -> 64 -> 10`. The supplied loop trains each under identical conditions.

```{r}
#| label: task-2-1-configurations-student-response
# YOUR CODE HERE: architecture_configs <- list(...)
```
:::

::: {.sol}
```{r}
#| label: task-2-1-configurations-solution
architecture_configs <- list(
  narrow = c(32),
  wider = c(128),
  deeper = c(128, 64)
)
```
:::

```{r}
#| label: architecture-experiment
architecture_results <- imap(architecture_configs, function(hidden, name) {
  torch_manual_seed(42)
  fit <- train_classifier(make_mlp(hidden)(), fashion_loaders,
                          epochs = 5, lr = 1e-3, verbose = FALSE)
  fit$history |> mutate(configuration = name)
}) |> bind_rows()

architecture_results |>
  ggplot(aes(epoch, accuracy, colour = configuration, linetype = split)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  labs(linetype = "data split") +
  theme_minimal()
```

## Task 2.2 — Change the learning rate

::: {.direction}
Using the wider architecture, compare learning rates `1e-2`, `1e-3`, and `1e-4`. Complete only the vector; the loop is supplied.

```{r}
#| label: task-2-2-learning-rates-student-response
# YOUR CODE HERE: learning_rates <- ...
```
:::

::: {.sol}
```{r}
#| label: task-2-2-learning-rates-solution
learning_rates <- c(`1e-2` = 1e-2, `1e-3` = 1e-3, `1e-4` = 1e-4)
```
:::

```{r}
#| label: learning-rate-experiment
learning_rate_results <- imap_dfr(learning_rates, function(lr, name) {
  torch_manual_seed(42)
  fit <- train_classifier(make_mlp(128)(), fashion_loaders,
                          epochs = 5, lr = lr, verbose = FALSE)
  fit$history |> mutate(learning_rate = name)
})

learning_rate_results |>
  pivot_longer(c(loss, accuracy), names_to = "metric", values_to = "value") |>
  ggplot(aes(epoch, value, colour = learning_rate, linetype = split)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~metric, scales = "free_y") +
  labs(linetype = "data split") +
  theme_minimal()
```

::: {.direction}
Use the plots—not an assumption—to identify any insufficient capacity, useful added capacity, slow learning, unstable learning, or overfitting. Why is a larger/deeper network not automatically better? Keep one trained MLP (`mlp_fit`) as the baseline for Part 3.

**Your answer:** *Write your answer here.*
:::

::: {.sol}
A model shows insufficient capacity when both training and validation performance plateau at poor values. Added capacity is useful only if validation performance improves. A very small learning rate may improve too slowly within the epoch budget; a large rate may oscillate or worsen because updates overshoot useful parameter regions. Greater depth or width adds parameters and computation, can make optimization harder, and can overfit. The empirical validation curve therefore decides whether the extra capacity helped.
:::

# Part 3 — Images: MLP versus CNN {#part-3}

Flattening made each image a 784-value vector. A CNN instead receives a one-channel $28 \times 28$ grid, matching the lecture's idea that model architecture should reflect data structure.

## Task 3.1 — Build a small CNN

::: {.direction}
Complete the network with a 2D convolution (8 output channels and a `3 x 3` kernel), ReLU, `2 x 2` max pooling, flattening, and a ten-logit output layer. With padding of one, pooling changes `28 x 28` to `14 x 14`.

::: {.callout-tip title="CNN syntax and tensor dimensions"}
Use the official references for [`nn_conv2d()`](https://torch.mlverse.org/docs/reference/nn_conv2d.html) and [`nn_max_pool2d()`](https://torch.mlverse.org/docs/reference/nn_max_pool2d.html). Generic constructor forms are `nn_conv2d(in_channels, out_channels, kernel_size, padding = ...)` and `nn_max_pool2d(kernel_size = ...)`; ReLU uses the function already introduced in Task 1.2.

Track the batch separately through this task:

- input: `(batch, 1, 28, 28)`;
- after the padded convolution: `(batch, 8, 28, 28)`;
- after `2 x 2` pooling: `(batch, 8, 14, 14)`;
- after flattening all non-batch dimensions: `(batch, 8 * 14 * 14)`.

Therefore, derive the final linear layer's input width from the last expression and give it 10 outputs. This shape reasoning is provided so the task remains about spatial architecture, not debugging an opaque matrix-size error.
:::

```{r}
#| label: task-3-1-cnn-student-response
# YOUR CODE HERE: define fashion_cnn.
```
:::

::: {.sol}
```{r}
#| label: task-3-1-cnn-solution
fashion_cnn <- nn_module(
  initialize = function() {
    self$conv <- nn_conv2d(1, 8, kernel_size = 3, padding = 1)
    self$pool <- nn_max_pool2d(kernel_size = 2)
    self$output <- nn_linear(8 * 14 * 14, 10)
  },
  forward = function(x) {
    x <- self$pool(nnf_relu(self$conv(x)))
    x <- torch_flatten(x, start_dim = 2)
    self$output(x)
  }
)
```
:::

```{r}
#| label: cnn-training-and-comparison
as_images <- function(x) x$view(c(x$size(1), 1, 28, 28))

torch_manual_seed(42)
cnn_fit <- train_classifier(fashion_cnn(), fashion_loaders,
                            epochs = 8, lr = 1e-3, transform = as_images)
plot_learning_curves(cnn_fit$history)

count_parameters <- function(model) {
  sum(vapply(model$parameters, function(p) p$numel(), numeric(1)))
}

image_comparison <- tibble(
  model = c("MLP", "CNN"),
  parameters = c(count_parameters(mlp_fit$model), count_parameters(cnn_fit$model)),
  validation_accuracy = c(
    evaluate_classifier(mlp_fit$model, fashion_loaders$validation)[["accuracy"]],
    evaluate_classifier(cnn_fit$model, fashion_loaders$validation, as_images)[["accuracy"]]
  ),
  test_accuracy = c(
    evaluate_classifier(mlp_fit$model, fashion_loaders$test)[["accuracy"]],
    evaluate_classifier(cnn_fit$model, fashion_loaders$test, as_images)[["accuracy"]]
  )
)
image_comparison
```

::: {.direction}
What neighborhood information is no longer explicit after flattening? What spatial assumption does convolution introduce, and how does weight sharing reuse a filter at different locations? Did the CNN actually outperform the MLP in your run? Discuss both accuracy and parameter count; do not assume that the specialized model must win.

**Your answer:** *Write your answer here.*
:::

::: {.sol}
Flattening retains pixel values but no explicit two-dimensional neighborhood: adjacent vector entries are not always adjacent pixels. A convolution assumes that local patterns matter and applies the same learned filter across locations, greatly reducing separate parameters. Whether that inductive bias produces better held-out accuracy is an empirical question answered by `image_comparison`; randomness, subset size, architecture, and training budget can alter the result.
:::

# Part 4 — Sequential data: MLP versus RNN {#part-4}

The Store Item Demand competition contains five years of daily sales for 50 items in 10 stores. A retailer might use a next-day forecast for replenishment. We select one complete store-item series and use the previous 28 days to predict the next day.

## Task 4.1 — Construct chronological rolling windows

The scaler is fitted on the earliest 70% of observations only. Windows are assigned by their **target date**, giving chronological 70%/15%/15% train/validation/test sets without leaking future targets into training.

```{r}
#| label: demand-data
demand_raw <- read_csv(demand_path, show_col_types = FALSE) |>
  mutate(date = as.Date(date)) |>
  arrange(store, item, date)

series_sizes <- demand_raw |> count(store, item, name = "n") |> arrange(desc(n))
chosen <- slice_head(series_sizes, n = 1)

sales_series <- demand_raw |>
  filter(store == chosen$store, item == chosen$item) |>
  arrange(date) |>
  select(date, sales)

n_days <- nrow(sales_series)
train_end <- floor(0.70 * n_days)
validation_end <- floor(0.85 * n_days)
sales_mean <- mean(sales_series$sales[seq_len(train_end)])
sales_sd <- sd(sales_series$sales[seq_len(train_end)])
scaled_sales <- (sales_series$sales - sales_mean) / sales_sd

window_length <- 28L
target_rows <- (window_length + 1L):n_days
windows <- t(vapply(target_rows, function(i) scaled_sales[(i - window_length):(i - 1)],
                    numeric(window_length)))
targets <- scaled_sales[target_rows]
target_dates <- sales_series$date[target_rows]

window_split <- case_when(
  target_rows <= train_end ~ "train",
  target_rows <= validation_end ~ "validation",
  TRUE ~ "test"
)

make_regression_loader <- function(split, shuffle = FALSE) {
  ids <- which(window_split == split)
  dataloader(
    tensor_dataset(
      torch_tensor(windows[ids, , drop = FALSE], dtype = torch_float()),
      torch_tensor(matrix(targets[ids], ncol = 1), dtype = torch_float())
    ),
    batch_size = 64,
    shuffle = shuffle
  )
}

demand_loaders <- list(
  train = make_regression_loader("train", TRUE),
  validation = make_regression_loader("validation"),
  test = make_regression_loader("test")
)

tibble(split = window_split, date = target_dates) |>
  group_by(split) |>
  summarise(first_date = min(date), last_date = max(date), n = n())
```

## Task 4.2 — Fit the lag-vector MLP baseline

The regression helpers use MSE for optimization and report RMSE in the original sales units. As before, batching and training mechanics are supplied.

```{r}
#| label: regression-helpers
evaluate_regressor <- function(model, loader, transform = identity) {
  model$eval()
  predictions <- list()
  observed <- list()
  with_no_grad({
    coro::loop(for (batch in loader) {
      predictions <- append(predictions, list(as.numeric(model(
        transform(batch[[1]]$to(device = device)))$cpu()
      )))
      observed <- append(observed, list(as.numeric(batch[[2]])))
    })
  })
  predicted_scaled <- unlist(predictions)
  observed_scaled <- unlist(observed)
  tibble(
    observed = observed_scaled * sales_sd + sales_mean,
    predicted = predicted_scaled * sales_sd + sales_mean
  )
}

train_regressor <- function(model, loaders, epochs = 20, lr = 1e-3,
                            transform = identity) {
  model <- model$to(device = device)
  optimizer <- optim_adam(model$parameters, lr = lr)
  history <- vector("list", epochs)
  for (epoch in seq_len(epochs)) {
    model$train()
    coro::loop(for (batch in loaders$train) {
      x <- transform(batch[[1]]$to(device = device))
      y <- batch[[2]]$to(device = device)
      optimizer$zero_grad()
      loss <- nnf_mse_loss(model(x), y)
      loss$backward()
      optimizer$step()
    })
    validation_predictions <- evaluate_regressor(model, loaders$validation, transform)
    history[[epoch]] <- tibble(
      epoch = epoch,
      validation_rmse = sqrt(mean((validation_predictions$observed -
                                   validation_predictions$predicted)^2))
    )
  }
  list(model = model, history = bind_rows(history))
}
```

::: {.direction}
Complete an MLP that treats the 28 lagged observations as a flat vector: `28 -> 32 -> 1`. Use ReLU in the hidden layer and one linear output because sales forecasting is regression.

Refer back to the `nn_module()`, `nn_linear()`, and `nnf_relu()` example in Task 1.2; this task reuses exactly that API pattern with new dimensions and a regression output.

```{r}
#| label: task-4-2-demand-mlp-student-response
# YOUR CODE HERE
```
:::

::: {.sol}
```{r}
#| label: task-4-2-demand-mlp-solution
demand_mlp <- nn_module(
  initialize = function() {
    self$hidden <- nn_linear(28, 32)
    self$output <- nn_linear(32, 1)
  },
  forward = function(x) self$output(nnf_relu(self$hidden(x)))
)

torch_manual_seed(42)
demand_mlp_fit <- train_regressor(demand_mlp(), demand_loaders, epochs = 20)
```
:::

## Task 4.3 — Preserve order with an RNN

::: {.direction}
Complete a small RNN with one feature per time step and 16 hidden units. In `forward()`, pass the sequence through the recurrent layer, select its final hidden representation, and map that representation to one predicted sales value.

::: {.callout-tip title="RNN syntax and shapes"}
Consult [`nn_rnn()`](https://torch.mlverse.org/docs/reference/nn_rnn.html), especially its input/output shape section. The generic constructor is `nn_rnn(input_size = feature_count, hidden_size = hidden_count, batch_first = TRUE)`.

Here each batch initially has shape `(batch, 28)`. Add a final singleton feature dimension before the model so the RNN receives `(batch, sequence_length = 28, input_size = 1)`. Setting `batch_first = TRUE` makes that ordering explicit; without it, the RNN expects sequence length before batch. The returned hidden state contains the final representation for each recurrent layer and batch; select the representation for the single layer, then pass its 16 features to the output layer. Reuse the module and linear-layer pattern from Task 1.2.
:::

```{r}
#| label: task-4-3-rnn-student-response
# YOUR CODE HERE
```
:::

::: {.sol}
```{r}
#| label: task-4-3-rnn-solution
demand_rnn <- nn_module(
  initialize = function() {
    self$rnn <- nn_rnn(input_size = 1, hidden_size = 16, batch_first = TRUE)
    self$output <- nn_linear(16, 1)
  },
  forward = function(x) {
    recurrent_result <- self$rnn(x)
    final_hidden <- recurrent_result[[2]][1, , ]
    self$output(final_hidden)
  }
)

as_sequence <- function(x) x$unsqueeze(3)

torch_manual_seed(42)
demand_rnn_fit <- train_regressor(demand_rnn(), demand_loaders,
                                  epochs = 20, transform = as_sequence)
```
:::

```{r}
#| label: sequence-comparison
validation_curves <- bind_rows(
  demand_mlp_fit$history |> mutate(model = "MLP"),
  demand_rnn_fit$history |> mutate(model = "RNN")
)

validation_curves |>
  ggplot(aes(epoch, validation_rmse, colour = model)) +
  geom_line(linewidth = 0.9) +
  theme_minimal()

mlp_test <- evaluate_regressor(demand_mlp_fit$model, demand_loaders$test) |>
  mutate(model = "MLP", date = target_dates[window_split == "test"])
rnn_test <- evaluate_regressor(demand_rnn_fit$model, demand_loaders$test, as_sequence) |>
  mutate(model = "RNN", date = target_dates[window_split == "test"])
test_predictions <- bind_rows(mlp_test, rnn_test)

test_predictions |>
  group_by(model) |>
  summarise(
    RMSE = sqrt(mean((observed - predicted)^2)),
    MAE = mean(abs(observed - predicted))
  )

test_predictions |>
  filter(date <= min(date) + 89) |>
  select(date, model, observed, predicted) |>
  pivot_longer(c(observed, predicted), names_to = "series", values_to = "sales") |>
  ggplot(aes(date, sales, colour = series)) +
  geom_line() +
  facet_wrap(~model, ncol = 1) +
  labs(colour = NULL) +
  theme_minimal()
```

::: {.direction}
Answer from the code and output:

1. Where is sequential structure represented in the RNN input, and what is less explicit in the flat MLP input?
2. How does the recurrent hidden state carry information from earlier days?
3. Why would randomly splitting overlapping time windows give an unrealistic evaluation?
4. Which architecture achieved lower held-out RMSE/MAE in your run? Does the specialized architecture actually help on this deliberately simple series?

Transformers provide a more general attention-based approach to sequences, as discussed in the lecture, but are not required here.

**Your answer:** *Write your answer here.*
:::

::: {.sol}
The RNN receives 28 ordered time steps with one feature each; recurrence updates a hidden state at every step, so the final state summarizes earlier observations. The MLP receives positions as separate features and has no shared temporal transition mechanism. A random split lets nearly identical overlapping windows—and potentially future patterns—appear on both sides of the split, exaggerating generalization. The reported test RMSE and MAE determine the winner. An RNN need not win: a short, regular univariate series may be modeled efficiently by lag features, and optimization and sample size also matter.
:::

::: {.direction}

# Wrap-up {#wrap-up}

Complete the final column using your held-out results.

| Data | Representation | Model | Structural assumption | Task/loss | Validation or test performance |
|---|---|---|---|---|---|
| Fashion images | flattened pixels | MLP | fully connected | multiclass classification / cross-entropy | |
| Fashion images | 2D pixels | CNN | local/spatial | multiclass classification / cross-entropy | |
| Sales history | flattened lag vector | MLP | fully connected | regression / MSE | |
| Sales history | ordered sequence | RNN | temporal recurrence | regression / MSE | |

**Central takeaway:** **Data structure influences architecture; prediction task determines output layer and loss; architecture and training choices are then evaluated empirically.**

<!--
::: {.callout-tip title="Session 8 survey"}
{{< survey session_08 exercise >}}
:::
-->

:::
