---
title: "Session 4: ML Baselines (Regression)"
session_id: session-04
format:
  html: default
---

:::{.direction}

In this exercise, we cover:

| Exercise part | Time (min) |
|---|---:|
| [Baseline regression](#baseline) | 18 |
| [Regression inference and model summary](#inference) | 17 |
| [Regression diagnostics](#diagnostics) | 22 |
| [Subgroup analysis](#subgroup-analysis) | 18 |
| [Deployment](#deployment) | 10 |
| [Wrap-up](#wrap-up) | 5 |
| **Overall** | **90** |

Work directly in this Quarto (`.qmd`) exercise using R in RStudio, Positron, or VS Code.

# Case: Employee performance at MetaCorp

You are a business analyst working for MetaCorp.
The company wants to understand which factors are associated with employee performance.
A central question in this exercise is whether **remote work** is associated with employee performance once we also account for other relevant variables.

MetaCorp is interested in questions such as:

- Does remote work seem to be associated with higher or lower performance?
- Does this relationship remain once we control for experience, training, salary, and management quality?
- Is the relationship between remote work and performance similar across departments?
- When should we be careful not to overinterpret regression results?

The target variable is `performance`, a quantitative employee performance score.

Potential predictors include:

- `experience`: years of work experience
- `training_hours`: annual training hours
- `salary`: annual salary in EUR
- `remote_share`: share of working time spent remotely (between 0 and 1)
- `team_size`: size of the employee's team
- `manager_quality`: manager quality on a 1–5 scale
- `department`: department (`Sales`, `IT`, `HR`)

:::

# Setup and dataset

## Load and inspect the dataset

```{r}
library(tidyverse)

df <- read_csv(
  "https://raw.githubusercontent.com/fs-ise/machine-learning-for-big-data/main/exercises/data/employee_performance_data.csv",
  show_col_types = FALSE
) |>
  mutate(
    department = factor(
      department,
      levels = c("HR", "IT", "Sales")
    )
  )

glimpse(df)
summary(df)
```

Setting the factor levels explicitly makes **HR the reference category**. R's formula interface will automatically represent the other departments with contrasts; we do not need to create indicator columns manually.
## Variable overview

| Variable | Meaning | Type |
|---|---|---|
| `performance` | employee performance score | target variable |
| `experience` | years of experience | numeric |
| `training_hours` | annual training hours | numeric |
| `salary` | annual salary in EUR | numeric |
| `remote_share` | share of remote work between 0 and 1 | numeric |
| `team_size` | size of employee team | numeric |
| `manager_quality` | manager quality rating from 1 to 5 | numeric |
| `department` | employee department | categorical |

# Part A — Baseline regression <a id="baseline"></a>

We begin with one regression model for the full company. This gives us a first answer to the question:

> How is `remote_share` associated with performance when we control for other variables at the same time?

This is useful because simple bivariate comparisons can be misleading. For example, employees with higher remote work shares may also differ in their department, training patterns, or salaries.

## Task A1: Fit a regression model

:::{.direction}
Use the full dataset to estimate a linear regression model for `performance`.

Your tasks:

1. Specify the regression formula with `performance` as the dependent variable.
2. Include `experience`, `training_hours`, `salary`, `remote_share`, `team_size`, `manager_quality`, and `department` as predictors.
3. Fit the model with `lm()` and store it as `model`.
4. Inspect its coefficients with `coef(model)`.
5. Identify the intercept and pay special attention to `remote_share`.
6. Identify how R represented the `department` factor.

```{r}
# Add your solution below.
```

The identifying fields `name`, `employee_id`, and `role` simply do not appear in the formula. Because `department` is a factor, the formula interface handles its contrasts automatically.
:::

:::{.sol}

```{r}
model <- lm(
  performance ~
    experience +
    training_hours +
    salary +
    remote_share +
    team_size +
    manager_quality +
    department,
  data = df
)

coef(model)
```

The coefficient names `departmentIT` and `departmentSales` show that R compares IT and Sales with the reference category, HR.
:::
## Task A2: Interpret the coefficients

:::{.direction}
Answer the following questions.
:::

1. Which variables have a positive association with performance?

::: {.sol}
The positive coefficients are for `experience`, `training_hours`, `salary`, `team_size`, `manager_quality`, `departmentIT`, and `departmentSales`. Higher values of the numeric variables are associated with higher predicted performance, holding the other predictors constant. The department coefficients represent differences relative to HR.
:::

2. Which variable has a negative association with performance?

::: {.sol}
`remote_share` has a negative coefficient of approximately **-0.936**. More remote work is therefore associated with slightly lower predicted performance in the pooled model, holding the other variables constant.
:::

3. How would you interpret the coefficient of `remote_share`?

::: {.sol}
Because `remote_share` ranges from 0 to 1, moving from fully on-site (`0`) to fully remote (`1`) is associated with about **0.94 points lower** predicted performance, holding the other variables constant. An increase of `0.10`—10 percentage points more remote work—is associated with about **0.094 points lower** predicted performance.
:::

4. Is the `remote_share` association large or small in practical terms?

::: {.sol}
It appears small. Even a change from fully on-site to fully remote corresponds to less than one performance point. By comparison, the `manager_quality` coefficient is about **5.82** per scale point, while the IT and Sales differences relative to HR are about **13.93** and **12.76** points.
:::

5. How should `departmentIT` and `departmentSales` be interpreted?

::: {.sol}
R created contrasts for the factor automatically. With HR as the explicit reference category:

- `departmentIT` is about **13.93**, so IT employees have about 13.93 points higher predicted performance than otherwise comparable HR employees.
- `departmentSales` is about **12.76**, so Sales employees have about 12.76 points higher predicted performance than otherwise comparable HR employees.

These coefficients compare each department with HR; they do not directly compare IT with Sales.
:::

:::{.direction}

::: {.callout-note title="Interpretation reminder"}
A coefficient expresses the expected change in the dependent variable when the predictor increases by one unit, **holding the other variables constant**.
:::

::: {.callout-important title="Important note on remote_share"}
Because `remote_share` ranges from 0 to 1, a one-unit increase means moving from **fully on-site** to **fully remote**. In practice, smaller changes such as 0.10, or 10 percentage points, are often easier to interpret.
:::

:::
# Part B — Regression inference and model summary <a id="inference"></a>

Base R's `lm()` provides estimation, prediction, and statistical inference in the same fitted model. We can examine coefficient uncertainty and overall model fit without refitting the model through another library.

## Task B1: Examine the fitted model

:::{.direction}
Use the same `model` from Part A.

1. Display its regression summary with `summary(model)`.
2. Display coefficient confidence intervals with `confint(model)`.
3. Retrieve only the confidence interval for `remote_share`.

```{r}
# Add your solution below.
```

In the summary, locate the coefficient estimates, standard errors, t-statistics, p-values, residual standard error, R², adjusted R², and overall F-test.
:::

:::{.sol}

```{r}
summary(model)
confint(model)
confint(model)["remote_share", ]
```

:::

## Task B2: Interpret inference and model fit

:::{.direction}
Use the model output to answer the following questions.
:::

1. Which predictors are statistically significant at the 5% level?

::: {.sol}
All included predictors except `remote_share` are statistically significant at the 5% level. The department contrasts are also statistically significant relative to HR.
:::

2. What do the p-value and confidence interval say about `remote_share`?

::: {.sol}
The estimate is about **-0.936**, its t-statistic is about **-0.76**, and its p-value is about **0.449**. Its 95% confidence interval is approximately **[-3.36, 1.49]**, which includes zero. Thus, the pooled data do not provide strong evidence that the association differs from zero.
:::

3. What do R² and adjusted R² tell you?

::: {.sol}
R² is approximately **0.816**, so the model explains about 81.6% of the observed variation in performance. Adjusted R² is approximately **0.814** and accounts for the number of predictors. Strong model fit does not prove that the specification is correct or that its associations are causal.
:::

4. How do statistical significance and practical relevance differ here?

::: {.sol}
Statistical significance concerns whether the data provide evidence that an association differs from zero. Practical relevance concerns whether its magnitude matters for decisions. The small `remote_share` estimate is neither statistically significant in the pooled model nor large in practical terms, but context and subgroup differences still deserve investigation.
:::

5. Why must we avoid a causal interpretation?

::: {.sol}
This observational regression controls only for measured predictors. Unobserved differences, selection into remote work, reverse causality, or measurement error could explain associations. The coefficient therefore describes a conditional association, not necessarily the effect of changing remote-work policy.
:::

6. What does the formula interface allow us to express directly, and how does R handle `department`?

::: {.sol}
The formula makes the dependent variable and predictors explicit and readable: `performance ~ experience + ... + department`. Because `department` is a factor, R automatically constructs contrasts relative to HR and reports them as `departmentIT` and `departmentSales`.
:::
# Part C — Regression diagnostics <a id="diagnostics"></a>

::: {.direction}
Regression coefficients are only part of the story. Before relying on model predictions, assess whether the model reasonably represents the data.

A key linear-regression assumption is **linearity**. Residuals capture what the fitted model does not explain. If the model is appropriate, they should look like random noise around zero. Curvature or other systematic patterns can indicate missing structure, while a funnel shape can indicate non-constant variance.

Even if `remote_share` has a positive or negative coefficient, that does **not** mean that the model is perfectly specified or that the relationship is causal.
:::

## Task C1: Residual plot

:::{.direction}
Create a data frame containing fitted values and residuals from `model`, then create a residual plot with ggplot2.

```{r}
# Add your solution below.
```
:::

:::{.sol}

```{r}
diagnostics_df <- tibble(
  fitted = fitted(model),
  residual = residuals(model)
)

ggplot(
  diagnostics_df,
  aes(x = fitted, y = residual)
) +
  geom_point() +
  geom_hline(
    yintercept = 0,
    linetype = "dashed"
  ) +
  labs(
    title = "Residual plot",
    x = "Predicted performance",
    y = "Residuals"
  )
```
:::

Questions:

1. Do the residuals look roughly random?

::: {.sol}
Yes. They appear roughly randomly scattered around zero, suggesting that a linear model is a reasonable approximation for these data.
:::

2. Do you see systematic patterns that may indicate nonlinearity or other model problems?

::: {.sol}
No clear curvature or systematic structure is visible. The plot therefore shows no obvious misspecification, although a single plot cannot establish that every regression assumption holds.
:::

3. What would a clear funnel shape suggest?

::: {.sol}
A funnel shape would suggest **heteroscedasticity**: residual variance changes with fitted values. Non-constant variance can make conventional standard errors, confidence intervals, and statistical tests unreliable.
:::
# Part D — Subgroup analysis: does one model fit all departments? <a id="subgroup-analysis"></a>

::: {.direction}
Sometimes a single regression for the whole company may hide important differences. Internal feedback at MetaCorp suggests that complaints about remote work are more frequent in some departments than in others. Coordination-intensive work may face communication challenges, while other work may benefit from concentration and flexibility.

Separate department models provide an intuitive first exploration of whether relationships look similar across departments, whether some variables matter more in certain contexts, and whether the association with `remote_share` changes with the work environment.
:::

## Task D1: Department-specific regressions

Estimate separate regressions for Sales, IT, and HR. Omit `department` from each formula because each subset contains only one department.

:::{.direction}

```{r}
departments <- unique(df$department)
results <- list()

# Add a loop below. For each department:
# 1. filter df;
# 2. fit the subgroup model with lm();
# 3. store it in results;
# 4. print summary(model_sub).
```
:::

:::{.sol}

```{r}
departments <- unique(df$department)
results <- list()

for (dept in departments) {
  df_sub <- df |>
    filter(department == dept)

  model_sub <- lm(
    performance ~
      experience +
      training_hours +
      salary +
      remote_share +
      team_size +
      manager_quality,
    data = df_sub
  )

  results[[as.character(dept)]] <- model_sub

  cat("\n=== Regression results for", as.character(dept), "===\n")
  print(summary(model_sub))
}
```
:::
## Task D2: Compare the subgroup models

1. Do the coefficients differ across departments?

::: {.sol}
Yes. Most predictors remain positively associated with performance, but magnitudes differ. For example, `experience` is about **1.36** in HR, **1.67** in IT, and **2.07** in Sales. `training_hours` is very similar across groups at about **0.90–0.91**.
:::

2. Is `remote_share` associated with performance in the same way in every department?

::: {.sol}
No. Its coefficient is approximately **-6.41** in HR (p ≈ 0.027), **4.07** in IT (p ≈ 0.043), and **-2.92** in Sales (p ≈ 0.114). Thus it is negative and statistically significant in HR, positive and statistically significant in IT, and negative but not statistically significant in Sales. These are subgroup associations, not causal effects.
:::

3. Does `manager_quality` seem equally important in all departments?

::: {.sol}
It is positive and statistically significant in every department, but its magnitude differs: approximately **4.72** in HR, **7.45** in IT, and **4.98** in Sales. The estimated association is strongest in IT.
:::

4. What does this tell us about heterogeneity in company data?

::: {.sol}
Relationships are not identical across departments, and the `remote_share` coefficient even changes direction. Departments may differ in tasks, work structures, and coordination needs, producing different performance relationships.
:::

5. Why might a single pooled model be too simplistic?

::: {.sol}
The pooled model allows department intercepts to differ but assumes that slopes are the same. The subgroup results suggest that some slopes vary, so the pooled coefficient can mask important contextual differences.
:::

::: {.callout-note title="From subgroups to interactions"}
Separate regressions are intuitive for exploration. Differences can also be represented in one model with interaction terms, for example:

```{r}
#| eval: false
performance ~
  experience +
  training_hours +
  salary +
  remote_share * department +
  team_size +
  manager_quality
```

The interaction allows the `remote_share` slope to differ by department while estimating one combined model.
:::

# Part E — Deployment <a id="deployment"></a>

Saving and loading a model is useful in practice, but it is not the main conceptual focus of this exercise. It is included as an optional extension after estimation, interpretation, diagnostics, and subgroup analysis.

## Task E1 (optional): Save a fitted model

:::{.direction}

```{r}
# Save model to data/employee_performance_model.rds.
```
:::

:::{.sol}

```{r}
saveRDS(
  model,
  "data/employee_performance_model.rds"
)
```
:::
## Task E2 (optional): Load the model and predict a new case

::: {.direction}
Load the fitted model, create a tibble for an IT employee with the original predictors, and call `predict()`.

```{r}
# Add your solution below.
```

> **Important:** The fitted `lm()` object retains its formula and factor contrasts. You do not need to construct department indicator variables manually. Production use still requires validation that incoming values, units, factor levels, and other preparation rules match the training data.
:::

:::{.sol}

```{r}
loaded_model <- readRDS(
  "data/employee_performance_model.rds"
)

new_employee <- tibble(
  experience = 8,
  training_hours = 45,
  salary = 56000,
  remote_share = 0.4,
  team_size = 7,
  manager_quality = 4,
  department = factor(
    "IT",
    levels = levels(df$department)
  )
)

predict(
  loaded_model,
  newdata = new_employee
)
```

The predicted performance is approximately **169.23**. The saved model knows how the department factor was represented during fitting, so prediction uses the original predictor rather than manually created indicator columns.
:::

## Task E3: Reflect on practical and organizational considerations

:::{.direction}
In practice, deploying a model like this involves more than saving and loading it. Think about how such a model would be used in a real company like MetaCorp.
:::

**Questions:**

1. What practical challenges might arise when using this model to support decisions about employees?

::: {.sol}
- The model may rely on incomplete or imperfect data, omitting factors such as motivation or task complexity.
- New inputs must use the same variables, units, factor levels, and preparation rules as the training data.
- The model may not generalize after company practices or policies change.
- Managers may misinterpret conditional associations as causal effects.
:::

2. What risks could emerge if the model is used without careful interpretation or oversight?

::: {.sol}
- Decisions could be based on misleading relationships or causal interpretations.
- The model could unintentionally disadvantage departments, work styles, or other groups.
- Overreliance on model output could displace human judgment and relevant context.
- Employees may lose trust if decisions appear opaque or unfair.
:::

3. What safeguards are important for fair, transparent, and responsible use, including legal and organizational requirements?

::: {.sol}
- Ensure legal compliance from the outset, particularly GDPR and rules governing employee data and monitoring.
- Involve employee representatives, such as works councils, early; employee-related model use may require formal participation and approval.
- Define whether the model supports people or makes automated decisions, and align that use with legal and organizational rules.
- Document the model, its data, limitations, and appropriate interpretation.
- Communicate openly with employees about its purpose and limits.
- Assess and address bias or unfair effects across groups.
- Maintain meaningful human oversight.
- Validate and monitor the model as conditions change.

Responsible use requires aligning technical implementation with legal requirements, organizational processes, and ethical considerations.
:::

:::{.direction}
## Wrap-up <a id="wrap-up"></a>

🎉🎈 You have completed the exercise—good work! 🎈🎉

In this exercise, we have learned to

- Fit and interpret linear regression models in R
- Use regression summaries for statistical inference
- Interpret coefficients, confidence intervals, and statistical significance
- Diagnose regression models using residual plots
- Compare relationships across subgroups
- Save, load, and apply fitted models for prediction
- Reflect on responsible organizational use of predictive models

<!--
::: {.callout-tip title="Session 4 survey"}
{{< survey session_04 exercise >}}
:::
-->

:::
