---
session: "Session 6"
title: "Machine Learning Foundations — Workflow"
session_id: session-06
format:
  html: default
---

::: {.direction}

In this exercise, we cover:

| Part        | Topic                                    | Time (min) |
|-------------|------------------------------------------|-----------:|
| 1           | Define the problem and split the data    |         12 |
| 2           | Specify preprocessing without leakage    |         10 |
| 3–4         | Specify models and workflows             |         10 |
| 5–6         | Cross-validation and baseline evaluation |         15 |
| 7           | Tune a decision tree                     |         15 |
| 8           | Compare and select models                |          8 |
| 9           | Final held-out test evaluation           |         12 |
| 10          | Generalization reflection                |          8 |
| **Overall** |                                          |     **90** |

## Learning objective

Implement a supervised machine-learning workflow in R and explain how splitting, resampling, preprocessing, tuning, and final testing support **generalization**.

We continue the LendWise credit-default case from Session 5. There, the focus was how classification probabilities and metrics work. Here, the focus is whether a model will perform on **unseen customers** and how to compare model classes without contaminating the final evaluation.

A short reminder:

- supervised learning estimates a relationship between features and a known target;
- unsupervised learning searches for structure without a predefined target;
- our supervised target is `default`.

### Package setup

Before starting this exercise, make sure the required R packages are installed.

Run this command once in the R console. You do not need to run it every time you open the exercise. You can skip this step if these packages are already installed.

```{r}
#| label: package-setup-install-packages
#| eval: false
install.packages(c(
  "tidyverse",
  "tidymodels"
))
```

:::

```{r}
#| label: package-setup-load-packages
#| solution-setup: true
library(tidyverse)
library(tidymodels)
```

We use the `tidymodels` ecosystem to implement the machine-learning workflow.
It includes `yardstick`, which provides functions for evaluating model performance (e.g., accuracy, sensitivity, ROC AUC, and confusion matrices).
In Session 5, we calculated some of these metrics manually.
Here, we use standardized functions to evaluate models consistently across cross-validation and final testing.

# Part 1 — Define the problem and split the data {#split}

We use the complete `credit_default.csv` dataset and retain `customer_category` as a predictor. The event of interest is **default**.  We therefore define `default` as the first factor level, following the convention used by our evaluation functions.

::: {.direction}

## Task 1.1 — Load and inspect the data

1. Load the csv from `https://raw.githubusercontent.com/fs-ise/machine-learning-for-big-data/main/exercises/data/credit_default.csv`.
2. Convert the numeric outcome to a factor whose levels are ordered `default`, then `no_default`.
3. Inspect the data, row count, and outcome proportions.

```{r}
#| label: task-1-1-load-and-inspect-the-data-student-response
# YOUR CODE HERE
```
:::

::: {.sol}

```{r}
#| label: task-1-1-load-and-inspect-the-data-df
df <- read_csv(
  "https://raw.githubusercontent.com/fs-ise/machine-learning-for-big-data/main/exercises/data/credit_default.csv",
  show_col_types = FALSE
) |>
  mutate(
    default = factor(
      default,
      levels = c(1, 0),
      labels = c("default", "no_default")
    )
  )
glimpse(df)
count(df, default)
prop.table(table(df$default))
levels(df$default)
```

The factor-level order confirms that `default` is treated as the positive class (ground truth) when calculating ROC AUC and sensitivity.

:::

## Task 1.2 — Create a held-out test set

::: {.direction}
Create a reproducible 80/20 split, stratified by `default`. Store the partitions as `train_data` and `test_data`, then compare row counts and class proportions.

```{r}
#| label: task-1-2-create-a-held-out-test-set-student-response
# YOUR CODE HERE
```

Why must `test_data` remain untouched until model selection is complete?

**Your answer:** *Write your answer here.*
:::

::: {.sol}

```{r}
#| label: task-1-2-create-a-held-out-test-set-set-seed
set.seed(42)

data_split <- initial_split(
  df,
  prop = 0.8,
  strata = default
)

train_data <- training(data_split)
test_data <- testing(data_split)

split_summary <- tibble(
  partition = c("Full", "Training", "Test"),
  rows = c(nrow(df), nrow(train_data), nrow(test_data)),
  default_rate = c(
    mean(df$default == "default"),
    mean(train_data$default == "default"),
    mean(test_data$default == "default")
  )
)

split_summary
```

Stratification keeps the outcome distribution consistent across partitions. The test set must remain untouched because it is reserved for one independent estimate of generalization performance. Repeatedly consulting it would make model choices adapt to this particular sample and introduce selection bias.
:::

# Part 2 — Specify preprocessing without leakage {#recipe}

A [recipe](https://recipes.tidymodels.org/reference/recipe.html) records what transformations should happen. It does not estimate normalization statistics until a workflow is fitted. During cross-validation, those statistics are learned separately from each analysis fold and then applied to its assessment fold.

::: {.direction}

## Task 2.1 — Create the recipe

Use `train_data` to specify a recipe that:

1. predicts `default` from all remaining variables;
2. creates indicator columns for nominal predictors such as `customer_category`;
3. normalizes numeric predictors.

```{r}
#| label: task-2-1-create-the-recipe-student-response
# YOUR CODE HERE
```

Normalization is useful for many models, including logistic regression when predictors have very different units. Trees generally do not require it, but using one common recipe makes the comparison concise; the tree's split ordering is unchanged by a monotonic rescaling.
:::

::: {.sol}

```{r}
#| label: task-2-1-create-the-recipe-model-recipe
model_recipe <- recipe(
  default ~ .,
  data = train_data
) |>
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors())
model_recipe
```

::: {.callout-tip}

## Order of preprocessing steps matters

`step_dummy()` creates numeric 0/1 variables. If followed by

`step_normalize(all_numeric_predictors())`, these dummy variables

are normalized too.

**To preserve 0/1 encoding:** Normalize numeric predictors **before**

creating dummy variables.

:::

:::

{{< needspace 10 >}}

# Part 3 — Specify two model classes {#models}

We deliberately compare two different model classes:

- logistic regression is relatively constrained and represents effects through a linear log-odds function;
- a decision tree can represent nonlinear relationships and interactions, but greater flexibility can increase overfitting risk.

::: {.direction}

## Task 3.1 — Specify the models

Create:

1. an untuned logistic regression using the `glm` engine;
2. a classification tree using the `rpart` engine, with `cost_complexity`, `tree_depth`, and `min_n` marked for tuning.

```{r}
#| label: task-3-1-specify-the-models-student-response
# YOUR CODE HERE
```
:::

::: {.sol}

```{r}
#| label: task-3-1-specify-the-models-logistic-spec
logistic_spec <- logistic_reg() |>
  set_engine("glm") |>
  set_mode("classification")

logistic_spec
```

```{r}
#| label: task-3-1-specify-the-models-tree-spec
tree_spec <- decision_tree(
  cost_complexity = tune(),
  tree_depth = tune(),
  min_n = tune()
) |>
  set_engine("rpart") |>
  set_mode("classification")

tree_spec
```

:::

{{< needspace 20 >}}

# Part 4 — Bundle preprocessing and models {#workflows}

::: {.direction}

A [workflow](https://workflows.tidymodels.org/) bundles a recipe and model specification. This ensures the same transformations are applied consistently in resampling, final fitting, and prediction.

## Task 4.1 — Create two workflows

Create `logistic_workflow` and `tree_workflow` by combining the shared recipe with each model specification.

```{r}
#| label: task-4-1-create-two-workflows-student-response
# YOUR CODE HERE
```
:::

::: {.sol}

We bundle two separate workflow objects and check their summaries:

```{r}
#| label: task-4-1-create-two-workflows-logistic-workflow
logistic_workflow <- workflow() |>
  add_recipe(model_recipe) |>
  add_model(logistic_spec)

logistic_workflow
```

```{r}
#| label: task-4-1-create-two-workflows-tree-workflow
tree_workflow <- workflow() |>
  add_recipe(model_recipe) |>
  add_model(tree_spec)

tree_workflow
```
:::

# Part 5 — Cross-validation {#cross-validation}

::: {.direction}

Cross-validation creates temporary analysis and assessment partitions **inside the training data**. These folds support model comparison and tuning; they do not replace the final held-out test set.

## Task 5.1 — Create stratified folds

Create five reproducible folds from `train_data`, stratified by `default`. Confirm that the object contains five splits.

```{r}
#| label: task-5-1-create-stratified-folds-student-response
# YOUR CODE HERE
```

:::

::: {.sol}

We create five folds from `train_data`, stratified by `default`:

```{r}
#| label: task-5-1-create-stratified-folds-set-seed
set.seed(42)

folds <- vfold_cv(
  train_data,
  v = 5,
  strata = default
)

folds
```
:::

Discuss:

1. What role do the assessment portions of the folds play?

::: {.direction}
**Your answer:** *Write your answer here.*
:::

::: {.sol}
The data are split into analysis and assessment folds. The analysis portion is used to train the model, while the assessment portion evaluates its performance on observations not used for fitting that fold’s model.
:::

2. How do they differ from `test_data`?

::: {.direction}
**Your answer:** *Write your answer here.*
:::

::: {.sol}
Fold assessments are reused for development across CV iterations; `test_data` is reserved for the final evaluation only.
:::

3. Why would repeated model selection using test performance bias the reported final result?

::: {.direction}
**Your answer:** *Write your answer here.*
:::

::: {.sol}
Choosing repeatedly by test performance adapts decisions to random peculiarities of that test sample. The resulting estimate becomes optimistically biased through selection.
:::

{{< needspace 10 >}}

# Part 6 — Evaluate the logistic baseline {#baseline}

We now evaluate the logistic regression model using `yardstick`.

Instead of calculating metrics manually, `metric_set()` combines ROC AUC, accuracy, and sensitivity into a reusable collection of evaluation metrics.

We apply the same metrics during cross-validation, model tuning, and final testing.

::: {.direction}

## Task 6.1 — Resample the baseline workflow

1. Create `classification_metrics` with `roc_auc`, `accuracy`, and `sens`.
2. Fit the logistic workflow across `folds`.
3. Collect the mean and standard error for each metric.

```{r}
#| label: task-6-1-resample-the-baseline-workflow-student-response
# YOUR CODE HERE
```
:::

::: {.sol}

```{r}
#| label: task-6-1-resample-the-baseline-workflow-classification-metrics
classification_metrics <- metric_set(
  roc_auc,
  accuracy,
  sens
)

set.seed(42)

logistic_resamples <- fit_resamples(
  logistic_workflow,
  resamples = folds,
  metrics = classification_metrics,
  control = control_resamples(save_pred = TRUE)
)

logistic_cv_metrics <- collect_metrics(logistic_resamples)
logistic_cv_metrics
```

::: {.callout-tip}

## Understanding `fit_resamples()`

- **`object`** (`logistic_workflow`): The workflow containing the preprocessing recipe and model.
- **`resamples`** (`folds`): The cross-validation splits. For each split, the model is trained on the analysis portion and evaluated on the assessment portion.
- **`metrics`** (`classification_metrics`): The performance measures to calculate (ROC AUC, accuracy, sensitivity).
- **`control`**: Additional settings. `save_pred = TRUE` retains assessment predictions for later inspection using `collect_predictions()`.

`collect_metrics()` then summarizes the performance metrics across folds.

These are validation estimates aggregated across folds—not training performance and not final test performance.

[Documentation: `fit_resamples()`](https://tune.tidymodels.org/reference/fit_resamples.html)
:::

:::

{{< needspace 15 >}}

# Part 7 — Tune the decision tree {#tuning}

The tree's hyperparameters control its flexibility:

- **`cost_complexity`:** Penalizes complex trees. Higher values encourage smaller trees.
- **`tree_depth`:** Maximum number of splits along a path. Higher values allow deeper trees.
- **`min_n`:** Minimum number of observations required in a node to attempt a split. Higher values restrict splitting.

We will examine the decision-tree algorithm in the next session. For now, we use a small tuning grid to keep computation time short.

::: {.direction}

## Task 7.1 — Create and evaluate a tuning grid

Use [expand_grid](https://tidyr.tidyverse.org/reference/expand_grid.html) to create all combinations of:

- `cost_complexity`: 0.0001, 0.001, 0.01;
- `tree_depth`: 2, 4, 6;
- `min_n`: 10, 25.

Evaluate the grid using the **same training folds and metrics**. Select the configuration with the best mean ROC AUC.

```{r}
#| label: task-7-1-create-and-evaluate-a-tuning-grid-student-response
# YOUR CODE HERE
```

:::

::: {.sol}

```{r}
#| label: task-7-1-create-and-evaluate-a-tuning-grid-tree-grid
tree_grid <- expand_grid(
  cost_complexity = c(0.0001, 0.001, 0.01),
  tree_depth = c(2L, 4L, 6L),
  min_n = c(10L, 25L)
)

set.seed(42)
tree_results <- tune_grid(
  tree_workflow,
  resamples = folds,
  grid = tree_grid,
  metrics = classification_metrics,
  control = control_grid(save_pred = TRUE)
)

best_tree <- select_best(tree_results, metric = "roc_auc")
best_tree
show_best(tree_results, metric = "roc_auc", n = 5)
```

:::

At any point in this task, did the tree see `test_data`?

::: {.direction}
**Your answer:** *Write your answer here.*
:::

::: {.sol}
- No. Every tuning decision uses only cross-validation partitions created from `train_data`.
:::

## Task 7.2 — Visualize tuning results

Visualize how maximum tree depth affects predictive performance.

::: {.direction}

Use `collect_metrics(tree_results)` to retrieve the results and:

- Keep only ROC AUC results.

- Fix `min_n = 25` and `cost_complexity = 0.001`.

- Plot `tree_depth` on the X-axis and mean ROC AUC on the Y-axis.

- Connect the points to illustrate changes in performance.

```{r}
#| label: task-7-2-visualize-tuning-results-student-response
# YOUR CODE HERE
```

:::

::: {.sol}

```{r}
#| label: task-7-2-visualize-tuning-results
#| out-width: 57%
#| fig-align: center

collect_metrics(tree_results) |>
  filter(
    .metric == "roc_auc",
    min_n == 25,
    cost_complexity == 0.001
  ) |>
  ggplot(aes(x = tree_depth, y = mean)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Effect of tree depth on predictive performance",
    x = "Maximum tree depth",
    y = "Mean validation ROC AUC"
  )
```

::: {.direction}
**Interpret the results:**

1. Which tree depth achieves the highest mean ROC AUC?
2. Does increasing tree depth always improve performance?
:::

::: {.sol}
**Interpretation:**

- Increasing maximum tree depth from 2 to 4 improves mean ROC AUC (approximately 0.61 to 0.64).
- Increasing it further to 6 reduces performance (approximately 0.63).
- Greater model flexibility does not necessarily improve generalization.

**Conclusion:** For the fixed values of `min_n = 25` and
`cost_complexity = 0.001`, a maximum tree depth of 4 achieves
the highest observed mean ROC AUC.

Different patterns may emerge for other hyperparameter combinations.
We therefore use the complete tuning results to select the
hyperparameter combination with the highest mean ROC AUC.
:::

:::

{{< needspace 8 >}}

# Part 8 — Compare and select models {#comparison}

::: {.direction}

We now compare the logistic regression baseline with the tuned decision tree.
Model selection uses cross-validation results, **not** the held-out test set.
We compare mean ROC AUC and consider the uncertainty of the estimates.

:::

## Task 8.1 — Compare validation ROC AUC

::: {.direction}

Work through the provided code step by step. Inspect the intermediate results
and focus on what the comparison means rather than writing another transformation
pipeline from scratch.

:::

**Step 1 — Prepare the logistic regression results**

::: {.direction}

Inspect `logistic_cv_metrics`. Then keep ROC AUC with `filter()` and
use `transmute()` to create consistently named columns.

```{r}
#| label: task-8-1-logistic-summary-assignment
logistic_cv_metrics

logistic_summary <- logistic_cv_metrics |>
  filter(.metric == "roc_auc") |>
  transmute(
    model = "Logistic regression",
    mean_roc_auc = mean,
    std_err = std_err
  )
logistic_summary
```

:::

::: {.sol}

```{r}
#| label: task-8-1-logistic-summary-solution
logistic_cv_metrics

logistic_summary <- logistic_cv_metrics |>
  filter(.metric == "roc_auc") |>
  transmute(
    model = "Logistic regression",
    mean_roc_auc = mean,
    std_err = std_err
  )
logistic_summary
```

:::

**Step 2 — Retrieve the best decision tree**

::: {.direction}

`show_best()` retrieves the configuration with the highest mean validation
ROC AUC across the entire tuning grid. The column names must match Step 1.

```{r}
#| label: task-8-1-tree-summary-assignment
tree_summary <- show_best(
  tree_results,
  metric = "roc_auc",
  n = 1
) |>
  transmute(
    model = "Decision tree",
    mean_roc_auc = mean,
    std_err = std_err
  )
tree_summary
```

:::

::: {.sol}

```{r}
#| label: task-8-1-tree-summary-solution
tree_summary <- show_best(
  tree_results,
  metric = "roc_auc",
  n = 1
) |>
  transmute(
    model = "Decision tree",
    mean_roc_auc = mean,
    std_err = std_err
  )
tree_summary
```

:::

**Step 3 — Compare and select**

::: {.direction}

`bind_rows()` combines the summaries and `arrange()` sorts the models.
The selection is stored in `selected_model` for Part 9.

```{r}
#| label: task-8-1-select-model-assignment
model_comparison <- bind_rows(
  logistic_summary,
  tree_summary
) |>
  arrange(desc(mean_roc_auc))

model_comparison

selected_model <- if_else(
  model_comparison$model[1] == "Decision tree",
  "decision_tree",
  "logistic_regression"
)

selected_model
```

**Interpret the results:**

1. Which model has the higher mean validation ROC AUC, and how large is the difference?
2. What do the standard errors tell us about the precision of the estimated mean ROC AUC values?
3. Does the observed improvement justify the additional complexity of the tree, or could logistic regression be sufficient?

Remember: the reported standard errors describe uncertainty in the individual
estimated means, **not directly in their difference**.

:::

::: {.sol}

```{r}
#| label: task-8-1-select-model-solution
model_comparison <- bind_rows(
  logistic_summary,
  tree_summary
) |>
  arrange(desc(mean_roc_auc))

model_comparison

selected_model <- if_else(
  model_comparison$model[1] == "Decision tree",
  "decision_tree",
  "logistic_regression"
)

selected_model
```

**Interpretation:**

- Identify the model with the higher observed mean ROC AUC and report the difference.
- Interpret the standard errors as uncertainty in each model's estimated mean;
  they do not establish whether the difference is statistically significant.
- Consider the performance difference alongside model complexity. The tree's
  tuning results were used to select its hyperparameters, so its reported
  mean may also be optimistic due to selection.

The predefined selection rule chooses the model with the higher mean
validation ROC AUC. This is a development decision, not proof that the
selected model is statistically superior. The held-out test set remains
untouched until Part 9.

:::

{{< needspace 20 >}}

# Part 9 — Finalize and evaluate once on the test set {#final-test}

Only now do we use the held-out test observations. We finalize the selected workflow, refit it on all training observations, and evaluate exactly once on `test_data` through `last_fit()`.

::: {.direction}

## Task 9.1 — Finalize the selected workflow

If the tree won, insert `best_tree` into `tree_workflow`. Otherwise, use `logistic_workflow` unchanged. Then call `last_fit()` with the original `data_split`.

```{r}
#| label: task-9-1-finalize-the-selected-workflow-student-response
# YOUR CODE HERE
```

:::

::: {.sol}

```{r}
#| label: task-9-1-finalize-the-selected-workflow-if
if (selected_model == "decision_tree") {
  final_workflow <- finalize_workflow(
    tree_workflow,
    best_tree
  )
} else {
  final_workflow <- logistic_workflow
}
set.seed(42)
final_fit <- last_fit(
  final_workflow,
  split = data_split,
  metrics = classification_metrics
)
final_test_metrics <- collect_metrics(final_fit)
final_test_metrics
```

These test metrics estimate performance on unseen observations. They are for reporting and reflection, not another round of tuning.

:::

## Task 9.2 — Inspect final predictions

Use yardstick to create a test confusion matrix and ROC curve.

::: {.sol}

```{r}
#| label: task-9-2-inspect-final-predictions
#| out-width: 60%
#| fig-align: center

final_predictions <- collect_predictions(final_fit)
final_confusion_matrix <- final_predictions |>
  conf_mat(
    truth = default,
    estimate = .pred_class
  )
final_confusion_matrix
autoplot(final_confusion_matrix, type = "heatmap")
```

```{r}
#| label: task-9-2-inspect-final-predictions-roc
#| out-width: 40%
#| fig-align: center

final_roc <- final_predictions |>
  roc_curve(
    truth = default,
    .pred_default,
    event_level = "first"
  )
final_roc |>
  autoplot() +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed"
  ) +
  labs(title = "Final held-out test ROC curve")
```

The `.pred_default` column corresponds to the explicitly defined event level. CV metrics supported model development; this confusion matrix and ROC curve describe the one-time held-out evaluation.

:::

{{< needspace 8 >}}

# Part 10 — Reflect on generalization {#reflection}

::: {.direction}

Answer the following questions concisely.

:::

1. Why could training accuracy be misleading?

::: {.sol}

A model is optimized on training observations and can learn their noise as well as their signal. High training accuracy may therefore reflect overfitting rather than performance on new customers.

:::

2. Why must preprocessing be inside the workflow and resampling process?

::: {.sol}

Each normalization or encoding step must be estimated without assessment-fold or test information. A workflow re-estimates preprocessing on each analysis fold and applies it to the corresponding assessment fold, preventing information leakage.

:::

3. Why should the test set be used only after model choices are finalized?

::: {.sol}

If model type, hyperparameters, or preprocessing are chosen based on repeated test results, choices adapt to that sample. This selection bias makes the reported test performance too optimistic and destroys its role as an independent estimate.

:::

4. Why can a more flexible model perform worse on unseen observations?

::: {.sol}

Flexibility can reduce training bias but increase variance by fitting sample-specific fluctuations. This bias–variance trade-off means a simpler model can generalize better even when the flexible model fits training data more closely.

:::

5. What is the role of cross-validation?

::: {.sol}

Cross-validation repeatedly estimates validation performance within the training data. It supports model and hyperparameter selection while preserving the independent test set, and it reveals variability across different training/validation partitions.

:::

6. What should happen after final test evaluation and before deployment?

::: {.sol}

Before deployment, anticipate challenges beyond predictive performance:

- **Data quality:** Will the required predictors be available, reliable, and consistent for new customers?
- **Operational costs:** What are the consequences of false positives and false negatives, and how should classification thresholds be chosen?
- **Fairness:** Does the model perform differently across customer groups, potentially leading to unequal outcomes?
- **Monitoring:** How will we detect changes in customer characteristics, default patterns, or model performance over time?
- **Governance:** Who is responsible for approving, documenting, monitoring, and updating the model?

Document and lock the selected workflow, preprocessing steps, and final evaluation. Define monitoring and retraining procedures and obtain appropriate organizational and legal review.

**Importantly:** Do not repeatedly tune the model against the same test set. Further development requires new independent data for evaluation.

:::

::: {.direction}

## Wrap-up

🎉🎈 You have completed the exercise—good work! 🎈🎉

You implemented the standard supervised workflow:

```text
split → recipe → workflows → training-data cross-validation
→ model selection → finalization → one held-out test evaluation
```

You also connected each step to generalization, overfitting, leakage, selection bias, and the bias–variance trade-off.

<!--

::: {.callout-tip title="Session 6 survey"}

{{< survey session_06 exercise >}}

:::

-->

:::
