---
title: "Exercise: R Setup, IDE, and Basic Data Workflow"
subtitle: "From a clean session to loading and saving data"
format:
  html:
    toc: true
    code-copy: true
    code-overflow: wrap
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-important}
### Work together

Work through this exercise with the person next to you. Compare what you see in your IDEs and help each other when something behaves differently.

If you get stuck and cannot see how to move forward, **indicate that you need help** so that a TA or instructor can assist you.
:::

## Learning goals

After this exercise, you should be able to:

- distinguish the **editor**, **R console**, **file browser**, and **in-memory workspace** in your IDE;
- run R code interactively and from a Quarto (`.qmd`) document;
- create, inspect, and remove objects in memory;
- install and load R packages;
- create and inspect a data frame;
- save tabular data to disk and load it again;
- restart R and reproduce your work from the code in this document.

In this notebook, you will get started with R by running calculations, creating and inspecting variables, working with data frames, importing and exporting data, installing and loading packages, and using project-based relative file paths with the tidyverse.

::: {.callout-note}
### Files versus objects

Keep these two ideas separate throughout the exercise:

- **Files** live on disk, for example `exercise_01.qmd` or `data/students.csv`.
- **Objects** live in the current R session, for example a data frame called `students`.

Removing an R object does **not** delete the corresponding file from disk.
:::

## Choose and orient yourself in an IDE

Start by working with your neighbor to orient yourselves in the IDE. You can complete this exercise in **RStudio**, **Positron**, or **Visual Studio Code**. You only need one of them.

Open the folder containing this `.qmd` file as your project/workspace. In RStudio, use an RStudio Project for that folder; in Positron or VS Code, open the folder itself.

| Task | RStudio | Positron | VS Code |
|---|---|---|---|
| Edit this `.qmd` file | Source pane | Editor | Editor |
| Run R interactively | Console pane | Console | R terminal |
| Inspect objects in memory | Environment pane | Variables pane | R Workspace viewer |
| Browse files | Files pane | Explorer | Explorer |
| View the rendered `.qmd` | Viewer pane | Viewer | Internal Quarto preview |

### Turn on automatic preview updates

Configure your IDE so that the rendered version of this document updates when you **save** the `.qmd` file. You should not need to click **Render** after every change.

- **RStudio:** In the `.qmd` editor toolbar, enable **Render on Save**. In the gear/options menu, select **Preview in Viewer Pane**.
- **Positron:** Enable **Render on Save** in the editor toolbar. If needed, open Settings, search for `quarto.render`, and make sure the preview type is **Internal** so that the result appears in the Viewer.
- **VS Code:** Open Settings, search for `quarto.render`, and enable **Render on Save**. Keep **Quarto > Render: Preview Type** set to **Internal**. If no preview is open yet, run **Quarto: Preview** once; after that, saving the file should refresh the preview.

::: {.callout-tip}
### Exercise: find the main interface elements

With your neighbor, try to locate the following parts of your IDE:

1. this `.qmd` file in the editor;
2. an interactive R console or R terminal;
3. the file browser/explorer;
4. the environment/variables/workspace viewer;
5. the Viewer or internal preview showing the rendered document.

For now, the goal is simply to **find these elements in the IDE**. The exercises below will use them one by one so that you can see what they do and how they relate to your R session and files.

If you use VS Code, make sure the **R extension** and **Quarto extension** are installed, then start an R terminal from the Command Palette with **R: Create R terminal**. Positron includes native R support and bundles Quarto.
:::

## Test the R setup

Start with the R console. Run:

```{r}
R.version.string
```

Now use R as a calculator:

```{r}
3 * 5
7 + 2
100 / 5
99 - 1
```

Try a few functions:

```{r}
sqrt(144)
log(10)
mean(c(4, 6, 8, 10))
```

::: {.callout-tip}
### Exercise: console versus document

1. Run `2^10` directly in the **console**.
2. Add a line below that computes `17 * 23`.
3. Run that line from the `.qmd` document.
4. **Save the `.qmd` file** and check whether the rendered version in the Viewer/preview updates automatically and shows your change. Do not click **Render**.

```{r}
# TODO: calculate 17 * 23
```

What is the difference between code typed only in the console and code stored in this document?
:::

## Create variables and inspect memory

R stores values in named **objects**. The assignment operator is `<-`.

```{r}
a <- 3 * 5
b <- 7 + 2
c <- a * b

c
```

Create objects of different types:

```{r}
course <- "Machine Learning for Big Data"
session_number <- 1
setup_complete <- TRUE
```

List the objects currently in memory:

```{r}
ls()
```

Inspect individual objects:

```{r}
class(course)
class(session_number)
class(setup_complete)
```

Also find these objects in your IDE's **Environment**, **Variables**, or **Workspace** view.

::: {.callout-tip}
### Exercise: create and remove an object

1. Create an object called `answer` containing the result of `6 * 7`.
2. Confirm that it appears in `ls()` and in your IDE's workspace view.
3. Remove only `answer` with `rm()`.
4. Confirm that it has disappeared.



```{r}
# TODO: create answer

# TODO: inspect the workspace

# TODO: remove answer

# TODO: inspect the workspace again
```
:::

## Install and load packages

R packages add functions and datasets to R. Installation and loading are different operations:

- `install.packages()` installs a package on your computer, usually **once**.
- `library()` loads an installed package into the **current R session**.

Install the tidyverse from the **R console**:

```{r}
#| eval: false
install.packages("tidyverse")
```

Do not put package installation into code that should run every time you render an analysis. After installation, load the package in the document:

```{r}
library(tidyverse)
```

Check the installed version:

```{r}
packageVersion("tidyverse")
```

If you use **VS Code**, its R support also expects the `languageserver` package. Install it once from the console if necessary:

```{r}
#| eval: false
install.packages("languageserver")
```

::: {.callout-tip}
### Exercise: package help

Open the help page for `read_csv()`:

```{r}
#| eval: false
?read_csv
```

Find the argument used to specify how missing values are represented in an input file.
:::

## Work with a data frame

A **data frame** represents rectangular tabular data: rows are observations and columns are variables.

Create a small data frame:

```{r}
students <- data.frame(
  name = c("Alex", "Bao", "Casey", "Dina", "Emil"),
  study_hours = c(2.5, 4.0, 3.5, 5.5, 1.5),
  score = c(68, 82, 76, 91, 61),
  passed = c(TRUE, TRUE, TRUE, TRUE, FALSE)
)

students
```

Inspect its structure:

```{r}
class(students)
str(students)
dim(students)
names(students)
head(students)
```

The tidyverse can work directly with ordinary R data frames. It also commonly uses **tibbles**, which are modern data-frame subclasses.

```{r}
students_tbl <- as_tibble(students)

class(students_tbl)
is.data.frame(students_tbl)
glimpse(students_tbl)
```

::: {.callout-tip}
### Exercise: inspect the data

Use R functions to answer the following questions:

- How many rows does `students` have?
- How many columns does it have?
- What is the mean of `score`?
- What are the values in the `name` column?

```{r}
# TODO: number of rows

# TODO: number of columns

# TODO: mean score

# TODO: names of students
```
:::

## Display a data frame in the IDE

So far, you have inspected data by printing it in the console or in this document. Your IDE provides an **additional graphical data viewer**, which can be more convenient for browsing rows and columns in a table.

Run this interactively rather than relying on it when rendering the document:

```{r}
#| eval: false
View(students)
```

Depending on your IDE, you may also be able to open the data viewer by clicking the `students` object in the **Environment**, **Variables**, or **Workspace** view.

Then try the following:

```{r}
head(students, 3)
summary(students)
glimpse(students)
```

::: {.callout-tip}
### Exercise: compare three views

Compare:

1. printing `students`;
2. `glimpse(students)`;
3. your IDE's graphical data viewer.

Which is most useful for quickly checking **values**, **column types**, and **larger tables**?
:::

## Work with files and relative paths

Check the current working directory:

```{r}
getwd()
```

For this course, prefer opening the exercise folder as a project/workspace and using **relative paths** rather than changing the working directory with `setwd()`.

Create a `data` directory if it does not already exist:

```{r}
dir.create("data", showWarnings = FALSE)
```

Save the data frame as a CSV file:

```{r}
write_csv(students, "data/students.csv")
```

Confirm that the file exists:

```{r}
file.exists("data/students.csv")
```

Find `students.csv` in the IDE's file browser. Open it and inspect the plain-text CSV representation.

::: {.callout-tip}
### Exercise: file versus memory

At this point, `students` exists both:

- as an object in the R session; and
- as the file `data/students.csv` on disk.

Remove the in-memory object:

```{r}
rm(students)
ls()
```

Now check whether the file still exists:

```{r}
file.exists("data/students.csv")
```

Explain why the result is `TRUE` even though `students` no longer appears in `ls()`.
:::

## Load data from a file

Read the CSV file back into memory with `read_csv()`:

```{r}
students <- read_csv("data/students.csv")
```

Inspect the result:

```{r}
students
glimpse(students)
class(students)
is.data.frame(students)
```

Notice that `read_csv()` returns a **tibble**, and `is.data.frame(students)` is still `TRUE`.

::: {.callout-tip}
### Exercise: save a modified file

Create a new object called `students_updated` in which every student receives 2 additional points. Then save it as `data/students_updated.csv`.

Hints: use `mutate()` and `write_csv()`.

```{r}
# TODO: create students_updated

# TODO: save it to data/students_updated.csv

# TODO: confirm that the file exists
```
:::

## Clean the workspace

Remove one object:

```{r}
rm(students_tbl)
```

To remove **all user-created objects** from the current R session, you can run:

```{r}
#| eval: false
rm(list = ls())
```

This affects objects in memory; it does not delete your `.qmd` file or CSV files.

A stronger habit is to restart R regularly and verify that your analysis can be recreated from the saved code.

::: {.callout-warning}
### Do not rely on a saved workspace

The durable record of your analysis should be your **code and data files**, not whatever happens to be present in memory. A clean R session helps reveal missing steps and hidden dependencies.
:::

## Final reproducibility check

Complete this checkpoint at the end:

1. Save this `.qmd` file.
2. Restart the R session.
3. Run `ls()` and confirm that your old objects are gone.
4. Run the document again from the top.
5. Confirm that `students` is recreated from `data/students.csv`.
6. Confirm that your output file(s) exist in `data/`.
7. Save the `.qmd` file and confirm that the HTML preview refreshes successfully.

```{r}
#| eval: false
# Run interactively immediately after restarting R:
ls()
```

::: {.callout-tip}
### Optional challenge

Without looking back at the earlier code, write the shortest sequence of commands that:

1. loads the tidyverse;
2. reads `data/students.csv`;
3. displays the first three rows;
4. lists the objects in memory;
5. removes the loaded data frame again.

```{r}
# TODO
```
:::

## Further learning and inspiration

The structure and exercises above are also informed by these introductory R materials:

- [RStudio User Guide: Get Started](https://docs.posit.co/ide/user/ide/get-started/) — IDE panes, packages, environment, projects, and a first analysis.
- [Positron: First data analysis with R in a Quarto document](https://positron.posit.co/tutorial-get-started-quarto.html) — project setup, packages, Quarto, Variables pane, and Data Explorer.
- [Visual Studio Code: R in Visual Studio Code](https://code.visualstudio.com/docs/languages/r) — R extension setup, R terminal, workspace viewer, data and plot viewers.
- [R for Data Science (2e): Workflow — scripts and projects](https://r4ds.hadley.nz/workflow-scripts) — reproducible projects, relative paths, clean sessions, and scripts as the source of truth.
- [Quarto: Computations with R](https://quarto.org/docs/get-started/computations/rstudio.html) — executable R code in `.qmd` documents and rendering workflows.
- [Quarto: VS Code](https://quarto.org/docs/tools/vscode.html) — Quarto editing, preview, and code execution in VS Code.
