---
title: "Session 3: EDA"
session_id: session-03
format:
  html: default
---

:::{.direction}

In this exercise, we cover:

| Exercise part | Time (min) |
|---|---:|
| [EDA: Clustering](#clustering) | 25 |
| [Wrap-up](#wrap-up) | 10 |
| **Overall** | **35** |

Work directly in this Quarto (`.qmd`) exercise using R in RStudio, Positron, or VS Code. Work through the tasks in R and compare your work with the solution.

### Package setup

Before starting this exercise, make sure the required R packages are installed.

Run this command once in the R console. You do not need to run it every time you open the exercise. You can skip this step if these packages are already installed.

```{r}
#| label: package-setup-install-packages
#| eval: false
install.packages("tidyverse")
```

:::

# EDA: Clustering {#clustering}

An organization wants to understand whether employees show recurring **work profiles**.
Instead of defining employee types in advance, we use clustering to identify groups with similar working patterns.

Dataset: `employee_work_profiles.csv`

```{r}
#| label: eda-clustering-load-packages
#| output: false

library(tidyverse)

employee_work_df <- read_csv(
  "https://raw.githubusercontent.com/fs-ise/machine-learning-for-big-data/main/exercises/data/employee_work_profiles.csv",
  show_col_types = FALSE
)

employee_work_df |>
  slice_head(n = 5)
```

The dataset contains measures such as weekly meeting hours, weekly focus hours, the share of work performed remotely, performance, and job satisfaction.

:::{.direction}

## Tasks

### 1. Prepare the data for clustering

We want the clusters to represent **how employees work**, rather than their outcomes.
Use the following three variables for clustering:

- `meeting_hours`
- `focus_hours`
- `remote_share`

Create a data frame `X` containing these variables:

```{r}
#| label: 1-prepare-the-data-for-clustering-x
X <- employee_work_df |>
  select(meeting_hours, focus_hours, remote_share)

# Print X at the end.
X
```

Before clustering, do a quick exploratory check so you know what kind of patterns you are looking for.

```{r}
#| label: 1-prepare-the-data-for-clustering-summary
summary(X)
```

```{r}
#| label: 1-prepare-the-data-for-clustering-x-pipeline
X |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ variable, scales = "free") +
  labs(
    x = NULL,
    y = "Count"
  )
```

Considering that k-means is sensitive to different measurement scales, we need to standardize the values.
This ensures that the three variables have approximately equal influence on the clusters. Base R's `scale()` centers each variable around zero and scales it to unit standard deviation.

Run the following code and observe how the scale of the variables changes:

```{r}
#| label: 1-prepare-the-data-for-clustering-x-scaled
X_scaled <- scale(X)
head(X_scaled)
```

Before running a clustering algorithm, view two of the standardized variables in a scatterplot:

```{r}
#| label: 1-prepare-the-data-for-clustering-x-scaled-df
X_scaled_df <- as.data.frame(X_scaled)

ggplot(
  X_scaled_df,
  aes(x = meeting_hours, y = focus_hours)
) +
  geom_point() +
  labs(
    title = "Standardized work-pattern data (no clustering)",
    x = "Meeting hours (scaled)",
    y = "Focus hours (scaled)"
  )
```

How many clusters would you expect? Briefly justify your answer based on the plot.

**Your answer:** *Write your answer here.*

### 2. Apply k-means clustering

Next, run the k-means algorithm for different values of `k`.
Use a `for` loop to test values from **2 to 10**. For each value of `k`, create a scatterplot showing the resulting clusters.

Remember that the clustering itself uses all three variables in `X_scaled`, even though the plot displays only `meeting_hours` and `focus_hours`.

```{r}
#| label: 2-apply-k-means-clustering-student-response
# YOUR CODE HERE
# Use a for loop to try k = 2, 3, ..., 10.
# For each k:
# 1. run kmeans()
# 2. add the cluster labels to X_scaled_df
# 3. create and print a scatterplot
```

Which number of clusters seems appropriate? Consider both the separation of the groups and whether adding more clusters reveals meaningful additional structure.

**Your answer:** *Write your answer here.*

### 3. Add cluster labels to the dataset

Based on your result above, fit a final k-means model and assign the cluster labels returned by `kmeans()` to the original data frame (`employee_work_df`).
Store the labels as a factor.

```{r}
#| label: 3-add-cluster-labels-to-the-dataset-student-response
# YOUR CODE HERE
```

### 4. Compare outcomes across the employee profiles

The clusters were created only from work-pattern variables. We can now ask whether the resulting employee profiles also differ in outcomes.

Create grouped boxplots to compare **performance** and **job satisfaction** across the clusters.
Try to complete the task using the [`ggplot2` boxplot documentation](https://ggplot2.tidyverse.org/reference/geom_boxplot.html).

```{r}
#| label: 4-create-grouped-boxplots-student-response
# YOUR CODE HERE
```

What do the boxplots tell you about the employee profiles represented by the clusters?

**Your answer:** *Write your answer here.*

:::

:::{.sol}

## Solution

### 1. Prepare the data for clustering

We use `meeting_hours`, `focus_hours`, and `remote_share` to describe employees' working patterns:

```{r}
#| label: prepare-clustering-data-solution
X <- employee_work_df |>
  select(meeting_hours, focus_hours, remote_share)

X |>
  slice_head(n = 5)
```

#### Exploratory check

```{r}
#| label: exploratory-check-summary
summary(X)
```

```{r}
#| label: exploratory-check-x-pipeline
X |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ variable, scales = "free") +
  labs(
    x = NULL,
    y = "Count"
  )
```

These summaries and histograms help us understand the distributions and the different scales of the variables.

#### Standardization

Since k-means is sensitive to scale differences, we standardize the variables. `scale()` centers each variable around zero and scales it to unit standard deviation.

```{r}
#| label: standardization-x-scaled
X_scaled <- scale(X)
head(X_scaled)
```

#### Scatterplot before clustering

```{r}
#| label: scatterplot-before-clustering-x-scaled-df
X_scaled_df <- as.data.frame(X_scaled)

ggplot(
  X_scaled_df,
  aes(x = meeting_hours, y = focus_hours)
) +
  geom_point() +
  labs(
    title = "Standardized work-pattern data (no clustering)",
    x = "Meeting hours (scaled)",
    y = "Focus hours (scaled)"
  )
```

**Interpretation:**
The scatterplot suggests approximately **three groups**: employees with relatively high focus and low meeting time, employees with high meeting and low focus time, and a larger group between these extremes.

### 2. Apply k-means clustering

We try different values of `k` and visually inspect the cluster structure. Setting `nstart = 25` makes the result less sensitive to a single random initialization.

```{r}
#| label: 2-apply-k-means-clustering-for
for (k in 2:10) {
  set.seed(42)

  km <- kmeans(
    X_scaled,
    centers = k,
    nstart = 25
  )

  cluster_plot_data <- X_scaled_df |>
    mutate(cluster = factor(km$cluster))

  centers <- as.data.frame(km$centers)

  cluster_plot <- ggplot(
    cluster_plot_data,
    aes(
      x = meeting_hours,
      y = focus_hours,
      color = cluster
    )
  ) +
    geom_point() +
    geom_point(
      data = centers,
      aes(x = meeting_hours, y = focus_hours),
      inherit.aes = FALSE,
      shape = 4,
      size = 4,
      stroke = 1.5
    ) +
    labs(
      title = paste("k =", k),
      x = "Meeting hours (scaled)",
      y = "Focus hours (scaled)",
      color = "Cluster"
    )

  print(cluster_plot)
}
```

**Conclusion:**
A value of **k = 3** captures the main structure well. Larger values increasingly subdivide groups that already appear internally similar rather than revealing clearly distinct additional profiles.

### 3. Add cluster labels to the dataset

We now fit the final model reproducibly and store the cluster labels as a factor because they represent categories rather than quantities:

```{r}
#| label: 3-add-cluster-labels-to-the-dataset-set-seed
set.seed(42)

k <- 3
km <- kmeans(
  X_scaled,
  centers = k,
  nstart = 25
)

employee_work_df <- employee_work_df |>
  mutate(cluster = factor(km$cluster))

employee_work_df |>
  slice_head(n = 5)
```

To help interpret the clusters, we can summarize the average work patterns and outcomes in each group:

```{r}
#| label: 3-summarize-clusters
employee_work_df |>
  group_by(cluster) |>
  summarise(
    across(
      c(
        meeting_hours,
        focus_hours,
        remote_share,
        performance,
        job_satisfaction
      ),
      mean
    ),
    .groups = "drop"
  )
```

The numerical cluster labels themselves have no inherent meaning. Interpret each cluster based on its observed characteristics.

### 4. Compare outcomes across the employee profiles

We compare performance and job satisfaction across clusters. Because the two measures use different scales, we reshape the data and allow each facet to use its own y-axis.

```{r}
#| label: 4-create-grouped-boxplots-ggplot
employee_work_df |>
  select(cluster, performance, job_satisfaction) |>
  pivot_longer(
    cols = c(performance, job_satisfaction),
    names_to = "outcome",
    values_to = "value"
  ) |>
  ggplot(aes(x = cluster, y = value)) +
  geom_boxplot() +
  facet_wrap(~ outcome, scales = "free_y") +
  labs(
    title = "Employee outcomes by cluster",
    x = "Cluster",
    y = NULL
  )
```

**Interpretation:**
The clusters differ not only in their work patterns but also in their outcome distributions. This helps us characterize the groups as organizationally meaningful employee profiles rather than merely geometric groupings in the clustering variables.

:::

:::{.direction}

## Wrap-up {#wrap-up}

🎉🎈 You have completed the exercise—good work! 🎈🎉

In this exercise, we have learned to

* Explore distributions and relationships in employee work data
* Select and standardize variables for clustering
* Compare different numbers of clusters using a `for` loop
* Apply and interpret k-means clustering
* Use variables not included in the clustering to characterize the resulting profiles

:::
