---
title: "Notes: Session 1: Big Data"
session_id: session-01
event_id: null
---

## Lecture

| Time (min) | Duration | Topic |
|------------|----------|-------|
| 0–10 | 10 | Growth of data and data-volume units |
| 10–20 | 10 | Sources of data production |
| 20–30 | 10 | Defining Big Data: the 4Vs |
| 30–55 | 25 | Business cases for the 4Vs |
| 55–65 | 10 | Metadata and location intelligence |
| 65–75 | 10 | Critically analyzing Big Data examples |
| 75–82 | 7 | Are all Vs necessary? |
| 82–90 | 8 | Summary and analytical setup |

::: objective
Explain why the growth and changing characteristics of data matter for business analytics.
:::

## Data production

Highlight **progression**, note user-generated/crowdsourced content, new devices, ..., AI-generated content.

Emphasize that big data is not simply “lots of data.” Each V presents a distinct managerial dilemma: decision complexity (volume), decisions while events unfold (velocity), integration of previously separate data worlds (variety), and whether apparent signals can be trusted (veracity).

Use the UPS, Visa, Walmart, and digital-advertising cases to connect each characteristic to models, operational decisions, and business outcomes.

{{< pagebreak >}}

## Analytical environment

- Briefly highlight the importance or R in the industry (R or Python) - see [survey](https://xebia.com/de/blog/data-survey-2018-2019-data-50/)
- Ask students: Who has RStudio installed?
- Who is familiar with qmd documents?
- Step-by-step: create qmd (header, context), activate render on save, create an R codeblock (with/out curly parentheses for execution), show ctrl+Enter execution, show environment

::: {.callout-note title="Important"}
Download the qmd files. do not work in the repository (html rendering fails due to `_quarto.yml` settings)
:::

- Work in pairs.

{{< pagebreak >}}

## Report.qmd exercise

- On Kaggle, search for ``Advertising`` and download the ``Sales Prediction (Simple Linear Regression)`` dataset (by ``ASHISH``).
- Add an R block using the interface
- Run simple ``read_csv()`` and inspect results, then add the assignment, run again, and add the ``show_col_types = FALSE`` based on the message
- Inspect the environment
- Render the html, show the "display output only" option in the R code chunk
- For ggplot, start with ``ggplot() + geom_point()`` - run and interpret errors. Add aesthetics based on docs
- Beautify: ``head(advertising) |> knitr::kable()``

````markdown
---
title: "Advertising and Sales"
format: html
------------

## Business question

Is higher TV advertising expenditure associated with higher sales?

## Data

The dataset contains advertising expenditures for different channels and sales.

```{{r}}
#| warning: false
library(tidyverse)
advertising <- read_csv(
  "advertising.csv",
  show_col_types = FALSE
)
head(advertising)
```

## Analysis

```{{r}}
ggplot(advertising, aes(x = TV, y = Sales)) +
  geom_point()
```

## Interpretation

There is a positive relationship between TV advertising expenditure and sales:
observations with higher TV advertising expenditure tend to have higher sales.
````

{{< pagebreak >}}

## Output

```{r}
#| message: false
library(tidyverse)

advertising <- read_csv(
  "https://raw.githubusercontent.com/JWarmenhoven/ISLR-python/master/Notebooks/Data/Advertising.csv",
  show_col_types = FALSE
)
head(advertising)
```


```{r}
#| echo: false
#| out-width: "40%"

ggplot(advertising, aes(x = TV, y = Sales)) +
  geom_point() +
  labs(
    x = "TV advertising",
    y = "Sales"
  )
```

Extensions:

- ggplot: included in the example
- Bibliography: add ``bibliography: references.bib`` to the yaml header + include ``@cit-key``, add explicit ``# References`` at the end.
- PDF rendering: add ``pdf: default`` in the format section, use the Render dropdown.
  **Note:** The default PDF template and citation style will be accepted for submissions.


At the end: highlight that one goal is to get a good understanding of well-structured and contextualized analytical reports.

{{< pagebreak >}}

## Note on extracting/importing functionality for operational systems

Put executable logic in `.R` files and `source()` them from Quarto. This lets the **report and the operational system use exactly the same code**.
For example:

```text
project/
├── report.qmd
├── R/
│   ├── functions.R
│   └── pipeline.R
```

`R/functions.R`:

```r
prepare_data <- function(data) {
  data |> dplyr::filter(!is.na(churn))
}
```

`R/pipeline.R`:

```r
train_model <- function(data) {
  glm(
    churn ~ age + monthly_spend,
    data = data,
    family = binomial
  )
}

predict_churn <- function(model, data) {
  predict(model, newdata = data, type = "response")
}
```

Then in `report.qmd`:

````markdown
```{{r}}
source("R/functions.R")
source("R/pipeline.R")

customers <- readr::read_csv("data/customers.csv")
customers <- prepare_data(customers)
model <- train_model(customers)
customers$churn_probability <- predict_churn(model, customers)
```
````

> **Quarto does not have to contain the analytical implementation. It can call the same reusable functions and pipeline that an operational system calls.**

<!--
TODO: one thing that students said last semester was that there is still a difference between a report and an operational system that updates with new data.
-> agree
-> show quarto dasboards: https://quarto.org/docs/gallery/#dashboards
-> illustrate extraction of analytical code (notebook for contextualization and illustration, same code also imported and used in operational systems)
-> if the operational systems uses R, you can import R code, if it is in Python, import Python.
TODO: also make sure the example is in R!
-->

{{< pagebreak >}}

## Exercise

| Time (min) | Duration | Topic |
|------------|----------|-------|
| 0–15 | 15 | RStudio and Quarto walkthrough |
| 15–30 | 15 | Importing and inspecting data |
| 30–50 | 20 | Building the first analytical report |
| 50–65 | 15 | Visualization with `ggplot2` |
| 65–75 | 10 | Rendering and presentation |
| 75–85 | 10 | Optional extensions |
| 85–90 | 5 | Reproducibility and reusable analytical code |

- Goal: broadly cover all skills required to navigate RStudio, Quarto reports and R code (libraries, in-memory objects).
  You probably know several of these things, but nevertheless go through the notebook.
- Do not render the whole document at the beginning, work through it step-by-step, fill out cells and run them individually.
  Then, once the document is complete, you may render it completely (rerun the analyses).

Notes:

| Part | Solution / key point |
|---|---|
| **Console vs document** | `17 * 23` → **391**. Console code is transient; code in the `.qmd` is **saved and reproducible**. |
| **Quarto `eval: false`** | The non-running cell should contain `#| eval: false` followed by `print("This should not run")`. The code remains visible but is not executed during render. |
| **Objects / memory** | `answer <- 6 * 7`; `ls()` should include `answer`; `rm(answer)` removes it; subsequent `ls()` should not. Objects belong to the **R session**, not the filesystem. |
| **Package help** | Missing-value argument of `read_csv()` is **`na`**. |
| **Inspect `students`** | `nrow(students)` → **5**; `ncol(students)` → **4**; `mean(students$score)` → **75.6**; `students$name` → Alex, Bao, Casey, Dina, Emil. |
| **Different data views** | Printing: good for **small-table values**. `glimpse()`: best for **column names/types + sample values**. `View()`: best for **browsing larger tables**. |
| **File vs memory** | After `rm(students)`, `file.exists("students.csv")` remains `TRUE`: `rm()` removes the **R object**, not the CSV stored on disk. |
| **Modify data** | `students_updated <- students |> mutate(score = score + 2)`; then `write_csv(students_updated, "students_updated.csv")`; `file.exists("students_updated.csv")` → `TRUE`. |
| **Reproducibility check** | After restarting, old objects should be absent. Rendering must recreate everything **from code in document order**, showing that the analysis does not depend on manual execution or a saved workspace. |
| **Optional challenge** | `library(tidyverse)` → `students <- read_csv("students.csv")` → `head(students, 3)` → `ls()` → `rm(students)`. |
