---
title: "Notes: Session 1: Big Data"
session_id: session-01
event_id: null
---

## Lecture

| Time (min) | Duration | Topic | Additional materials |
|------------|----------|-------|----------------------|
| 0–20       | 20       | Growth and sources of data | |
| 20–90      | 70       | Big Data, the 4Vs, and business cases | |

::: objective
Explain why the growth and changing characteristics of data matter for business analytics.
:::

## Data production

- highlight progression, note user-generated/crowdsourced content, new devices, ..., AI-generated content

Emphasize that big data is not simply “lots of data.” Each V presents a distinct managerial dilemma: decision complexity (volume), decisions while events unfold (velocity), integration of previously separate data worlds (variety), and whether apparent signals can be trusted (veracity).

Use the UPS, Visa, Walmart, and digital-advertising cases to connect each characteristic to models, operational decisions, and business outcomes.


## Analytical environment

- Ask students: Who has RStudio installed?
- Who is familiar with qmd documents?
- Step-by-step: create qmd (header, context), activate render on save, create an R codeblock (with/out curly parentheses for execution), show ctrl+Enter execution, show environment

Work in pairs.

Base task: simple Kaggle

Next: students build their own report/test

Extensions ():

- Bibliography: add ``bibliography: references.bib`` to the yaml header + include ``@cit-key``, add explicit ``# References`` at the end.
- PDF rendering: add ``pdf: default`` in the format section, use the Render dropdown.
  **Note:** The default PDF template and citation style will be accepted for submissions.

- ggplot:

```{r}
library(tidyverse)

# Load CSV directly from the web
url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/main/inst/extdata/penguins.csv"

penguins <- read_csv(url)

# Inspect first rows
head(penguins)
```

```{r}
# Scatterplot
ggplot(penguins, aes(x = flipper_length_mm,
                     y = body_mass_g,
                     color = species)) +
  geom_point() +
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species"
  )
```

Show the "display output only" option in the R code chunk