---
title: "Notes: Session 5: Classification"
session_id: session-05
format: html
---

2026-04-14: 65 min for the lecture + 25 min for the discussion of group projects

| Time (min) | Duration | Topic                         | Additional materials |
|------------|----------|-------------------------------|----------------------|
| 0–30       | 30       | From regression to classification |                   |
| 30–60      | 30       | Logistic regression and probabilities |              |
| 60–90      | 30       | Thresholds, evaluation, and decisions |              |


TODO

- Use the linear-regression comparison to surface modeling tensions before introducing classification.
- Emphasize the distinction between probability estimation and the threshold-based classification decision.

```{r}
#| echo: false
library(tidyverse)
set.seed(1)
d <- tibble(x = seq(0, 10, length.out = 50), y = 2*x + 3 + rnorm(50, 0, 3))
ggplot(d, aes(x, y)) + geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Linear Regression (Continuous Outcome)", x = "X (continuous)", y = "Y (continuous)")
```

```{r}
#| echo: false
library(tidyverse)
set.seed(1)
d <- tibble(group = factor(sample(LETTERS[1:4], 80, replace = TRUE)),
            y = c(A=5, B=7, C=10, D=6)[group] + rnorm(80, 0, 1.5))
ggplot(d, aes(group, y)) + geom_jitter(width = .05, alpha = .6) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  labs(title = "Regression with 4-Level Categorical Predictor",
       x = "Categorical Predictor (4 levels)", y = "Y (continuous)")
```

```{r}
#| echo: false
library(tidyverse)
set.seed(1)
d <- tibble(x = seq(0, 10, length.out = 100), p = plogis(-5 + x), y = rbinom(100, 1, p))
ggplot(d, aes(x)) + geom_point(aes(y = y), alpha = .4) + geom_line(aes(y = p)) +
  labs(title = "Logistic Regression (Binary Outcome)",
       x = "X (continuous)", y = "Probability / Outcome (0 or 1)")
```


### Classification Approaches

* **Black box models**
  → Directly map input → class (low interpretability)

* **Score + threshold**
  → Compute a score, then classify via cutoff
  (separates prediction from decision)

* **Regions / rules**
  → Partition input space into decision areas
  (e.g., trees, rule systems)

### Logistic Regression Choice

We use the **score approach**, where the score is: $p(x)=P(Y=1\mid x)$

→ **Interpretable probability** + threshold ⇒ classification

## From Logit to Sigmoid (whiteboard)

Start with the **Logit formulation** and define:
$$\log\left(\frac{y}{1-y}\right)=\beta_0 + \beta_1 x = z$$

**Solve for (y)**

Exponentiate:
$$
\frac{y}{1-y} = e^{z}
$$

Rearrange:
$$
y = (1-y)\cdot e^{z}
$$

$$
y + y e^{z} = e^{z}
$$

$$
y(1 + e^{z}) = e^{z}
$$

$$
y = \frac{e^{z}}{1 + e^{z}}
$$


**Final step (multiply numerator and denominator by $e^{-z}$)**:

$$
y = \frac{e^{z}}{1 + e^{z}} \cdot \frac{e^{-z}}{e^{-z}}
= \frac{1}{1 + e^{-z}}
$$

## Exercise

2026-04-14: 90 min, did not start with Part 5 (students indicated that we could discuss solutions earlier)

| Time (min) | Duration | Topic                         | Additional materials |
|------------|----------|-------------------------------|----------------------|
| 0–30       | 30       | Group work assignment         |                      |
| 30–90      | 60       | Logistic regression           |                      |


Part 1.3: TODO: extract confidence and interpret practical and statistical significance? (the small coefficient for income could be practically significant because of the high values of income, which is not standardized)

{{< pagebreak >}}

## Materials

- good slides: https://harvard-iacs.github.io/2019-CS109A/lectures/lecture10/ (https://harvard-iacs.github.io/2019-CS109A/lectures/lecture11/presentation/Lecture11_LogReg2.pdf)
- good (simple; geographical) explanation:  https://www.youtube.com/watch?v=yIYKR4sgzI8
- Lecture, Ng: https://www.youtube.com/watch?v=4u81xU7BIOc
- https://web.stanford.edu/class/cme250/files/cme250_lecture2.pdf
- https://slds-lmu.github.io/i2ml/chapters/03_supervised_classification/03-04-classification-logistic/
- https://ubc-cs.github.io/cpsc330-2023W1/README.html
- Optional /extension: changing the data: https://ubc-cs.github.io/cpsc330-2023W1/lectures/09_classification-metrics.html#optional-changing-the-data
- https://harvard-iacs.github.io/2019-CS109A/lectures/lecture-11/notebook/
- https://github.com/UBC-CS/cpsc330-2023W1
- https://slds-lmu.github.io/i2ml/
- https://www2.stat.duke.edu/courses/Spring20/sta210.001/labs/lab-08-logistic.html

